---
title: "The Product Space (HS): Descriptive Analysis"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini, Beatrice Liuni, Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

# **Descriptive Analysis**

>  All the statistics in this analysis describe the
> distributed network, which is a *pruned backbone* of the full product space
> (a maximum spanning tree plus the links with $\varphi > 0.55$), not the complete
> proximity matrix. Weak proximity links were removed by construction, so measures
> such as density, degree, clustering, and distances refer to this strong-link
> backbone only. We keep this in mind when interpreting the results, and we note it
> explicitly where it matters.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(igraph)
library(here)
library(tidyverse)
```

## **Data Loading**

```{r load-graph}
g = readRDS(here("data", "processed", "product_space.rds"))
layout_xy = readRDS(here("data", "processed", "layout_xy.rds"))
summary(g)
```

## **Descriptive Analysis of the Data**

This notebook performs the descriptive analysis of the Product Space (HS) network.
It loads the cleaned graph produced in the `data_preparation.Rmd` and explores its
basic properties. We first check what the vertices and edges represent, whether we observe the full
system or only a part, and whether there is sampling or missing data.

```{r view-tables}
nodes_tbl = igraph::as_data_frame(g, what = "vertices") |> as_tibble()
edges_tbl = igraph::as_data_frame(g, what = "edges")    |> as_tibble()

nodes_tbl |> select(name, pid, product, community_ref, size_layout) |> head(10)

edges_tbl |> head(10)
```

The edge table shows two metric columns, `width_layout` and `weight`, with the same values; this is on purpose. 

The original value is the line *thickness* used to draw the network in the Atlas
visualization. According to the dataset documentation this thickness is
non-linear and meant for layout only, so it is **not** the true proximity score
$\varphi$ and should not be read as a real edge weight. We keep it under two names
for two reasons: `width_layout` records what the value really is (a drawing
thickness), while `weight` is the name that igraph looks for when running weighted
computations. Our main analyses treat the network as **unweighted**; we use the
weight only later, as a robustness check, to confirm that results do not depend on
this layout-only value.

### **Communities**

The dataset assigns each product to one of twelve **communities**, stored in the
`community_ref` attribute. These were recovered from the official Product Space
colours: in the original visualization each product is coloured by the group it
belongs to, and we mapped those colours back to a community label. It is important
to be clear about their status: the communities are a given attribute of the dataset, not something we compute. They are a ready-made grouping that we use as
a key to read the rest of the analysis. Later, in the modelling part, we will run
a community-detection algorithm and use these official communities as a benchmark
to compare against; here we simply describe them.

We present the communities now, before the structural analysis, so that the
following sections (degree, clustering, centrality) can be read with these groups
already in mind.

### **Community Sizes**

```{r community-sizes}
nodes_tbl |> count(community_ref, sort = TRUE)
```

The twelve communities are very uneven in size. The largest, community 2, contains
252 products (about 29% of the network), while the smallest, community 1, contains
only 19. A few large communities account for most of the products, and several
small ones cover the rest.

### **What each Community represents**

To turn these groups from numbers into something interpretable, we look at a sample 
of products from each one.

```{r community-products}
nodes_tbl |>
  select(community_ref, product) |>
  arrange(community_ref) |>
  group_by(community_ref) |>
  slice_head(n = 6) |>
  print(n = 72)
```

Reading the products in each group, the communities line up with recognisable
economic sectors. The labels below are **our own descriptive reading**, based on
the most representative products; they are not official categories provided by the
dataset.

| Community | Size | Descriptive label |
|-----------|------|--------------------|
| 2  | 252 | Heavy machinery, transport equipment and metalworking |
| 11 | 103 | Livestock and fishing products |
| 8  | 90  | Fine and specialty chemicals |
| 3  | 82  | Raw minerals and metals |
| 4  | 81  | Textiles and apparel |
| 7  | 59  | Electronics and precision machinery |
| 9  | 50  | Mixed goods (no single clear sector) |
| 12 | 43  | Paper, wood and construction materials |
| 6  | 38  | Agri-food and basic chemicals |
| 10 | 27  | Fresh agricultural products and tobacco |
| 5  | 22  | Light consumer goods and miscellany |
| 1  | 19  | Precision instruments and luxury goods |

Most communities are clearly tied to a sector: textiles and apparel (community 4),
raw minerals (3), fine chemicals (8), livestock and fishing (11), electronics (7),
and so on. This coherence is also a sanity check on the data: products grouped
together do share an economic logic, which confirms that the community labels are
meaningful and correctly aligned with the products.

A few products sit in a community that does not match their obvious type,for
example a woven-wool fabric inside the machinery community. This is expected rather
than an error: the communities come from **co-export patterns**, not from a product
classification, so a product is grouped with others exported by the same countries,
which need not be of the same kind. Community 9 is the hardest to label, mixing
silk, printed matter, fishing vessels and explosives; we keep it as "mixed".

These descriptive labels are used from here on as a reading key for the network: in
the following sections and in the map, each community can be referred to by its
sector rather than by a number alone.

### **Data Quality Check**

Before analysing the structure, we check that the data are clean: no missing
values, no duplicated edges, no self-loops, and no isolated products.

```{r data-quality}
sum(is.na(nodes_tbl))
sum(is.na(edges_tbl))

any(which_multiple(g))

any(which_loop(g))

sum(degree(g) == 0)
```

The dataset is clean at the source: there are no missing values in either table, no
duplicated edges, no self-loops, and no isolated products. No additional cleaning is
needed before analysis. This also answers the *missingness* question from the
network-mapping framework: there is no accidental missing data. The only reduction
is the backbone construction (maximum spanning tree + $\varphi > 0.55$), which
removes weak links on purpose, not by mistake.

## **Network Structure**

Having described what the dataset contains, its products, links, communities, and
overall quality, we now analyse the structure of the network: how the products
are connected, how the connections are distributed, and which products and links
play the most important roles. We move from global properties (size, connectivity)
to local ones (degree, clustering, centrality), and finally to the links
themselves (edge attributes). Throughout, we read the results with the twelve
communities introduced above in mind.

### **Network Size and Connectivity**

We begin with the global properties of the network: its size, how densely the
products are connected, whether it forms a single piece, and how far apart products
are from one another.

```{r network-size}
vcount(g)
ecount(g)

edge_density(g)

components(g)$no

diameter(g, weights = NA)

mean_distance(g, weights = NA)
```

The network has 866 products and 2,532 proximity links. Its density is very low
($\approx 0.0068$): only about 0.7% of all possible edges are present, as expected
for a pruned backbone. Despite being sparse, the network forms a **single connected
component**, so every product is reachable from every other. This is partly
guaranteed by construction: the maximum spanning tree at the core of the backbone
connects all products by design.

The diameter is 21 (the longest shortest path) and the average distance is
$\approx 7.1$ steps between two products. We will interpret these values together
with the degree and centrality results in the next sections, rather than drawing
conclusions now.

### **Degree**

The degree of a product is the number of other products it is directly connected
to. It is the simplest measure of how connected a product is, and the starting
point for understanding the local structure of the network.

```{r degree}
deg = degree(g)
summary(deg)
```

The summary shows a strong imbalance. The median degree is only 2, so half of the
products have two connections or fewer, while the mean is about 5.8 and the maximum
reaches 48. When the mean sits well above the median, the distribution is skewed to
the right: most products have few links, and a small number act as highly connected hubs.

```{r degree-top10}
nodes_tbl|>
  mutate(degree = deg)|>
  arrange(desc(degree))|>
  select(product, community_ref, degree)|>
  head(10)
```

The ten most connected products are mostly heavy-manufacturing goods (forged steel
balls, transmission shafts, cast iron articles) together with a few others such as
bricks and T-shirts. Notably, most of them belong to community 2 (heavy machinery
and metalworking), the largest community. We return to what this concentration
means when we look at centrality and at the network map.

#### **Degree Distribution**

```{r degree-histogram}
deg_tbl = tibble(degree = deg)

ggplot(deg_tbl, aes(x = degree)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Degree distribution",
       x = "Degree", y = "Number of products") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The histogram confirms the imbalance seen in the summary.
The large majority of products have a very low degree, with a tall peak at degree 1
(about 285 products) that drops off quickly. To the right there is a long, thin
tail of products with much higher degree, up to the single product with degree 48.
The shape is strongly
right-skewed: many low-degree products and a few high-degree hubs.

The histogram shows that the distribution has a heavy tail, but on a linear scale
that tail is compressed against the axis and its exact shape is hard to judge. A
common way to inspect heavy-tailed distributions is to plot them on a log-log
scale, where both axes are logarithmic. The reason is mathematical: if the degree
distribution follows a power law, $P(k) \propto k^{-\alpha}$, taking logs of both
sides turns the curve into a straight line,

$$
\log P(k) = c - \alpha \, \log k,
$$

so a power law, which is hard to recognise on a linear scale, appears as a
descending straight line with slope $-\alpha$. We use this plot to check whether
the degree distribution has that shape.

```{r degree-loglog}
deg_dist = as.data.frame(table(deg)) |>
  rename(degree = deg, freq = Freq) |>
  mutate(degree = as.numeric(as.character(degree)))

ggplot(deg_dist, aes(x = log(degree), y = log(freq))) +
  geom_point(color = "darkblue") +
  labs(title = "Degree distribution (log-log scale)",
       x = "log(Degree)", y = "log(Frequency)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

In our plot the points fall roughly along such a descending line, which is
consistent with a power-law degree distribution, though a
straight log-log shape alone does not prove a power law. The points are cleaner on
the left (low degrees), where each degree value is shared by many products, and
noisier on the right (high degrees), where only a few products take each value and
the frequencies are estimated less reliably. In the next step we estimate the exponent that describes this shape.

#### **Estimating the Power-Law Exponent**

To estimate the exponent $\alpha$, we fit a model to the degree-frequency data.

A common shortcut is to fit a linear model to the log-frequencies against the
log-degrees and read the slope as $-\alpha$. We avoid this, because the assumptions
of linear regression are not satisfied here. The response is the frequency of
each degree value, a *count* (how many products have a given degree), which is
discrete, non-negative, and has a variance that grows with its mean. Linear
regression instead assumes a continuous response with normal, constant-variance
errors. Taking the logarithm of the frequencies does not fix this: it distorts the
weighting of the points, giving the sparse, noisy high-degree tail (where many
frequencies equal 1, so their log is 0) the same weight as the well-populated
low-degree part. A linear fit would therefore give an unreliable estimate.

The appropriate tool for a count response is a Poisson regression, fitted with
`glm()`. It models the counts directly through a log link, without logging the
response, and uses the Poisson distribution, which is built for non-negative counts
with mean-dependent variance.

```{r degree-glm}
fit_glm = glm(freq ~ log(degree), family = poisson, data = deg_dist)
summary(fit_glm)
```

The coefficient of `log(degree)` is about $-1.33$ (z $= -43$, p $< 0.001$), so the
estimated exponent is $\alpha \approx 1.33$. The very small p-value and the large
drop from the null deviance (2032) to the residual deviance (46) confirm that
degree explains the frequencies well.

```{r degree-glm-plot}
glm_int   = coef(fit_glm)[1]
glm_slope = coef(fit_glm)[2]

ggplot(deg_dist, aes(x = log(degree), y = log(freq))) +
  geom_point(color = "darkblue") +
  geom_abline(intercept = glm_int, slope = glm_slope,
              color = "red", linewidth = 0.5) +
  labs(title = "Degree Distribution with fitted Poisson slope",
       x = "log(Degree)", y = "log(Frequency)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

This exponent is on the low side. We leave its assessment to the modelling part,
where it can be compared with the values produced by the generative models. As
noted in the backbone caveat, our network is also a pruned backbone, which affects
the tail of the distribution.

The degree distribution is strongly
heavy-tailed: most products have very few links, while a small number of hubs are
highly connected, the most connected reaching degree 48. These hubs are mostly
heavy-machinery goods from community 2. This uneven distribution is the first
structural feature of the network; the next sections examine whether these hubs are
also the most clustered and the most central.

### **Clustering and Transitivity**

After looking at how connected each product is, we ask whether the
neighbours of a product are also connected to each other, i.e. whether the network
forms *triangles*.

For a vertex $v$, let $\tau_\triangle(v)$ be the number of triangles involving $v$,
and $\tau_3(v) = \binom{d_v}{2}$ the number of connected triples (wedges) centred on
$v$. The local clustering coefficient is

$$
cl(v) = \frac{\tau_\triangle(v)}{\tau_3(v)},
$$

defined for vertices with degree at least 2. It measures, among the pairs of
neighbours of $v$, how many are actually connected.

These local values can be aggregated to the whole network in two ways:

$$
cl(G) = \frac{1}{|V'|}\sum_{v \in V'} cl(v); 
\qquad cl_T(G) = \frac{\sum_{v \in V'}\tau_\triangle(v)}{\sum_{v \in V'}\tau_3(v)}.
$$

The first is a simple average of the local coefficients; the second, the
transitivity, is a weighted average. Following Bollobás and Riordan, we use the
transitivity $cl_T(G)$ as our main measure, since the weighting makes it less
sensitive to low-degree vertices. Values close to 1 indicate a highly clustered
network; values close to 0 indicate few triangles.

```{r clustering}
# Transitivity: cl_T(G)
transitivity(g, type = "global")

# Clustering coefficient: cl(G)
transitivity(g, type = "average")
```

The transitivity is $cl_T(G) \approx 0.43$ and the simple clustering coefficient is
$cl(G) \approx 0.28$. Both are moderate-to-high values: the neighbours of a product
are often connected to each other, so the network forms a fair number of triangles.
This is the quantitative counterpart of the sector groups described in earlier sections:
products that belong to the same community tend to be linked to one another, which
produces triangles within each group. The value is notable given that we work with
a pruned backbone, where removing weak links also removes potential triangles.

The two measures also differ ($cl_T(G) > cl(G)$). In the transitivity, each vertex
$v$ is weighted by $\tau_3(v) = \binom{d_v}{2}$, the number of connected triples
centred on it. Since this grows quickly with degree, hubs receive much larger
weights than peripheral nodes. The fact that the weighted measure is higher than the
simple average therefore suggests that the more connected products (the hubs) tend
to be more clustered than the peripheral ones: their neighbours are also often
linked to one another.

Together with the heavy-tailed degree distribution, this clustering completes a
consistent picture so far: a few highly connected, densely interconnected hubs
alongside many peripheral products, with connections concentrated inside the sector
communities.

#### **Clustering Within Communities**

The global transitivity describes the network as a whole. A natural next question
is whether clustering is uniform across the network or whether some communities are
more tightly knit than others. We compute the transitivity of each community
separately, by taking the subgraph of the products in that community and measuring
how often their links close into triangles.

```{r clustering-by-community}
# Transitivity computed within each community's subgraph
comm_clustering <- nodes_tbl |>
  distinct(community_ref) |>
  arrange(community_ref) |>
  mutate(
    n_products   = map_int(community_ref, ~ sum(V(g)$community_ref == .x)),
    transitivity = map_dbl(community_ref, function(c) {
      sub = induced_subgraph(g, V(g)[V(g)$community_ref == c])
      transitivity(sub, type = "global")
    })
  )

comm_clustering |> arrange(desc(transitivity))
```

The within-community transitivity varies widely,
from 0.73 down to 0. The global value of 0.43 was therefore an average that hides
strong differences between groups. The most tightly knit communities are textiles
and apparel (community 4, 0.73), agri-food and basic chemicals (6, 0.56), and
paper, wood and construction (12, 0.52): in these sectors the products are almost
all interconnected. At the other end, raw minerals (3, 0.13), luxury and precision
goods (1, 0) and the mixed community 9 (0) form few or no internal triangles, which
fits the fact that these groups are more heterogeneous.

A notable point concerns community 2 (heavy machinery), the largest community and
the one containing the most central products. Its internal transitivity is only
0.27, below the global average. So community 2 is large and central without forming
a dense internal block: its products are not heavily interconnected among
themselves. Smaller communities such as textiles are far more self-contained: fewer
products, but almost fully interconnected. Why community 2 is nonetheless so central
becomes clearer when we look at centrality and at the links between communities.

This refines the earlier picture: clustering is not uniform but concentrated inside
certain sector communities, while the central hub community is comparatively open.

### **Centrality**

The degree section asked how many links a product has. We now ask a broader
question: how important is a product in the network? This is the idea of
centrality. There is no single definition of importance, so rather than pick one
measure, we consider four standard ones, each built on a different idea, and then
select those that are most informative for this network.

- Degree centrality counts direct neighbours. Already studied in Section 3.2,
  here it serves as a baseline.
- Closeness centrality is based on *distance*: a product is important if it is
  close, on average, to all others through short paths.
- Betweenness centrality is based on *flow*: a product is important if it lies
  on many shortest paths between other products, acting as a bridge.
- **Eigenvector centrality** is based on *feedback*: a product is important if it is
  connected to other important products.

For a vertex $v \in V$:

**Closeness** is the inverse of the total distance from $v$ to all other vertices,

$$
c_{Cl}(v) = \frac{1}{\sum_{u \in V} \mathrm{dist}(v, u)},
$$

so a small total distance gives a large closeness. It is only well-defined on a
connected graph; ours is a single component, so this is not a problem.

**Betweenness** counts how often $v$ sits on shortest paths between other vertices,

$$
c_B(v) = \sum_{s \neq t \neq v \in V} \frac{\sigma(s, t \mid v)}{\sigma(s, t)},
$$

where $\sigma(s, t)$ is the number of shortest paths between $s$ and $t$, and
$\sigma(s, t \mid v)$ the number of those passing through $v$.

**Eigenvector centrality** (Bonacich) defines importance recursively, as
proportional to the sum of the neighbours' importances,

$$
c_{Ei}(u) = \alpha \sum_{\{u, v\}} c_{Ei}(v),
$$

which in matrix form is the leading-eigenvector problem for the adjacency matrix
$A$: $A\,\vec{c}_{Ei} = \alpha^{-1}\,\vec{c}_{Ei}$. All these measures are computed
on the pruned backbone, since they depend on the shortest paths within it.

#### **Selecting the Measures**

We do not use all four measures blindly. A measure is useful only if it adds
information that degree does not already give: if it ranks products almost like
degree, it is redundant. To check this, we plot each of the three other measures
against degree.

```{r centrality-explore}
clos = closeness(g, weights = NA)
betw = betweenness(g, weights = NA)
eig  = eigen_centrality(g, weights = NA)$vector

cent_explore = tibble(
  degree      = deg  / max(deg),
  closeness   = clos / max(clos),
  betweenness = betw / max(betw),
  eigenvector = eig  / max(eig)
)
```

> We rescale the measures in order to make a better visual comparison of the centrality measures considered.

```{r centrality-pairs}
cent_explore |>
  pivot_longer(c(closeness, betweenness, eigenvector),
               names_to = "measure", values_to = "value") |>
  ggplot(aes(x = degree, y = value)) +
  geom_point(alpha = 0.4, color = "darkblue") +
  facet_wrap(~ measure, scales = "free_y") +
  labs(title = "Degree vs other centrality measures (rescaled)",
       x = "Degree (rescaled)", y = "Other measure (rescaled)") +
  theme_minimal()
```

The three panels show how much each measure adds beyond degree.

* **Betweenness (left)**: The points form a widely dispersed cloud with no clear
trend. At any given degree, betweenness ranges from near zero up to high values:
several products with very low degree (around 0.05–0.15 on the rescaled axis) reach
betweenness of 0.4–0.6, while many products with high degree have betweenness close
to zero. Knowing a product's degree tells us almost nothing about its betweenness.
This measure is therefore strongly independent of degree and carries genuinely
new information: it identifies products that act as bridges regardless of how many
links they have.

* **Closeness (centre)**: The shape is structured but not redundant. There is a dense
vertical band at low degree where closeness spans almost the entire range (from
about 0.35 to 0.9): products with the same low degree can be very differently
placed, some close to the rest of the network, some far. As degree increases,
closeness rises and then flattens towards a ceiling near 1, so high-degree products
are consistently close to everything. Closeness thus partly follows degree, but
adds clear information among the low-degree products, separating those that are
well positioned from those that are peripheral despite the same number of links.

* **Eigenvector (right)**: The points trace a clean, increasing curve: eigenvector
stays near zero for almost all products until degree becomes high (above $\approx 0.4$ on
the rescaled axis), then rises steeply towards 1. This is close to a deterministic
function of degree, knowing the degree essentially predicts the eigenvector
ranking. The measure is therefore largely redundant with degree and adds little.

Reading the three together, betweenness brings the most new information (fully
independent of degree), closeness brings useful additional information concentrated
among low-degree products, and eigenvector brings little beyond what degree already
tells us. We therefore keep betweenness and **closeness** for the analysis,
alongside degree as a baseline, and set eigenvector centrality aside.

#### **Analysis on Selected Measures**

We store the selected measures as vertex attributes and examine which products they
rank as central.

```{r centrality-compute}
V(g)$degree      = deg
V(g)$betweenness = betw
V(g)$closeness   = clos
vertex_attr_names(g)
```

The three measures are now part of the network object. We build a table and look at
the most central products under each of the two selected measures.

```{r centrality-tables}
cent_tbl <- tibble(
  product       = V(g)$product,
  community_ref = V(g)$community_ref,
  degree        = V(g)$degree,
  betweenness   = V(g)$betweenness,
  closeness     = V(g)$closeness
)
```

```{r centrality-top-betweenness}
# Top 10 by betweenness (the bridging role)
cent_tbl |> arrange(desc(betweenness)) |>
  select(product, community_ref, betweenness, degree) |> head(10)
```

```{r centrality-top-closeness}
# Top 10 by closeness (the short-reach role)
cent_tbl |> arrange(desc(closeness)) |>
  select(product, community_ref, closeness, degree) |> head(10)
```

The top products by betweenness are mostly metal and machinery goods, but three of
them have very low degree: electrical machinery parts (degree 6), aircraft engines
(degree 3), and dried potatoes (degree 4). The top products by closeness, in
contrast, all have medium-to-high degree (22–48). The forged steel balls (degree
48) rank first in both lists, and most of the central products in both belong to
community 2 (heavy machinery).

```{r centrality-scatter-betweenness}
ggplot(cent_tbl, aes(x = degree, y = betweenness)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  labs(title = "Degree vs betweenness", x = "Degree", y = "Betweenness") +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```

```{r centrality-scatter-closeness}
ggplot(cent_tbl, aes(x = degree, y = closeness)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  labs(title = "Degree vs closeness", x = "Degree", y = "Closeness") +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```

The scatter plots explain the difference. Degree versus betweenness is a dispersed
cloud with no clear trend: betweenness is not determined by degree, so it captures
an independent bridging role, and this is why low-degree products can still rank at
the top. Degree versus closeness instead shows a clear shape: closeness rises with
degree and then flattens towards a ceiling, and high closeness appears only together
with at least medium degree. Closeness therefore tracks degree more closely, while
still distinguishing low-degree products that are well positioned from those that
are peripheral.

Overall, centrality points to two kinds of important products: well-embedded
products scoring high on both degree and closeness (forged steel balls, transmission
shafts), and pure bridges scoring high on betweenness despite low degree (aircraft
engines, dried potatoes). No single measure captures both roles. The concentration
of central products in community 2 confirms, from a different angle, that this
community acts as the connective core of the network, consistent with its large
size and comparatively low internal clustering seen earlier.