---
title: "The Product Space (HS): Dataset Construction"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini, Beatrice Liuni, Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(igraph)
library(here)
```

# **Introduction**

This document explains **how the Product Space network is built** and then
**constructs it from the raw files** saved in `data/raw/`. The companion
`dataset_description.md` is the concise technical reference; here we give the
conceptual reasoning behind the data, since understanding the construction is
essential to interpreting every later result.

# **What the network represents**

The Product Space connects **traded products**. Two products are linked when the
same countries tend to be competitive exporters of both. The economic intuition
(Hidalgo & Hausmann) is that if a country is good at exporting two products, those
products probably require **similar productive capabilities**, comparable skills,
infrastructure, institutions and technology. Products that share exporters are
therefore "close" in capability space.

Crucially, the network is a **one-mode projection** of a bipartite
country-product network. Countries are used to *build* it but are then collapsed
away: the files contain products only. The country identities cannot be recovered.

# **How an edge is built**

## **1. From export values to competitive exports (RCA)**

The raw input is for every country, how much it exports of every product (in dollars). Using dollar values directly is misleading, because large economies export a lot of everything. Example: a small country exporting $1B$ of steel may be highly *specialised* in steel, while a giant exporting $50B$ may simply be big. Absolute values would hide that.

The **Revealed Comparative Advantage (RCA)** fixes this by comparing shares rather than absolute values:

$$
RCA_{c,p} \;=\;
\frac{\; x_{c,p} \big/ \sum_{p'} x_{c,p'} \;}
     {\; \sum_{c'} x_{c',p} \big/ \sum_{c'}\sum_{p'} x_{c',p'} \;}
$$

The numerator is product $p$'s share of country $c$'s exports; the denominator is
$p$'s share of total world exports. If $RCA_{c,p} \ge 1$, product $p$ weighs more
in country $c$ than it does worldwide, so the country is **specialised** in it. This
normalises away country size, so we compare *specialisations*, not magnitudes.

We then reduce everything to a binary country x product matrix:

$$
M_{c,p} =
\begin{cases}
1 & \text{if } RCA_{c,p} \ge 1 \quad(\text{country } c \text{ is competitive in } p)\\
0 & \text{otherwise.}
\end{cases}
$$

## **2. Projecting onto products (proximity)**

Two products are "close" if the **same countries** are competitive in both. For
products $i$ and $j$, proximity is the minimum of two conditional probabilities:

$$
\varphi_{i,j} \;=\;
\min\Big[\,
P\!\left(M_{\cdot,i}=1 \mid M_{\cdot,j}=1\right),\;
P\!\left(M_{\cdot,j}=1 \mid M_{\cdot,i}=1\right)
\,\Big]
$$

*Worked example* - Five countries, two products:

| Country | T-shirts | Jackets |
|---|:---:|:---:|
| Bangladesh | 1 | 1 |
| Vietnam | 1 | 1 |
| China | 1 | 1 |
| Italy | 0 | 1 |
| Germany | 0 | 0 |

Countries competitive in T-shirts: 3; in Jackets: 4.

- $P(\text{T-shirts}\mid\text{Jackets}) = 3/4 = 0.75$
- $P(\text{Jackets}\mid\text{T-shirts}) = 3/3 = 1.0$
- $\varphi = \min(0.75, 1.0) = \mathbf{0.75}$ -> high -> the two products are close.

By contrast, T-shirts vs microchips share almost no exporters -> $\varphi \approx 0$
-> no edge.

**Why the minimum?** To neutralise *ubiquitous* products (exported by almost
everyone). For such a product, $P(\text{ubiquitous}\mid\text{anything})\approx 1$
purely because everyone exports it, not because of shared capabilities. Requiring
*both* directions to be high (the minimum) prevents a one-sided overlap from
inflating the score.

## **3. From a dense matrix to a backbone**

Proximity is defined for *every* pair, so the full network is almost complete
(~374,000 possible edges), mostly negligible. The distributed network is a
**backbone**: a maximum spanning tree (guaranteeing one connected component) plus
the links with $\varphi > 0.55$ (Hidalgo et al., 2007). 

 A *spanning tree* is a minimal set of
edges that connects every node into a single component without forming cycles
(exactly *n − 1* edges for *n* nodes), the smallest possible skeleton that leaves
no node isolated. Among all possible spanning trees, the *maximum* spanning tree
is the one that maximises the total edge weight; here, where the weight is
proximity, it is the skeleton that connects all products using the **strongest**
available links. This step matters because applying the $\varphi > 0.55$ threshold alone
could strip some products of all their links and fragment the network. By first
laying down the maximum spanning tree and then overlaying every edge above the
threshold, the construction yields a network that is simultaneously **sparse**
(only strong links) and **fully connected** (no isolated products). This is
precisely why the network we analyse has a single connected component and no
isolated nodes.

Every statistic in this project therefore describes the backbone, not the raw proximity matrix.

## **4. What is kept and what is pruned**

The construction *consumes* information that is then absent from the files: the
country layer, the matrix $M$, the RCA values and the raw proximity matrix are all
used and discarded. No country-level or geographic quantity is recoverable. Only
the product nodes and their backbone edges remain.

# **Constructing the network from raw files**

> Install packages once, in the Console (not on knit): `install.packages(c("here","igraph"))`.

## **Read and clean columns names**

`check.names = FALSE` preserves the original headers; we strip the leading `#`.

```{r load-raw}
nodes_raw = read.csv(here("data", "raw", "nodes.csv"),
                      stringsAsFactors = FALSE, check.names = FALSE)
edges_raw = read.csv(here("data", "raw", "edges.csv"),
                      stringsAsFactors = FALSE, check.names = FALSE)
gprops    = read.csv(here("data", "raw", "gprops.csv"),
                      stringsAsFactors = FALSE, check.names = FALSE)

clean_names = function(x) trimws(gsub("^#\\s*", "", x))
names(nodes_raw) = clean_names(names(nodes_raw))
names(edges_raw) = clean_names(names(edges_raw))
names(gprops)    = clean_names(names(gprops))
```

## **Fix mislabelled columns and flag layout-only attributes**

Headers follow the SITC layout, but this is the HS file: the column named `leamer` holds the **product name** and `name` holds the **hex colour**. We also rename `size`/`width` to make clear they are non-linear, layout-only.

```{r fix-columns}
nodes = data.frame(
  index       = as.integer(nodes_raw$index),
  pid         = nodes_raw$pid,                 # HS product id
  product     = trimws(nodes_raw$leamer),      # real product name
  size_layout = nodes_raw$size,                # market share, non-linear, LAYOUT ONLY
  color       = trimws(nodes_raw$name),        # hex colour = author community
  stringsAsFactors = FALSE
)
```

## **Recover the community from the colour**

Each distinct colour is one author-defined community; we map it to an integer id
to use later as a reference partition.

```{r community-ref}
nodes$community_ref = as.integer(factor(nodes$color))
cat("Reference communities:", length(unique(nodes$community_ref)), "\n")
print(table(nodes$community_ref))
```

## **Parse layout coordinates**

```{r parse-pos}
parse_pos = function(s) {
  nums = regmatches(s, gregexpr("-?[0-9]+\\.?[0-9]*", s))
  do.call(rbind, lapply(nums, function(v) as.numeric(v[1:2])))
}
xy = parse_pos(nodes_raw$pos)
nodes$x = xy[, 1]; nodes$y <- xy[, 2]
```

## **Clean edges**

The `width` is layout-only and non-linear. We keep it as `width_layout` for visual plotting purposes and never as the proximity $\varphi$.

```{r clean-edges}
edges = data.frame(
  source       = as.integer(edges_raw$source),
  target       = as.integer(edges_raw$target),
  width_layout = edges_raw$width,             # drawing thickness, non-linear, LAYOUT ONLY
  stringsAsFactors = FALSE
)
```

## **Build the graph with provenance metadata**

We explicitly do not assign an igraph `weight` attribute, ensuring the network is treated as unweighted by default in all structural computations. Provenance and the construction rule are stored as graph attributes so they travel with the object.

```{r build-graph}
vertices_df = nodes[, c("index", "pid", "product", "community_ref",
                         "size_layout", "color", "x", "y")]
edges_df    = edges[, c("source", "target", "width_layout")]

g = graph_from_data_frame(d = edges_df, directed = FALSE, vertices = vertices_df)

g = set_graph_attr(g, "name", "Product Space (HS)")
g = set_graph_attr(g, "source", "Coscia PS_HS.zip; Hidalgo et al. Science 2007")
g = set_graph_attr(g, "construction", "Maximum Spanning Tree + proximity phi > 0.55")
g = set_graph_attr(g, "note", "network is strictly unweighted; width_layout is for visual plotting only")

layout_xy = as.matrix(nodes[, c("x", "y")])
```

## **Verify and inspect**

```{r verify}
stopifnot(
  vcount(g) == 866, ecount(g) == 2532,
  !any(which_loop(g)), !any(which_multiple(g)), min(degree(g)) >= 1
)
cat("Nodes:", vcount(g), "| Edges:", ecount(g),
    "| Components:", components(g)$no,
    "| Weighted:", is_weighted(g),
    "| Reference communities:", length(unique(V(g)$community_ref)), "\n")
summary(g)
head(igraph::as_data_frame(g, what = "vertices"))
```

### **A note on the edge weight**

The graph reports `Weighted: FALSE`.

The dataset provides a `width` value for edges (ranging from 1 to 4), which represents the **drawing thickness** of the edge in the original Atlas visualization. The official documentation describes this attribute as non-linear and meant for layout purposes only. It is not the underlying proximity score $\varphi$ and cannot be converted back to it. Concretely, an edge with a width of 4 does not mean it represents a connection four times stronger than a width of 1; the number only preserves a visual ranking.

Because of this, our methodological choice is deliberate: we treat the network as strictly unweighted in all analyses, dropping the standard igraph `weight` attribute entirely to avoid accidental usage in routing or centrality algorithms. We retain these thickness values solely under the custom width_layout attribute, accessing them exclusively when scaling the links to draw the canonical network maps.

## **Save Processed Objetcs**

```{r save-processed}
saveRDS(g,         here("data", "processed", "product_space.rds"))
saveRDS(layout_xy, here("data", "processed", "layout_xy.rds"))
write.csv(nodes, here("data", "processed", "nodes_clean.csv"), row.names = FALSE)
write.csv(edges, here("data", "processed", "edges_clean.csv"), row.names = FALSE)
list.files(here("data", "processed"))
```

# **Summary**

The cleaned object is a strictly **unweighted**,  **undirected** network of **866 products** and **2,532 backbone edges**, in a single connected component, ccarrying a layout-only thickness attribute (used strictly for visualization) and a 12-community reference partition recovered from node colour. It is built from country export data via RCA and proximity, then reduced to a maximum-spanning-tree backbone with $\varphi > 0.55$. The country layer and the raw proximity values are not recoverable, so the analysis is product-side and topological.
