---
title: "The Product Space (HS): Descriptive Analysis"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini | Beatrice Liuni | Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

# **Descriptive Analysis**

>  All the statistics in this analysis describe the
> distributed network, which is a *pruned backbone* of the full product space
> (a maximum spanning tree plus the links with $\varphi > 0.55$), not the complete
> proximity matrix. Weak proximity links were removed by construction, so measures
> such as density, degree, clustering, and distances refer to this strong-link
> backbone only. We keep this in mind when interpreting the results, and we note it
> explicitly where it matters.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(igraph)
library(here)
library(tidyverse)
```

## **Data Loading**

```{r load-graph}
g = readRDS(here("data", "processed", "product_space.rds"))
layout_xy = readRDS(here("data", "processed", "layout_xy.rds"))
summary(g)
```

## **Descriptive Analysis of the Data**

This notebook performs the descriptive analysis of the Product Space (HS) network.
It loads the cleaned graph produced in the `data_preparation.Rmd` and explores its
basic properties. We first check what the vertices and edges represent, whether we observe the full
system or only a part, and whether there is sampling or missing data.

```{r view-tables}
nodes_tbl = igraph::as_data_frame(g, what = "vertices") |> as_tibble()
edges_tbl = igraph::as_data_frame(g, what = "edges")    |> as_tibble()

nodes_tbl |> select(name, pid, product, community_ref, size_layout) |> head(10)

edges_tbl |> head(10)
```

The edge table shows the metric column `width_layout`. 

The original value is the line *thickness* used to draw the network in the Atlas
visualization. According to the dataset documentation, this thickness is
non-linear and meant for layout only, so it is **not** the true proximity score
$\varphi$ and should not be read as a real edge weight. Because it is a layout-only
attribute, we treat the network as strictly unweighted in all our structural
analyses. 

### **Communities**

The dataset assigns each product to one of twelve **communities**, stored in the
`community_ref` attribute. These were recovered from the official Product Space
colours: in the original visualization each product is coloured by the group it
belongs to, and we mapped those colours back to a community label. It is important
to be clear about their status: the communities are a given attribute of the dataset, not something we compute. They are a ready-made grouping that we use as
a key to read the rest of the analysis. Later, in the modelling part, we will run
a community-detection algorithm and use these official communities as a benchmark
to compare against; here we simply describe them.

We present the communities now, before the structural analysis, so that the
following sections (degree, clustering, centrality) can be read with these groups
already in mind.

### **Community Sizes**

```{r community-sizes}
nodes_tbl |> count(community_ref, sort = TRUE)
```

The twelve communities are very uneven in size. The largest, community 2, contains
252 products (about 29% of the network), while the smallest, community 1, contains
only 19. A few large communities account for most of the products, and several
small ones cover the rest.

### **What each Community represents**

To turn these groups from numbers into something interpretable, we look at a sample 
of products from each one.

```{r community-products}
nodes_tbl |>
  select(community_ref, product) |>
  arrange(community_ref) |>
  group_by(community_ref) |>
  slice_head(n = 6) |>
  print(n = 72)
```

Reading the products in each group, the communities line up with recognisable
economic sectors. The labels below are **our own descriptive reading**, based on
the most representative products; they are not official categories provided by the
dataset.

| Community | Size | Descriptive label |
|-----------|------|--------------------|
| 2  | 252 | Heavy machinery, transport equipment and metalworking |
| 11 | 103 | Livestock and fishing products |
| 8  | 90  | Fine and specialty chemicals |
| 3  | 82  | Raw minerals and metals |
| 4  | 81  | Textiles and apparel |
| 7  | 59  | Electronics and precision machinery |
| 9  | 50  | Mixed goods (no single clear sector) |
| 12 | 43  | Paper, wood and construction materials |
| 6  | 38  | Agri-food and basic chemicals |
| 10 | 27  | Fresh agricultural products and tobacco |
| 5  | 22  | Light consumer goods and miscellany |
| 1  | 19  | Precision instruments and luxury goods |

Most communities are clearly tied to a sector: textiles and apparel (community 4),
raw minerals (3), fine chemicals (8), livestock and fishing (11), electronics (7),
and so on. This coherence is also a **sanity check** on the data: products grouped
together do share an economic logic, which confirms that the community labels are
meaningful and correctly aligned with the products.

A few products sit in a community that does not match their obvious type, for
example a woven-wool fabric inside the machinery community. This is expected rather
than an error, for two complementary reasons. First, the communities come from
**co-export patterns**, not from a product classification: a product is grouped with
others exported by the same countries, which need not be of the same kind. Second,
the community labels are defined on the **full** proximity network, while the links
we analyse belong to the **pruned backbone** (maximum spanning tree plus the
strongest links, as noted at the start). A product keeps its original label even if,
after pruning, its few surviving links connect it mostly to other groups: the strong
links that justified its community may have been removed, while weaker cross-group
links happened to survive. An apparent "misplacement" therefore reflects the
backbone construction, not a wrong assignment. Community 9 is the hardest to label,
mixing silk, printed matter, fishing vessels and explosives; we keep it as "mixed".

These descriptive labels are used from here on as a reading key for the network: in
the following sections and in the map, each community can be referred to by its
sector rather than by a number alone. We will also see, when analysing the links
between communities, that this gap between labels and pruned connections has a
visible effect on the network's structure.

### **Data Quality Check**

Before analysing the structure, we check that the data are clean: no missing
values, no duplicated edges, no self-loops, and no isolated products.

```{r data-quality}
sum(is.na(nodes_tbl))
sum(is.na(edges_tbl))

any(which_multiple(g))

any(which_loop(g))

sum(degree(g) == 0)
```

The dataset is clean at the source: there are no missing values in either table, no
duplicated edges, no self-loops, and no isolated products. No additional cleaning is
needed before analysis. This also answers the *missingness* question from the
network-mapping framework: there is no accidental missing data. The only reduction
is the backbone construction (maximum spanning tree + $\varphi > 0.55$), which
removes weak links on purpose, not by mistake.

## **Network Structure**

Having described what the dataset contains, its products, links, communities, and
overall quality, we now analyse the structure of the network: how the products
are connected, how the connections are distributed, and which products and links
play the most important roles. We move from global properties (size, connectivity)
to local ones (degree, clustering, centrality), and finally to the links
themselves (edge attributes). Throughout, we read the results with the twelve
communities introduced above in mind.

### **Network Size and Connectivity**

We begin with the global properties of the network: its size, how densely the
products are connected, whether it forms a single piece, and how far apart products
are from one another.

```{r network-size}
vcount(g)
ecount(g)

edge_density(g)

components(g)$no

diameter(g, weights = NA)

mean_distance(g, weights = NA)
```

The network has 866 products and 2,532 proximity links. Its density is very low
($\approx 0.0068$): only about 0.7% of all possible edges are present, as expected
for a pruned backbone. Despite being sparse, the network forms a **single connected
component**, so every product is reachable from every other. This is partly
guaranteed by construction: the maximum spanning tree at the core of the backbone
connects all products by design.

The diameter is 21 (the longest shortest path) and the average distance is
$\approx 7.1$ steps between two products. We will interpret these values together
with the degree and centrality results in the next sections, rather than drawing
conclusions now.

### **Degree**

The degree of a product is the number of other products it is directly connected
to. It is the simplest measure of how connected a product is, and the starting
point for understanding the local structure of the network.

```{r degree}
deg = degree(g)
summary(deg)
```

The summary shows a strong imbalance. The median degree is only 2, so half of the
products have two connections or fewer, while the mean is about 5.8 and the maximum
reaches 48. When the mean sits well above the median, the distribution is skewed to
the right: most products have few links, and a small number act as highly connected hubs.

```{r degree-top10}
nodes_tbl|>
  mutate(degree = deg)|>
  arrange(desc(degree))|>
  select(product, community_ref, degree)|>
  head(10)
```

The ten most connected products are mostly heavy-manufacturing goods (forged steel
balls, transmission shafts, cast iron articles) together with a few others such as
bricks and T-shirts. Notably, most of them belong to community 2 (heavy machinery
and metalworking), the largest community. We return to what this concentration
means when we look at centrality and at the network map.

#### **Degree Distribution**

```{r degree-histogram}
deg_tbl = tibble(degree = deg)

ggplot(deg_tbl, aes(x = degree)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  labs(title = "Degree distribution",
       x = "Degree", y = "Number of products") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The histogram confirms the imbalance seen in the summary.
The large majority of products have a very low degree, with a tall peak at degree 1
(about 285 products) that drops off quickly. To the right there is a long, thin
tail of products with much higher degree, up to the single product with degree 48.
The shape is strongly
right-skewed: many low-degree products and a few high-degree hubs.

The histogram shows that the distribution has a heavy tail, but on a linear scale
that tail is compressed against the axis and its exact shape is hard to judge. A
common way to inspect heavy-tailed distributions is to plot them on a log-log
scale, where both axes are logarithmic. The reason is mathematical: if the degree
distribution follows a power law, $P(k) \propto k^{-\alpha}$, taking logs of both
sides turns the curve into a straight line,

$$
\log P(k) = c - \alpha \, \log k,
$$

so a power law, which is hard to recognise on a linear scale, appears as a
descending straight line with slope $-\alpha$. We use this plot to check whether
the degree distribution has that shape.

```{r degree-loglog}
deg_dist = as.data.frame(table(deg)) |>
  rename(degree = deg, freq = Freq) |>
  mutate(degree = as.numeric(as.character(degree)))

ggplot(deg_dist, aes(x = log(degree), y = log(freq))) +
  geom_point(color = "darkblue") +
  labs(title = "Degree distribution (log-log scale)",
       x = "log(Degree)", y = "log(Frequency)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

In our plot the points fall roughly along such a descending line, which is
consistent with a power-law degree distribution, though a
straight log-log shape alone does not prove a power law. The points are cleaner on
the left (low degrees), where each degree value is shared by many products, and
noisier on the right (high degrees), where only a few products take each value and
the frequencies are estimated less reliably. In the next step we estimate the exponent that describes this shape.

#### **Estimating the Power-Law Exponent**

To estimate the exponent $\alpha$, we fit a model to the degree-frequency data.

A common shortcut is to fit a linear model to the log-frequencies against the
log-degrees and read the slope as $-\alpha$. We avoid this, because the assumptions
of linear regression are not satisfied here. The response is the frequency of
each degree value, a *count* (how many products have a given degree), which is
discrete, non-negative, and has a variance that grows with its mean. Linear
regression instead assumes a continuous response with normal, constant-variance
errors. Taking the logarithm of the frequencies does not fix this: it distorts the
weighting of the points, giving the sparse, noisy high-degree tail (where many
frequencies equal 1, so their log is 0) the same weight as the well-populated
low-degree part. A linear fit would therefore give an unreliable estimate.

The appropriate tool for a count response is a Poisson regression, fitted with
`glm()`. It models the counts directly through a log link, without logging the
response, and uses the Poisson distribution, which is built for non-negative counts
with mean-dependent variance.

```{r degree-glm}
fit_glm = glm(freq ~ log(degree), family = poisson, data = deg_dist)
summary(fit_glm)
```

The coefficient of `log(degree)` is about $-1.33$ (z $= -43$, p $< 0.001$), so the
estimated exponent is $\alpha \approx 1.33$. The very small p-value and the large
drop from the null deviance (2032) to the residual deviance (46) confirm that
degree explains the frequencies well.

```{r degree-glm-plot}
glm_int   = coef(fit_glm)[1]
glm_slope = coef(fit_glm)[2]

ggplot(deg_dist, aes(x = log(degree), y = log(freq))) +
  geom_point(color = "darkblue") +
  geom_abline(intercept = glm_int, slope = glm_slope,
              color = "red", linewidth = 0.5) +
  labs(title = "Degree Distribution with fitted Poisson slope",
       x = "log(Degree)", y = "log(Frequency)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

This exponent is on the low side. We leave its assessment to the modelling part,
where it can be compared with the values produced by the generative models. As
noted in the backbone caveat, our network is also a pruned backbone, which affects
the tail of the distribution.

The degree distribution is strongly
heavy-tailed: most products have very few links, while a small number of hubs are
highly connected, the most connected reaching degree 48. These hubs are mostly
heavy-machinery goods from community 2. This uneven distribution is the first
structural feature of the network; the next sections examine whether these hubs are
also the most clustered and the most central.

### **Clustering and Transitivity**

After looking at how connected each product is, we ask whether the
neighbours of a product are also connected to each other, i.e. whether the network
forms *triangles*.

For a vertex $v$, let $\tau_\triangle(v)$ be the number of triangles involving $v$,
and $\tau_3(v) = \binom{d_v}{2}$ the number of connected triples (wedges) centred on
$v$. The local clustering coefficient is

$$
cl(v) = \frac{\tau_\triangle(v)}{\tau_3(v)},
$$

defined for vertices with degree at least 2. It measures, among the pairs of
neighbours of $v$, how many are actually connected.

These local values can be aggregated to the whole network in two ways:

$$
cl(G) = \frac{1}{|V'|}\sum_{v \in V'} cl(v); 
\qquad cl_T(G) = \frac{\sum_{v \in V'}\tau_\triangle(v)}{\sum_{v \in V'}\tau_3(v)}.
$$

The first is a simple average of the local coefficients; the second, the
transitivity, is a weighted average. Following Bollobás and Riordan, we use the
transitivity $cl_T(G)$ as our main measure, since the weighting makes it less
sensitive to low-degree vertices. Values close to 1 indicate a highly clustered
network; values close to 0 indicate few triangles.

```{r clustering}
# Transitivity: cl_T(G)
transitivity(g, type = "global")

# Clustering coefficient: cl(G)
transitivity(g, type = "average")
```

The transitivity is $cl_T(G) \approx 0.43$ and the simple clustering coefficient is
$cl(G) \approx 0.28$. Both are moderate-to-high values: the neighbours of a product
are often connected to each other, so the network forms a fair number of triangles.
This is the quantitative counterpart of the sector groups described in earlier sections:
products that belong to the same community tend to be linked to one another, which
produces triangles within each group. The value is notable given that we work with
a pruned backbone, where removing weak links also removes potential triangles.

The two measures also differ ($cl_T(G) > cl(G)$). In the transitivity, each vertex
$v$ is weighted by $\tau_3(v) = \binom{d_v}{2}$, the number of connected triples
centred on it. Since this grows quickly with degree, hubs receive much larger
weights than peripheral nodes. The fact that the weighted measure is higher than the
simple average therefore suggests that the more connected products (the hubs) tend
to be more clustered than the peripheral ones: their neighbours are also often
linked to one another.

Together with the heavy-tailed degree distribution, this clustering completes a
consistent picture so far: a few highly connected, densely interconnected hubs
alongside many peripheral products, with connections concentrated inside the sector
communities.

#### **Clustering Within Communities**

The global transitivity describes the network as a whole. A natural next question
is whether clustering is uniform across the network or whether some communities are
more tightly knit than others. We compute the transitivity of each community
separately, by taking the subgraph of the products in that community and measuring
how often their links close into triangles.

```{r clustering-by-community}
# Transitivity computed within each community's subgraph
comm_clustering <- nodes_tbl |>
  distinct(community_ref) |>
  arrange(community_ref) |>
  mutate(
    n_products   = map_int(community_ref, ~ sum(V(g)$community_ref == .x)),
    transitivity = map_dbl(community_ref, function(c) {
      sub = induced_subgraph(g, V(g)[V(g)$community_ref == c])
      transitivity(sub, type = "global")
    })
  )

comm_clustering |> arrange(desc(transitivity))
```

The within-community transitivity varies widely,
from 0.73 down to 0. The global value of 0.43 was therefore an average that hides
strong differences between groups. The most tightly knit communities are textiles
and apparel (community 4, 0.73), agri-food and basic chemicals (6, 0.56), and
paper, wood and construction (12, 0.52): in these sectors the products are almost
all interconnected. At the other end, raw minerals (3, 0.13), luxury and precision
goods (1, 0) and the mixed community 9 (0) form few or no internal triangles, which
fits the fact that these groups are more heterogeneous.

A notable point concerns community 2 (heavy machinery), the largest community and
the one containing the most central products. Its internal transitivity is only
0.27, below the global average. So community 2 is large and central without forming
a dense internal block: its products are not heavily interconnected among
themselves. Smaller communities such as textiles are far more self-contained: fewer
products, but almost fully interconnected. Why community 2 is nonetheless so central
becomes clearer when we look at centrality and at the links between communities.

This refines the earlier picture: clustering is not uniform but concentrated inside
certain sector communities, while the central hub community is comparatively open.

### **Centrality**

The degree section asked how many links a product has. We now ask a broader
question: how important is a product in the network? This is the idea of
centrality. There is no single definition of importance, so rather than pick one
measure, we consider four standard ones, each built on a different idea, and then
select those that are most informative for this network.

- **Degree centrality** counts direct neighbours,
  here it serves as a baseline.
- **Closeness centrality** is based on *distance*: a product is important if it is
  close, on average, to all others through short paths.
- **Betweenness centrality** is based on *flow*: a product is important if it lies
  on many shortest paths between other products, acting as a bridge.
- **Eigenvector centrality** is based on *feedback*: a product is important if it is
  connected to other important products.

For a vertex $v \in V$:

**Closeness** is the inverse of the total distance from $v$ to all other vertices,

$$
c_{Cl}(v) = \frac{1}{\sum_{u \in V} \mathrm{dist}(v, u)},
$$

so a small total distance gives a large closeness. It is only well-defined on a
connected graph; ours is a single component, so this is not a problem.

**Betweenness** counts how often $v$ sits on shortest paths between other vertices,

$$
c_B(v) = \sum_{s \neq t \neq v \in V} \frac{\sigma(s, t \mid v)}{\sigma(s, t)},
$$

where $\sigma(s, t)$ is the number of shortest paths between $s$ and $t$, and
$\sigma(s, t \mid v)$ the number of those passing through $v$.

**Eigenvector centrality** (Bonacich) defines importance recursively, as
proportional to the sum of the neighbours' importances,

$$
c_{Ei}(u) = \alpha \sum_{\{u, v\}} c_{Ei}(v),
$$

which in matrix form is the leading-eigenvector problem for the adjacency matrix
$A$: $A\,\vec{c}_{Ei} = \alpha^{-1}\,\vec{c}_{Ei}$. All these measures are computed
on the pruned backbone, since they depend on the shortest paths within it.

#### **Selecting the Measures**

We do not use all four measures blindly. A measure is useful only if it adds
information that degree does not already give: if it ranks products almost like
degree, it is redundant. To check this, we plot each of the three other measures
against degree.

```{r centrality-explore}
clos = closeness(g, weights = NA)
betw = betweenness(g, weights = NA)
eig  = eigen_centrality(g, weights = NA)$vector

cent_explore = tibble(
  degree      = deg  / max(deg),
  closeness   = clos / max(clos),
  betweenness = betw / max(betw),
  eigenvector = eig  / max(eig)
)
```

> We rescale the measures in order to make a better visual comparison of the centrality measures considered.

```{r centrality-pairs}
cent_explore |>
  pivot_longer(c(closeness, betweenness, eigenvector),
               names_to = "measure", values_to = "value") |>
  ggplot(aes(x = degree, y = value)) +
  geom_point(alpha = 0.4, color = "darkblue") +
  facet_wrap(~ measure, scales = "free_y") +
  labs(title = "Degree vs other centrality measures (rescaled)",
       x = "Degree (rescaled)", y = "Other measure (rescaled)") +
  theme_minimal()
```

The three panels show how much each measure adds beyond degree.

* **Betweenness (left)**: The points form a widely dispersed cloud with no clear
trend. At any given degree, betweenness ranges from near zero up to high values:
several products with very low degree (around 0.05–0.15 on the rescaled axis) reach
betweenness of 0.4–0.6, while many products with high degree have betweenness close
to zero. Knowing a product's degree tells us almost nothing about its betweenness.
This measure is therefore strongly independent of degree and carries genuinely
new information: it identifies products that act as bridges regardless of how many
links they have.

* **Closeness (centre)**: The shape is structured but not redundant. There is a dense
vertical band at low degree where closeness spans almost the entire range (from
about 0.35 to 0.9): products with the same low degree can be very differently
placed, some close to the rest of the network, some far. As degree increases,
closeness rises and then flattens towards a ceiling near 1, so high-degree products
are consistently close to everything. Closeness thus partly follows degree, but
adds clear information among the low-degree products, separating those that are
well positioned from those that are peripheral despite the same number of links.

* **Eigenvector (right)**: The points trace a clean, increasing curve: eigenvector
stays near zero for almost all products until degree becomes high (above $\approx 0.4$ on
the rescaled axis), then rises steeply towards 1. This is close to a deterministic
function of degree, knowing the degree essentially predicts the eigenvector
ranking. The measure is therefore largely redundant with degree and adds little.

Reading the three together, betweenness brings the most new information (fully
independent of degree), closeness brings useful additional information concentrated
among low-degree products, and eigenvector brings little beyond what degree already
tells us. We therefore keep betweenness and **closeness** for the analysis,
alongside degree as a baseline, and set eigenvector centrality aside.

#### **Analysis on Selected Measures**

We store the selected measures as vertex attributes and examine which products they
rank as central.

```{r centrality-compute}
V(g)$degree      = deg
V(g)$betweenness = betw
V(g)$closeness   = clos
vertex_attr_names(g)
```

The three measures are now part of the network object. We build a table and look at
the most central products under each of the two selected measures.

```{r centrality-tables}
cent_tbl = tibble(
  product       = V(g)$product,
  community_ref = V(g)$community_ref,
  degree        = V(g)$degree,
  betweenness   = V(g)$betweenness,
  closeness     = V(g)$closeness,
)
```

```{r centrality-top-betweenness}
# Top 10 by betweenness (the bridging role)
cent_tbl |> arrange(desc(betweenness)) |>
  select(product, community_ref, betweenness, degree) |> head(10)
```

```{r centrality-top-closeness}
# Top 10 by closeness (the short-reach role)
cent_tbl |> arrange(desc(closeness)) |>
  select(product, community_ref, closeness, degree) |> head(10)
```

The top products by betweenness are mostly metal and machinery goods, but three of
them have very low degree: electrical machinery parts (degree 6), aircraft engines
(degree 3), and dried potatoes (degree 4). The top products by closeness, in
contrast, all have medium-to-high degree (22–48). The forged steel balls (degree
48) rank first in both lists, and most of the central products in both belong to
community 2 (heavy machinery).

```{r centrality-scatter-betweenness}
ggplot(cent_tbl, aes(x = degree, y = betweenness)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  labs(title = "Degree vs betweenness", x = "Degree", y = "Betweenness") +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```

```{r centrality-scatter-closeness}
ggplot(cent_tbl, aes(x = degree, y = closeness)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  labs(title = "Degree vs closeness", x = "Degree", y = "Closeness") +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```
The scatter plots explain the difference. Degree versus betweenness is a dispersed
cloud with no clear trend: betweenness is not determined by degree, so it captures
an independent bridging role, and this is why low-degree products can still rank at
the top. Degree versus closeness instead shows a clear shape: closeness rises with
degree and then flattens towards a ceiling, and high closeness appears only together
with at least medium degree. Closeness therefore tracks degree more closely, while
still distinguishing low-degree products that are well positioned from those that
are peripheral.

Overall, centrality points to two kinds of important products: those scoring high
on both degree and closeness (such as the forged steel balls and transmission
shafts), which are highly connected and close to the rest of the network, and pure
bridges scoring high on betweenness despite a low degree (such as aircraft engines
and dried potatoes). No single measure captures both roles, which is why we kept
more than one. Most of the central products under both measures belong to community
2 (heavy machinery). Together with its low internal clustering seen earlier, this
suggests that community 2 acts as a connective core: central to the network as a
whole without being densely knit inside. We examine the links between communities
in the next section.

### **Edge-Level Attributes**

We now compute two attributes of
the edges, to describe the links themselves rather than the products.

- **Edge betweenness** is betweenness applied to edges: for each link, the number of
  shortest paths in the network that pass through it. A link with high edge
  betweenness is a critical bridge, whose removal would lengthen or cut the
  connection between parts of the network.
- **Intra/inter-community flag** marks, for each link, whether it connects two
  products of the same community (*intra*) or of two different communities (*inter*).

```{r edge-attributes}
E(g)$betweenness = edge_betweenness(g, weights = NA)

el   = as_edgelist(g, names = FALSE)
comm = V(g)$community_ref
E(g)$cross_community = ifelse(
  comm[el[, 1]] == comm[el[, 2]], "intra", "inter"
)

edge_attr_names(g)
```

#### **Within- and Between-Community Links**

We first look at how many links stay inside a community and how many cross between
communities.

```{r edge-counts}
tibble(cross_community = E(g)$cross_community) |>
  count(cross_community) |>
  mutate(pct = round(n / sum(n) * 100, 2))
```

Of the 2,532 links, 1,966 (about 78%) stay within a community and only 566 (about
22%) cross between communities. Most connections therefore join products of the same
community. This is consistent with the high transitivity and the
sector communities described: products link mainly within their own
group, and only a minority of links bridge across groups.

#### **Edge Betweenness of Within- and Between-Community Links**

A common expectation is that between-community links carry more shortest paths than
within-community ones, since they are the few bridges connecting otherwise dense
groups, while inside a community there are many alternative short paths. We test
this by comparing the edge-betweenness distribution of the two types of link. We use
a violin plot on a logarithmic scale, because edge betweenness is highly
right-skewed (most links carry few paths, a few carry very many), and a log scale
makes the bulk of the distribution readable rather than compressed against the axis.
If the expectation holds, the *inter* distribution should sit higher than the
*intra* one

```{r edge-betweenness-dist}
tibble(
  betweenness     = E(g)$betweenness,
  cross_community = E(g)$cross_community
) |>
  ggplot(aes(x = cross_community, y = betweenness + 1, fill = cross_community)) +
  geom_violin(alpha = 0.5, color = NA) +
  geom_boxplot(width = 0.15, alpha = 0.7, outlier.size = 0.5) +
  scale_y_log10() +
  scale_fill_manual(values = c("inter" = "firebrick", "intra" = "steelblue")) +
  labs(title = "Edge betweenness: within vs between communities",
       x = NULL, y = "Edge betweenness (log scale)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), legend.position = "none")
```

```{r median-quartiles}
tibble(
  betweenness     = E(g)$betweenness,
  cross_community = E(g)$cross_community
) |>
  group_by(cross_community) |>
  summarise(
    median = median(betweenness),
    q1     = quantile(betweenness, 0.25),
    q3     = quantile(betweenness, 0.75)
  )
```  

The two distributions are broadly similar in shape: both are strongly right-skewed,
with most links carrying few shortest paths and a long upper tail reaching the same
high values in both groups. The strong expectation, that between-community links
are clearly the main bridges, is therefore not confirmed: high-betweenness links
exist within communities just as much as between them.

The log scale does, however, reveal a consistent difference in the expected
direction. On every quartile the *inter* links sit above the *intra* ones: median
edge betweenness is about 499 for between-community links against 361 for
within-community ones, with first and third quartiles of 149 vs 81 and 1211 vs 865.
So between-community links do tend to carry more shortest paths, consistent with a
bridging role, but the difference is modest (the medians differ by less than a
factor of 1.4) and the two distributions overlap heavily.

The most likely reason the gap is so small is the nature of the data. As noted in
Community-Description, community labels come from the full proximity network, while the links
we analyse are those of the pruned backbone. The maximum spanning tree at the core
of the backbone turns many links into obligatory passages of the skeleton, even when
they lie inside a community, so a large number of within-community links also act as
structural bridges. This compresses the difference between the two types and is a
reminder that these measures describe the backbone, not the full proximity network.

#### **Between-Community Connectivity**

The counts so far were global. We now return to a question left open in the previous
sections: whether community 2 really acts as a **connective core**. We saw that it
has the most central products but a low internal clustering, which suggested that it might be central not because its products are densely
linked among themselves, but because they connect outwards to other communities. We
can now test this directly, by measuring, for each community, what share of its
links goes to *other* communities. A community with a high share of
between-community links is relatively open and connective; one with a low share is
more self-contained.

```{r edge-inter-by-community}
edge_comm = tibble(
  c1   = comm[el[, 1]],
  c2   = comm[el[, 2]],
  type = E(g)$cross_community
)

bind_rows(
  edge_comm |> select(community = c1, type),
  edge_comm |> select(community = c2, type)
) |>
  count(community, type) |>
  pivot_wider(names_from = type, values_from = n, values_fill = 0) |>
  mutate(share_inter = round(inter / (inter + intra) * 100, 1)) |>
  arrange(desc(share_inter))
```

The share of between-community links varies widely. The most outward-facing is
community 9 (75%), which fits its status as a mixed group with no coherent sector:
lacking an internal theme, its products connect in many directions. Communities 12,
6 and 10 follow, with roughly half of their links going outside. At the other end,
textiles (community 4, 5%) and electronics (community 7, 6%) are highly
self-contained, keeping almost all their links internal.

The result for community 2 (heavy machinery) is more nuanced than expected. In
*relative* terms it is among the more closed communities: only about 14% of its
links cross to other communities, so most of its connections stay internal. Yet in
*absolute* terms it has the second largest number of outgoing links of any community (225),
because it is so large. So community 2 is not a connective core in the sense of
being proportionally open; its central position comes instead from its sheer size
and the resulting high absolute number of connections, both internal and external.
This refines, and partly corrects, the reading suggested by its low internal
clustering: that low clustering reflects the community's size and internal diversity
more than an unusually outward orientation.

These shares must be read with care, however. A large community such as community 2
covers nearly a third of all products, so by size alone a higher fraction of its
links is expected to fall inside it, regardless of any real tendency to be closed.
The percentages therefore partly reflect community size, not only structure: large
communities tend to look more self-contained and small ones more open as an artefact
of their size. Assessing whether a community is genuinely more or less closed than
its size alone would imply requires comparing the observed shares with those expected
under a suitable random reference, which we leave to the modelling part.

## **Network Visualization**

So far we have described the network through numbers. We now visualize it, to make
its structure directly visible. The dataset provides 
*canonical coordinates*, the
official Product Space layout, in which related products are placed close together,
so we do not need a force-directed algorithm (which on 866 nodes would tend to
produce an unreadable "hairball"). Using the given coordinates keeps the map
faithful to the way the Product Space is conventionally shown.

We build the map in two steps: a plain version showing the structure, then a
decorated version where colour encodes community and size encodes degree.

### **Unannotated Map - Product Space Layout**

We start with a plain map: every product placed at its canonical coordinates, with
no colour or size encoding, so that only the bare structure is visible.

```{r network-base, fig.width=10, fig.height=10}
plot(g,
     layout      = layout_xy,
     vertex.label = NA,
     vertex.size = 2,
     vertex.color = "grey40",
     edge.color  = "grey75",
     edge.width  = 0.3,
     main = "Product Space — Layout")
```

Even without decoration, the map already shows a clear structure: a dense central
band where many products are closely connected, and a sparser periphery of more
isolated products, with a few chains hanging off the edges. This is the visual
counterpart of the global numbers found earlier, low density, large diameter, and
long average distance, which together describe a network stretched between a
compact core and a far-reaching periphery. The decorated versions that follow add
community colour and node size to read this structure in more detail

### **Annotaded Map - Communities and Centrality**

We now decorate the map. Node colour encodes the community (using the official
Product Space colours, whose sectors we labelled), node size
encodes degree (so hubs stand out), and edge **thickness** uses `width_layout`. As
explained earlier, `width_layout` is a drawing-only attribute: we use it here solely
for the visual thickness of the links, never in any computation. A legend maps each
colour to the sector label we gave its community.

```{r network-decorated, fig.width=13, fig.height=10}
V(g)$plot_size = 1.5 + 3 * sqrt(V(g)$degree) / max(sqrt(V(g)$degree))

comm_palette = c("#0576a6", "#299fff", "#330000", "#33cc00", "#6e451e",
                  "#730049", "#99ffff", "#9c219c", "#9c9a87", "#bf8845",
                  "#d1ff00", "#ff0000")
comm_labels = c("1 – Precision instruments & luxury",
                 "2 – Heavy machinery & metalworking",
                 "3 – Raw minerals & metals",
                 "4 – Textiles & apparel",
                 "5 – Light consumer goods",
                 "6 – Agri-food & basic chemicals",
                 "7 – Electronics & precision machinery",
                 "8 – Fine & specialty chemicals",
                 "9 – Mixed goods",
                 "10 – Fresh agri & tobacco",
                 "11 – Livestock & fishing",
                 "12 – Paper, wood & construction")

par(mar = c(2, 2, 3, 12), xpd = TRUE)

plot(g,
     layout      = layout_xy,
     vertex.label = NA,
     vertex.size = V(g)$plot_size,
     vertex.color = V(g)$color,
     vertex.frame.color = NA,
     edge.color  = "grey75",
     edge.width  = E(g)$width_layout * 0.4,
     main = "Product Space — Communities and Centrality")

legend(x = 1.15, y = 1,
       legend = comm_labels,
       fill   = comm_palette,
       cex    = 0.7,
       bty    = "n",
       title  = "Communities (sectors)",
       title.adj = 0)
```

With colour, size and a sector legend, the map ties together what the numerical
analysis found. Three features stand out.

First, **the communities form distinct spatial regions** rather than being mixed:
electronics (cyan) in the upper left, heavy machinery (blue) across the centre,
fine chemicals (purple) in the lower left, agri-food and paper/construction (dark
red and red) in the centre-right, textiles (green) on the right, and livestock and
fishing (yellow) in the upper right. Products that the dataset assigns to the same
community are also placed close together and densely linked; it is the visual counterpart
of the high within-community transitivity measured before.

Second, **the network has a core–periphery organisation**. A dense central band
holds most of the large, highly connected products, while the periphery thins out
into scattered products and chains hanging off the edges (the dark raw-minerals and
brown fresh-agri groups at the bottom). This matches the global numbers: a low
density with a large diameter and long average distance describe a network stretched
between a compact core and a far-reaching periphery.

Third, **the hubs sit at the centre, inside the heavy-machinery community**. The
largest nodes lie in the blue region, consistent with the centrality results, where
the most central products belonged to community 2. This community also borders
almost every other one, while groups such as raw minerals and fresh agriculture sit
at the periphery, connected through fewer links, the spatial counterpart of
community 2's high centrality. At the same time, the map makes visible what the
clustering and edge analyses found: community 2 occupies the centre and spreads
widely rather than forming a tight, self-contained block; its products are numerous
and centrally placed, but loosely knit among themselves.

The map thus offers a single visual summary of the whole descriptive analysis: a
sector-structured network with a heavy-machinery core that borders most other
sectors, a sparse periphery, and hubs concentrated at the centre.

### **Locating the Pure Bridges**

The map so far shows the overall structure. We now use it to verify a specific
finding from the centrality analysis. In **Centrality** Section we identified three "pure
bridges": products with very high betweenness but very low degree (electrical
machinery parts, aircraft engines, and dried potatoes). Their high betweenness
suggested that they lie on the routes between otherwise separated parts of the
network, despite having few connections. The map lets us check this directly: if
these products really act as bridges, they should sit at points of passage rather
than buried inside a single dense region. We highlight them with black rings.

```{r network-bridges, fig.width=13, fig.height=10}
bridge_ids = c(123, 399, 518)

V(g)$highlight_size = V(g)$plot_size
V(g)$highlight_size[bridge_ids]  = 9

V(g)$highlight_frame = NA
V(g)$highlight_frame[bridge_ids] = "black"

par(mar = c(2, 2, 3, 12), xpd = TRUE)

plot(g,
     layout      = layout_xy,
     vertex.label = NA,
     vertex.size = V(g)$highlight_size,
     vertex.color = V(g)$color,
     vertex.frame.color = V(g)$highlight_frame,
     edge.color  = "grey80",
     edge.width  = E(g)$width_layout * 0.3,
     main = "Product Space - Pure bridges highlighted")
```

Their positions confirm their bridging role. None of them sits in the dense central
core where the high-degree hubs are; instead, each lies at a point of passage. The
electronics bridge (cyan, left) sits on the inner edge of its community, facing the
rest of the network; the machinery bridge (blue, centre) occupies a central transit
position despite having only three links; and the livestock bridge (yellow, upper
right) connects a peripheral group to the rest of the network from a relatively
isolated position. These products are important not because they are highly
connected, but because they sit on the routes between otherwise separated parts of
the network, exactly what high betweenness with low degree means.

## **Partial Conclusions**

The descriptive analysis has produced a consistent picture of the Product Space
backbone. The network is sparse (density $\approx 0.0068$) yet fully connected, with a
large diameter and an average distance of about seven steps, a structure stretched
between a compact core and a far-reaching periphery. Its degree distribution is
strongly heavy-tailed: most products have one or two links, while a few hubs reach
degree 48, and a Poisson fit to the degree-frequency data gives an exponent of
$\alpha \approx 1.33$. Clustering is moderate-to-high ($cl_T \approx 0.43$) but uneven:
within-community transitivity ranges from 0.73 in textiles down to near zero in the
mixed and luxury groups. Links are overwhelmingly internal — about 78% join products
of the same community — and the twelve sectors form distinct, spatially coherent
regions on the map.

Across several measures, one community stood out. Community 2 (heavy machinery) is the
largest and most central: its products dominate the degree and closeness rankings and
it borders almost every other group, yet its internal transitivity is only 0.27, well
below the tightly knit smaller sectors. It is, in short, large and central without
being internally dense: a connective region rather than a cohesive block.

These findings are descriptive: they measure *what* the network looks like, but they
cannot say *why* it looks this way, nor whether the patterns are more than the
by-product of size and density. Three questions, in particular, are left open, and
each is taken up in the modelling chapter:

* **Is the clustering genuine, or an artefact of density?** The transitivity of 0.43
is high in absolute terms, but only a comparison against models that share the
network's size and density can tell us whether it is structurally significant. The
global models (Erdős–Rényi, Barabási–Albert) and a parametric bootstrap address this.

* **Is community 2 genuinely self-contained, or does its apparent closure merely
reflect its size?** A large community accumulates internal links by size alone; only
by normalising against the number of *possible* connections can we judge its true
internal density. The stochastic blockmodel resolves this.

* **Does the degree exponent reflect a known generative mechanism?** The low value
$\alpha \approx 1.33$ is best assessed against a model that produces heavy tails by
construction — the role of Barabási–Albert.

More broadly, the descriptive analysis raises the central modelling question: the
sectors are clearly a strong organising principle, but *how much* of the network's
structure do they actually explain? The next chapter answers this by fitting a
sequence of models of increasing resolution, from global mechanisms that ignore the
communities to models that estimate connection rates within and between them, and
finally to a model that tries to recover the communities from the topology alone.

> Note: we save the enriched graph obtained through the section `02_descriptive_analysis.Rmd` that will be used for the `03_network_modelling.Rmd` part.

```{r graph-check}
vertex_attr_names(g)
edge_attr_names(g)

g = delete_vertex_attr(g, "plot_size")
g = delete_vertex_attr(g, "highlight_size")
g = delete_vertex_attr(g, "highlight_frame")
```

```{r saved-enriched-graph}
saveRDS(g, here("data", "processed", "product_space_enriched.rds"))
```