---
title: "The Product Space (HS): Modelling"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini | Beatrice Liuni | Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(here)
library(tidyverse)
library(igraph)
```

# **Network Modelling**

## **Data Loading**

```{r load-enriched-graph}
g = readRDS(here("data", "processed", "product_space_enriched.rds"))

summary(g)
```

## **Introduction and Objectives**

The descriptive analysis in Act 1 revealed a highly organized topology: an extremely
low density ($\approx 0.0068$) paired with a high global transitivity ($\approx 0.43$), a heavy-tailed
degree distribution concentrated in the heavy-machinery sector (community 2), and a
strong concentration of links within sectors (about 78% of all edges are
within-community). Descriptive statistics, however, summarize the observed patterns
without explaining the mechanisms that produce them.

The goal of this section is to move from description to a formal statistical
assessment of the structure of the Product Space. The guiding question is whether the
network can be explained by global mechanisms alone, or whether it requires the
explicit inclusion of the sector communities. We address this on two levels.

* **Global level — null and generative models.** We fit Erdős-Rényi and Barabási-Albert
models, using the size of the real network, to test whether the high clustering and
the long path lengths can be reproduced by density or by a heavy-tailed degree
sequence alone. These serve as baselines: they show what the network has *beyond*
randomness and degree.

* **Community level — modelling the sectors.** We then model the role of the
communities directly. First, an Exponential Random Graph Model (ERGM) with
dyad-independent terms tests whether sector membership predicts edge formation
(homophily). Second, a stochastic blockmodel using the 12 observed sectors as fixed
blocks estimates the rates of within- and between-community connection for each
block, which lets us revisit the question of community openness left open in the descriptive analysis.

## **Global Models**

Before modelling the community structure explicitly, we ask a more basic question:
how much of the Product Space can be explained by simple, global mechanisms that
ignore the sectors entirely? We consider two classical generative models. The
Erdős–Rényi model assumes that edges form completely at random; the Barabási–Albert
model assumes that the network grows by *preferential attachment*, where new products
attach preferentially to already well-connected ones. Neither model knows anything
about communities.

Our strategy is not simply to simulate one network from each model and compare a few
numbers. We proceed in two steps. First, we compare the global summary statistics
(transitivity, path length, maximum degree) of a simulated network against the
observed one. Second, we use a **parametric bootstrap**: we simulate many networks
from the null model, build the sampling distribution of a statistic under it, and
check whether the observed value is plausible or falls far outside its range. This
turns an informal comparison into a formal statistical assessment.

We expect both models to fail, but in informative ways: the question is *which*
features each one can and cannot reproduce, and what that failure tells us about the
forces shaping the network.

### **Erdős–Rényi: The Null Model**

The Erdős–Rényi (ER) model is the simplest random graph model and serves as our
baseline. In the $G(n, p)$ formulation, a graph has $n$ vertices and each of the
$\binom{n}{2}$ possible edges is present independently with the same probability $p$:

$$
A_{ij} \stackrel{iid}{\sim} \text{Bernoulli}(p), \quad i > j.
$$

Equivalently, in the $G(n, m)$ formulation used here, we fix the number of vertices
$n$ and edges $m$ and distribute the $m$ edges uniformly among all possible pairs.
We use $G(n, m)$ so that the simulated network has exactly the same size and density
as the observed one ($n = 866$, $m = 2532$), making the comparison fair.

This model produces homogeneous structure. Its degree distribution is approximately
Poisson (for $p = O(1/n)$), with exponentially decaying tails, and its expected local
clustering equals the edge probability, $\mathbb{E}[cl(v)] = p \approx 0.0068$, since
neighbours are connected only by chance. Both predictions clash with what we observed
in Act 1: a heavy-tailed degree distribution (Section 3.2) and a high transitivity
($cl_T(G) \approx 0.43$, Section 3.3). The ER model therefore lets us test directly
whether these features could arise from randomness alone, or whether they signal
genuine structure.

We simulate an ER graph of the same size as the Product Space and inspect its basic
properties.

```{r er-simulate}
SEED = 5836
set.seed(SEED)  

g_er = sample_gnm(n = vcount(g), m = ecount(g))

c(
  n_real = vcount(g),    n_er = vcount(g_er),
  m_real = ecount(g),    m_er = ecount(g_er),
  density_real = edge_density(g), density_er = edge_density(g_er)
)
```

### **Barabási–Albert: Preferential Attachment**

### **Barabási–Albert: Preferential Attachment**

While the Erdős–Rényi model fails on both clustering and the degree tail, the
Barabási–Albert (BA) model targets the second of these directly. Unlike ER, which is
static, BA is a *growth* model: starting from a small seed graph, the network adds
one vertex at a time, and each new vertex forms $m$ edges to existing vertices, chosen
with probability proportional to their current degree:

$$
P(\text{new vertex connects to } i) = \frac{d_i}{\sum_j d_j}.
$$

This rule, known as *preferential attachment*, means that well-connected vertices are
more likely to receive new links — a "rich-get-richer" dynamic. As the network grows,
this concentrates connections on a few vertices and produces a heavy-tailed degree
distribution that asymptotically follows a power law,

$$
P(d) \sim d^{-\gamma}, \qquad \gamma = 3,
$$

with exponent $\gamma = 3$ in the canonical model. This is the mechanism most often
invoked to explain the heavy tails observed in real networks, and it directly targets
the degree concentration we found in Section 3.2 (estimated exponent
$\alpha \approx 1.33$).

We simulate a BA network with $m = 3$ edges per new vertex, which yields about 2,600
edges, close to the observed 2,532, so that the density is comparable. Note that,
unlike $G(n, m)$, the BA model does not fix the edge count exactly, so the match is
approximate. The question we ask is whether reproducing the degree mechanism is
enough to recover the rest of the structure, in particular the high clustering that
Erdős–Rényi could not generate.

```{r ba-simulate}
set.seed(SEED)
# m = 3 edges added per new node gives ~2.6k edges, close to the real 2532
g_ba = sample_pa(n = vcount(g), m = 3, directed = FALSE)

c(
  n_ba = vcount(g_ba),
  m_ba = ecount(g_ba),
  density_ba = edge_density(g_ba)
)
```

We now compare the global properties of the two null models with those of the
observed network.

```{r global-comparison}
get_stats = function(graph, name) {
  tibble(
    Network         = name,
    Transitivity    = transitivity(graph, type = "global"),
    Avg_Path_Length = mean_distance(graph, weights = NA),
    Max_Degree      = max(degree(graph))
  )
}

bind_rows(
  get_stats(g,    "Empirical (Product Space)"),
  get_stats(g_er, "Erdős–Rényi"),
  get_stats(g_ba, "Barabási–Albert")
) |> knitr::kable(digits = 4)
```

The comparison shows that neither null model reproduces the structure of the Product
Space. Three points stand out.

The clearest failure is **clustering**. The empirical transitivity (0.434) is about
two orders of magnitude higher than that of both null models (0.006 for Erdős–Rényi,
0.020 for Barabási–Albert). The Erdős–Rényi value matches its theoretical prediction
($cl_T \approx p \approx 0.007$): in a random graph, neighbours are connected only by
chance, so almost no triangles form. Barabási–Albert, despite its hubs, also produces
very little clustering. The high transitivity of the Product Space therefore cannot
arise from randomness or from preferential attachment alone, it reflects genuine
structure that these global mechanisms do not capture.

The **degree** results are more nuanced. The empirical network has a maximum degree
of 48. Erdős–Rényi underestimates this sharply (maximum 16), as expected from its
light, Poisson-like tail. Barabási–Albert instead produces an even more extreme hub
(maximum 107): preferential attachment does generate a heavy tail, confirming that
this mechanism is relevant to the degree distribution, but in its pure form it
overshoots, creating hubs larger than any observed product. So BA captures the
*presence* of a heavy tail, but not its precise shape.

Finally, both null models have much shorter **average path lengths** (4.0 and 3.5)
than the empirical network (7.1). The Product Space is more stretched out than either
model: its products are, on average, farther apart, consistent with the
core–periphery layout seen on the map.

Taken together, the two global models fail in complementary ways. Erdős–Rényi misses
both the clustering and the hubs; Barabási–Albert recovers the hubs but not the
clustering, and exaggerates the degree concentration. Crucially, neither reproduces
the high transitivity, which is the signature of products grouping into densely
connected sectors. This is exactly the feature that a model with explicit community
structure should capture, and it motivates the community-level models that follow.



```{r er-bootstrap}
set.seed(SEED)
B = 500  # number of simulated networks

# Simulate B Erdős–Rényi graphs, record the transitivity of each
er_transitivity = replicate(B, {
  g_sim = sample_gnm(n = vcount(g), m = ecount(g))
  transitivity(g_sim, type = "global")
})

# Where does the observed value fall?
obs = transitivity(g, type = "global")
c(
  observed       = obs,
  er_mean        = mean(er_transitivity),
  er_sd          = sd(er_transitivity),
  er_max         = max(er_transitivity),
  std_distance   = (obs - mean(er_transitivity)) / sd(er_transitivity)
)
```

To put this comparison on a formal footing, we use a parametric bootstrap: we
simulate 500 Erdős–Rényi graphs of the same size as the Product Space and record the
global transitivity of each, building its sampling distribution under the null model.

The result is unambiguous. Under the ER model the transitivity averages 0.0068 — its
theoretical value $p$ — with a standard deviation of only 0.0012, and across all 500
simulations it never exceeds 0.012. The observed transitivity of 0.434 lies about
**353 standard deviations** above the null mean, and roughly 37 times higher than the
largest value the model produced in 500 attempts. The probability of observing such
clustering under random wiring is, for any practical purpose, zero.

This formalises what the summary table already suggested: the high transitivity of
the Product Space is not a by-product of its density. It is a genuine structural
signal: products cluster into tightly connected groups far more than chance allows.
The parametric bootstrap lets us reject the Erdős–Rényi model decisively, and points
to the organising principle the null models lack: the sector communities, which we
model next.

```{r ba-bootstrap}
set.seed(SEED)
B = 500

# Simulate B Barabási–Albert graphs, record the transitivity of each
ba_transitivity = replicate(B, {
  g_sim = sample_pa(n = vcount(g), m = 3, directed = FALSE)
  transitivity(g_sim, type = "global")
})

obs = transitivity(g, type = "global")
c(
  observed     = obs,
  ba_mean      = mean(ba_transitivity),
  ba_sd        = sd(ba_transitivity),
  ba_max       = max(ba_transitivity),
  std_distance = (obs - mean(ba_transitivity)) / sd(ba_transitivity)
)

```

The same conclusion holds for Barabási–Albert. Across 500 simulated BA networks the
transitivity averages 0.020, never exceeding 0.025, while the observed value of 0.434
lies about **290 standard deviations** above this mean, roughly 17 times the largest
value the model produced. So even though preferential attachment generates the hubs
that Erdős–Rényi lacks, it produces almost no clustering: the heavy tail and the
triangle-rich neighbourhoods are distinct features, and reproducing the first does
not reproduce the second. Both null models are therefore rejected on clustering, by
margins so large that randomness and preferential attachment can be ruled out as the
sole organising principles of the Product Space.





```{r blockmodel-known}
comm = V(g)$community_ref

n_q = as_tibble(table(comm)) |>
  rename(community = comm, size = n) |>
  mutate(community = as.integer(community))   # <- fix: keep integer

el = as_edgelist(g, names = FALSE)

edge_counts = tibble(
  q = pmin(comm[el[, 1]], comm[el[, 2]]),
  r = pmax(comm[el[, 1]], comm[el[, 2]])
) |>
  count(q, r, name = "edges")

pairs_tbl = expand_grid(q = sort(unique(comm)), r = sort(unique(comm))) |>
  filter(q <= r) |>
  left_join(n_q |> rename(q = community, n_q = size), by = "q") |>
  left_join(n_q |> rename(r = community, n_r = size), by = "r") |>
  mutate(
    possible = if_else(q == r, choose(n_q, 2), as.numeric(n_q) * n_r)
  )

block_tbl = pairs_tbl |>
  left_join(edge_counts, by = c("q", "r")) |>
  mutate(
    edges  = replace_na(edges, 0),
    pi_hat = round(edges / possible, 4),
    type   = if_else(q == r, "intra", "inter")
  ) |>
  select(q, r, type, edges, possible, pi_hat)

block_tbl
```


```{r blockmodel-intra}
block_tbl |>
  filter(type == "intra") |>
  left_join(n_q |> rename(q = community), by = "q") |>
  arrange(desc(pi_hat)) |>
  select(community = q, size, edges, pi_hat)
```


```{r blockmodel-summary}
block_tbl |>
  group_by(type) |>
  summarise(
    mean_pi   = round(mean(pi_hat), 4),
    median_pi = round(median(pi_hat), 4),
    max_pi    = round(max(pi_hat), 4)
  )
```


```{r blockmodel-check}
# Total edges across all blocks must equal the network's edge count
sum(block_tbl$edges)        # should be 2532
ecount(g)                   # 2532

# Intra vs inter totals must match Section 3.5.1 (1966 intra, 566 inter)
block_tbl |>
  group_by(type) |>
  summarise(total_edges = sum(edges))
```


```{r sbm-variational}
library(sbm)

# SBM works on the adjacency matrix
A = as_adjacency_matrix(g, sparse = FALSE)

# Estimate the SBM: the algorithm discovers the blocks and probabilities
set.seed(SEED)
sbm_fit = estimateSimpleSBM(
  A,
  model = "bernoulli",
  estimOptions = list(verbosity = 1, plot = FALSE, exploreMax = 12)
)
```


```{r}
# Quanti blocchi ha selezionato l'ICL?
sbm_fit$nbBlocks

# La matrice delle probabilità tra blocchi
round(coef(sbm_fit)$mean, 4)

# Le dimensioni dei blocchi trovati
table(sbm_fit$memberships)
```


```{r sbm-comparison}
tibble(
  official = factor(V(g)$community_ref),
  sbm      = factor(sbm_fit$memberships)
) |>
  count(official, sbm) |>
  pivot_wider(names_from = sbm, values_from = n, values_fill = 0)
```