---
title: "The Product Space (HS): Network Modelling"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini | Beatrice Liuni | Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

# **Network Modelling**

# **Setup and Data Loading**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(igraph)

g = readRDS(here("data", "processed", "product_space_enriched.rds"))

SEED = 5836   
```

# **Introduction and Objectives**

The descriptive analysis (`02_descriptive_analysis.Rmd`) revealed a highly organized topology: an extremely low
density ($\approx 0.0068$) paired with a high global transitivity ($\approx 0.43$), a
heavy-tailed degree distribution concentrated in the heavy-machinery sector
(community 2), and a strong concentration of links within sectors (about 78% of edges
are within-community). Descriptive statistics, however, summarize these patterns
without explaining the mechanisms that produce them.

This chapter moves from description to a formal statistical assessment. The guiding
question is not which single model "wins", but how much structure is needed to
explain the Product Space: can it be reproduced by simple global mechanisms, or
does it fundamentally require the sector communities? To answer this, we fit a
sequence of models of increasing structural resolution, all of which describe the
same quantity, the probability that an edge exists between two products, but
condition it on progressively more information.

* **Global models (no communities)**: Erdős–Rényi assumes edges form at random;
Barabási–Albert adds preferential attachment to generate a heavy-tailed degree
distribution. Here the edge probability ignores the sectors entirely. These are
baselines: they reveal what the network has *beyond* randomness and degree.

* **Community-level models (communities as a variable)**: We then condition edge
probability on the sectors. An ERGM with a homophily term tests whether sharing a
community predicts edge formation, distinguishing only *within* from *between*
sectors. A stochastic blockmodel on the 12 sectors refines this, estimating a
separate connection rate for *every pair* of communities, which also lets us revisit
the question of community openness left open in the descriptive analysis (`02_descriptive_analysis.Rmd`).

* **Latent structure (communities discovered).** Finally, a stochastic blockmodel
with *latent* blocks, fitted by variational EM, reverses the logic: instead of
assuming the sectors, it infers groups from the topology alone, and we ask whether
the recovered structure matches the official sectors.

Together these models form a single axis,  the edge probability under increasingly
rich assumptions, along which we can locate exactly where the structure of the
Product Space resides.

# **Global Models**

Before modelling the community structure explicitly, we ask a more basic question:
how much of the Product Space can be explained by simple, global mechanisms that
ignore the sectors entirely? We consider two classical generative models. The
Erdős–Rényi model assumes that edges form completely at random; the Barabási–Albert
model assumes that the network grows by *preferential attachment*, where new products
attach preferentially to already well-connected ones. Neither model knows anything
about communities.

Our strategy is not simply to simulate one network from each model and compare a few
numbers. We proceed in two steps. First, we compare the global summary statistics
(transitivity, path length, maximum degree) of a simulated network against the
observed one. Second, we use a **parametric bootstrap**: we simulate many networks
from the null model, build the sampling distribution of a statistic under it, and
check whether the observed value is plausible or falls far outside its range. This
turns an informal comparison into a formal statistical assessment.

We expect both models to fail, but in informative ways: the question is *which*
features each one can and cannot reproduce, and what that failure tells us about the
forces shaping the network.


## **Erdős–Rényi: The Null Model**

The Erdős–Rényi (ER) model is the simplest random graph model and serves as our
baseline. In the $G(n, p)$ formulation, a graph has $n$ vertices and each of the
$\binom{n}{2}$ possible edges is present independently with the same probability $p$:

$$
A_{ij} \stackrel{iid}{\sim} \text{Bernoulli}(p), \quad i > j.
$$

Equivalently, in the $G(n, m)$ formulation used here, we fix the number of vertices
$n$ and edges $m$ and distribute the $m$ edges uniformly among all possible pairs.
We use $G(n, m)$ so that the simulated network has exactly the same size and density
as the observed one ($n = 866$, $m = 2532$), making the comparison fair.

This model produces homogeneous structure. Its degree distribution is approximately
Poisson (for $p = O(1/n)$), with exponentially decaying tails, and its expected local
clustering equals the edge probability, $\mathbb{E}[cl(v)] = p \approx 0.0068$, since
neighbours are connected only by chance. Both predictions clash with what we observed
in `02_descriptive_analysis.Rmd`: a heavy-tailed degree distribution and a high transitivity
($cl_T(G) \approx 0.43$). The ER model therefore lets us test directly
whether these features could arise from randomness alone, or whether they signal
genuine structure.

We simulate an ER graph of the same size as the Product Space and inspect its basic
properties.

```{r er-simulate}
set.seed(SEED)
 g_er = sample_gnm(n = vcount(g), m = ecount(g))

c(
  n_real = vcount(g),    n_er = vcount(g_er),
  m_real = ecount(g),    m_er = ecount(g_er),
  density_real = edge_density(g), density_er = edge_density(g_er)
)
```

## **Barabási–Albert: Preferential Attachment**

The Erdős–Rényi model treats all vertices identically. The Barabási–Albert (BA) model
instead introduces heterogeneity through *growth* and *preferential attachment*.
Unlike ER, which places edges on a fixed vertex set, BA builds the network over time:
starting from a small seed graph, one vertex is added at a time, and each new vertex
forms $m$ edges to existing vertices, chosen with probability proportional to their
current degree,

$$
P(\text{new vertex attaches to } i) = \frac{d_i}{\sum_j d_j}.
$$

This "rich-get-richer" rule concentrates new links on already well-connected
vertices. As the network grows, it produces a heavy-tailed degree distribution that
asymptotically follows a power law,

$$
P(d) \sim d^{-\gamma}, \qquad \gamma = 3,
$$

with theoretical exponent $\gamma = 3$ in the canonical model. Preferential attachment
is the mechanism most often invoked to explain heavy tails in real networks, so it
directly targets the degree concentration we found in the descriptive analysis (estimated exponent
$\alpha \approx 1.33$, `02_descriptive_analysis.Rmd`). Unlike ER, however, BA makes no provision for clustering: it
builds hubs but not triangle-rich neighbourhoods.

We simulate a BA network with $m = 3$ edges per new vertex, which yields about 2,600
edges, close to the observed 2,532, so that the density is comparable. Note that,
unlike $G(n, m)$, the BA model does not fix the edge count exactly, so the match is
approximate. The question is whether reproducing the degree mechanism is enough to
recover the rest of the structure, in particular the high clustering that
Erdős–Rényi could not generate.

```{r ba-simulate}
set.seed(SEED)
g_ba = sample_pa(n = vcount(g), m = 3, directed = FALSE)

c(
  n_ba = vcount(g_ba),
  m_ba = ecount(g_ba),
  density_ba = edge_density(g_ba)
)
```

### **Comparison of Summary Statistics**

Having simulated both null models at the size of the Product Space, we compare them
with the observed network on three global statistics: the transitivity (clustering),
the average path length, and the maximum degree. Each captures a different structural
feature, so the comparison shows not just whether the models fail, but which
aspects of the network each one can or cannot reproduce.

```{r global-comparison}
get_stats = function(graph, name) {
  tibble(
    Network         = name,
    Transitivity    = transitivity(graph, type = "global"),
    Avg_Path_Length = mean_distance(graph, weights = NA),
    Max_Degree      = max(degree(graph))
  )
}

bind_rows(
  get_stats(g,    "Empirical (Product Space)"),
  get_stats(g_er, "Erdős–Rényi"),
  get_stats(g_ba, "Barabási–Albert")
) |> knitr::kable(digits = 4)
```

Neither null model reproduces the structure of the Product Space, and the way each
one fails is informative.

The clearest failure is **clustering**. The empirical transitivity (0.434) is one to
two orders of magnitude higher than both null models (0.006 for Erdős–Rényi, 0.020
for Barabási–Albert). The Erdős–Rényi value matches its theoretical prediction
($cl_T \approx p \approx 0.007$): in a random graph, neighbours are connected only by
chance, so almost no triangles form. Barabási–Albert, despite its hubs, produces
almost no clustering either. The high transitivity of the Product Space therefore
cannot arise from randomness or preferential attachment alone.

The **maximum degree** is more nuanced. The empirical network reaches 48. Erdős–Rényi
underestimates this sharply (16), as expected from its light, Poisson-like tail.
Barabási–Albert instead overshoots (68): preferential attachment does generate a
heavy tail, confirming the mechanism is relevant, but in its pure form it produces
hubs larger than any observed product. BA captures the *presence* of a heavy tail,
but not its precise shape.

Finally, both null models have much shorter **average path lengths** (4.0 and 3.6)
than the empirical network (7.1). The Product Space is more stretched out: its
products are, on average, farther apart, consistent with the core–periphery layout
seen on the map.

The summary table reports only the maximum degree, but the whole shape of the degree
distribution matters. To compare the tails directly, we plot the complementary
cumulative distribution $P(\text{degree} \geq x)$ for the three networks on a
log–log scale: this shows, for each model, how far its degree tail reaches and how
much probability mass sits at high degrees.

```{r degree-ccdf-plot}
deg_tbl = bind_rows(
  tibble(model = "Empirical",       degree = igraph::degree(g)),
  tibble(model = "Erdős–Rényi",     degree = igraph::degree(g_er)),
  tibble(model = "Barabási–Albert", degree = igraph::degree(g_ba))
) |>
  filter(degree > 0) |>
  group_by(model, degree) |> 
  summarise(count = n(), .groups = 'drop') |> 
  group_by(model) |> 
  arrange(desc(degree)) |> 
  mutate(ccdf = cumsum(count) / sum(count)) |> 
  ungroup()

ggplot(deg_tbl, aes(x = degree, y = ccdf, color = model)) +
  geom_step(linewidth = 0.9, direction = "vh") + # direction="vh" per scalini corretti
  scale_x_log10() + scale_y_log10() +
  scale_color_manual(values = c("Empirical" = "black",
                                "Erdős–Rényi" = "steelblue",
                                "Barabási–Albert" = "firebrick")) +
  labs(title = "Degree distributions: complementary CDF (log–log)",
       x = "Degree", y = "P(Degree ≥ x)", color = NULL) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

The complementary CDF makes the differences in the tails clear. The Erdős–Rényi tail
(blue) collapses early, almost no vertex exceeds degree 16, confirming that random
wiring cannot produce hubs. The empirical (black) and Barabási–Albert (red) tails are
far heavier and broadly similar to each other: both extend to high degrees, with BA
reaching a slightly larger maximum. In the mid-to-high range the empirical curve even
sits above BA, meaning the Product Space has somewhat more medium-hub products than
pure preferential attachment produces. So on the degree distribution, BA is a
reasonable approximation, far better than ER, which is consistent with preferential
attachment being a relevant mechanism for the heavy tail. The gap
between the models is not in the degree sequence but, as we saw, in the clustering.

### **Parametric Bootstrap Assessment**

The comparisons so far are informative but informal: they contrast the observed
network with a *single* simulated draw from each model. A single draw cannot tell us
whether the gap we see, especially in clustering, is a real structural difference
or just sampling variability. To settle this, we use a **parametric bootstrap**. For
each null model we simulate 500 networks of the same size as the Product Space,
compute the global transitivity of each, and build its sampling distribution under
that model. We then ask how far the observed transitivity (0.434) falls from what the
model can plausibly produce.

We focus on transitivity because it is the statistic where both null models failed
most clearly: the degree tail is at least partly captured by Barabási–Albert, but the
clustering is not reproduced by either.


```{r bootstrap-run}
set.seed(SEED)
B = 500
obs = transitivity(g, type = "global")

# One function, applied to both models
bootstrap_transitivity = function(simulate_fn) {
  replicate(B, transitivity(simulate_fn(), type = "global"))
}

er_transitivity = bootstrap_transitivity(
  function() sample_gnm(n = vcount(g), m = ecount(g))
)
ba_transitivity = bootstrap_transitivity(
  function() sample_pa(n = vcount(g), m = 3, directed = FALSE)
)

# Summary as a tidy tibble
bootstrap_summary = bind_rows(
  tibble(model = "Erdős–Rényi",     value = er_transitivity),
  tibble(model = "Barabási–Albert", value = ba_transitivity)
) |>
  group_by(model) |>
  summarise(
    mean_sim     = round(mean(value), 4),
    sd_sim       = round(sd(value), 4),
    max_sim      = round(max(value), 4),
    observed     = round(obs, 4),
    std_distance = round((obs - mean(value)) / sd(value), 1)
  )

bootstrap_summary
```

The table reports the gap as a single distance in standard deviations; the plot below
shows it directly. 

On a logarithmic scale, the two bootstrap distributions are clearly
separated: Erdős–Rényi around 0.006, Barabási–Albert around 0.02. Both sit
orders of magnitude below the observed transitivity (0.434), marked by the dashed
line.

```{r bootstrap-plot}
boot_tbl = bind_rows(
  tibble(model = "ER", transitivity = er_transitivity),
  tibble(model = "BA", transitivity = ba_transitivity)
)
obs = transitivity(g, type = "global")

ggplot(boot_tbl, aes(x = transitivity, fill = model)) +
  geom_histogram(alpha = 0.6, bins = 40, position = "identity") +
  geom_vline(xintercept = obs, linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = obs, y = Inf, label = "Observed (0.434)",
           hjust = 1.1, vjust = 2, size = 3.5) +
  scale_x_log10() +
  scale_fill_manual(values = c("ER" = "steelblue", "BA" = "firebrick")) +
  labs(title = "Bootstrap distributions of global transitivity (log scale)",
       x = "Global transitivity (log scale)", y = "Count", fill = NULL) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

The picture is unambiguous. Both null models concentrate in a narrow band well below
0.03, while the observed transitivity sits far to the right, separated from either
distribution by a large gap: about 353 standard deviations from the Erdős–Rényi mean
and 287 from the Barabási–Albert mean. No simulated network came remotely close to the
observed value.

This closes the analysis of the global models with a clear diagnosis. The two models
fail in complementary ways: Erdős–Rényi reproduces neither the hubs nor the
clustering, while Barabási–Albert captures the heavy degree tail, as the degree
distribution showed, but generates almost no clustering. What neither model can
reproduce is the **high transitivity** of the Product Space: the tendency of products
to form tightly connected groups. This is not a quantity that randomness or
preferential attachment can explain, because both treat the network as globally
homogeneous, with no notion of which products belong together.

The missing ingredient is therefore the **community structure** itself. The clustering
that the global models cannot generate is exactly what arises when products are
organised into sectors, with dense connections inside each sector and sparse ones
across them. This motivates the community-level models of the next section, which
build this structure into the model explicitly.

# **Community-Level Models**

The global models failed for a specific reason: they treat the network as
homogeneous, with no notion of which products belong together. The clustering they
cannot reproduce is precisely what arises when products are organised into sectors.
We now build this structure into the model, conditioning the probability of an edge
on the communities of the two products it would connect.

We do this in two steps of increasing resolution. First, an **exponential random
graph model (ERGM)** introduces the communities in the simplest possible way, through
a single *homophily* term that distinguishes only whether two products belong to the
same sector or to different ones. Because this term is *dyad-independent*, the model
reduces to a logistic regression on the edges and can be estimated exactly, avoiding
the degeneracy and heavy computation that affect general ERGMs. Then a **stochastic
blockmodel** refines this, estimating a separate connection rate for every pair of
sectors. The ERGM thus answers a first, coarse question: does sharing a community, in
itself, make an edge more likely?

## **Exponential Random Graph Model (ERGM)**

We begin with the exponential random graph model. An ERGM expresses the probability
of the whole network through a set of summary statistics; by choosing a statistic
that counts within-community edges, we can test directly whether community membership
drives edge formation.

### **The Model**

An exponential random graph model specifies the probability of the entire observed
graph $g$ through a set of *sufficient statistics* $T(g)$.

$$
P[G = g] = \frac{1}{\kappa(\theta)} \exp\left\{ \theta^\top T(g) \right\},
$$

where $T(g)$ is a vector of network statistics, $\theta$ the corresponding
parameters, and $\kappa(\theta)$ a normalising constant ensuring the probabilities
sum to one. The choice of $T(g)$ determines which structural features the model
targets: taking $T(g) = |E|$ (the number of edges) recovers the Erdős–Rényi model,
while adding the number of triangles would target clustering.

The statistics divide into two kinds, and the distinction is crucial here.
**Structural** statistics (such as the triangle count) make the existence of one edge
depend on others, so the likelihood involves the intractable normaliser
$\kappa(\theta)$ and must be estimated by Markov Chain Monte Carlo, an approach that
is computationally heavy and, for many parameter values, leads to *degenerate* models
that place almost all probability on the empty or the complete graph. **Dyad-
independent** statistics, by contrast, make each edge depend only on its two
endpoints. In that case the normaliser factorises and the ERGM reduces exactly to a
logistic regression on the edges:

$$
\text{logit}\, P(A_{ij} = 1) = \theta_{\text{edges}} + \delta \, \mathbb{1}[c_i = c_j],
$$

where $c_i$ denotes the community of product $i$ and the indicator $\mathbb{1}[c_i = c_j]$
is 1 when the two products share a sector. The coefficient $\delta$, that is the
**homophily** parameter (`nodematch`), measures the increase in the log-odds of an
edge when two products belong to the same community.

We deliberately restrict the model to this single dyad-independent term. It directly
encodes our question, whether sharing a sector drives connection, and, being
dyad-independent, it can be estimated exactly, free of the degeneracy and Monte Carlo
cost of structural terms. Adding a structural clustering statistic, even one designed
to be degeneracy-resistant, such as GWESP, would require Monte Carlo estimation and,
on a network as strongly and unevenly clustered as this one, is precisely the kind of
specification that tends to degenerate. We discuss this below, and leave the modelling
of clustering to the blockmodel, which represents it without structural terms.

### **Fitting the Homophily Model**

We convert the graph to the format used by the `ergm` package and fit the model in
two stages: a baseline with only the edge count, then the homophily term.

```{r ergm-setup}
library(intergraph)
library(network)
library(ergm)

g_net = asNetwork(g)
g_net
```

```{r ergm-models}
set.seed(SEED)

m1 = ergm(g_net ~ edges) # baseline, edges only

m2 = ergm(g_net ~ edges + nodematch("community_ref")) # add community homophily (dyad-independent term)

summary(m1)
summary(m2)
```

Note that both models are estimated by maximum pseudolikelihood, not MCMC: this is the practical confirmation that dyad-independent terms admit exact estimation, with none of the Monte Carlo machinery the general case requires.

The coefficients are read on the log-odds (logit) scale, so we convert them to
probabilities to interpret them. Recall the dyad-independent form of the model:

$$
\text{logit}\, P(A_{ij} = 1) = \theta_{\text{edges}} + \delta \, \mathbb{1}[c_i = c_j],
$$

so that

$$
P(A_{ij} = 1) = \text{logit}^{-1}\!\left( \theta_{\text{edges}} + \delta \,
\mathbb{1}[c_i = c_j] \right) = \frac{1}{1 + e^{-(\theta_{\text{edges}} + \delta\,
\mathbb{1}[c_i = c_j])}}.
$$

The indicator $\mathbb{1}[c_i = c_j]$ switches the homophily term on or off, giving
two regimes:

**Between sectors** ($c_i \neq c_j$, indicator $=0$):
$$
P_{\text{inter}} = \text{logit}^{-1}(\theta_{\text{edges}})
= \text{logit}^{-1}(-6.34) \approx 0.0018.
$$

**Within a sector** ($c_i = c_j$, indicator $=1$):
$$
P_{\text{intra}} = \text{logit}^{-1}(\theta_{\text{edges}} + \delta)
= \text{logit}^{-1}(-6.34 + 3.09) = \text{logit}^{-1}(-3.25) \approx 0.037.
$$

On the odds scale the homophily coefficient has a direct multiplicative reading: the
*odds* of a within-sector edge are $e^{\delta} = e^{3.09} \approx 22$ times the odds of
a between-sector edge. Sharing a community multiplies the propensity to connect by
roughly twenty-two.

The interpretation is unambiguous. The homophily coefficient is large and decisively
significant ($\delta = 3.09$, $z = 64.5$, $p < 0.001$): community membership is a
powerful predictor of edge formation. This is the formal counterpart of the
descriptive finding that 78% of edges fall within sectors. The sharp drop in AIC,
from 30352 to 25107, confirms that this single term captures a large part of the
structure that density alone misses.

It is worth recalling what an edge means here: two products are linked when the same countries tend to be competitive exporters of both (the proximity measure on which the Product Space is built), a proxy for shared productive capabilities. The strong sector homophily then has a natural economic reading: products in the same sector tend to require similar capabilities, so a country competitive in one is often competitive in the others, while crossing sectors demands different capabilities and is far rarer. We offer this interpretation with caution, since the network captures co-export proximity on a pruned backbone, not direct economic causation.

Yet this same reasoning exposes a limitation. If different sectors rest on different
capabilities, they need not be equally cohesive, and homophily, as specified here, cannot capture that: it forces a single within-sector probability ($\approx 0.037$) onto all communities alike. Given that the sectors differed greatly in size and internal density in the descriptive analysis, whether one uniform rate is adequate is exactly the question the blockmodel will address.

### **Why We Stop at Dyad-Independent Terms**

The homophily model captures *who* connects to whom at the sector level, but not
*clustering*; the tendency to form triangles that the global models failed to
reproduce. The natural extension would be to add a clustering statistic to the ERGM.

This is where dyad-independence ends and the difficulties begin. A clustering term
makes the probability of one edge depend on others, so the model can no longer be
estimated exactly: it requires Markov Chain Monte Carlo. The raw triangle count is
notoriously degenerate, triangles reinforce one another and the model collapses onto
the empty or the complete graph. The standard remedy, the *geometrically weighted
edgewise shared partner* (GWESP) term, was designed precisely to dampen this feedback
through a decay parameter that down-weights additional shared partners; lower decay
values are more resistant to degeneracy, higher ones closer to the raw triangle count.

Even so, such terms are well documented to degenerate on strongly clustered networks,
and the Product Space is an especially adverse case for two reasons established in the
descriptive analysis. First, its clustering is *extreme*: a global transitivity of
0.43 on a very sparse network means triangles are highly concentrated, the regime
where the triangle feedback is strongest. Second, the clustering is *highly
heterogeneous* across sectors, with internal densities ranging from about 0.18 in the
smallest communities to 0.007 in the largest. A GWESP term imposes a *single* global
clustering parameter on the whole network; it cannot represent that some sectors are
tightly knit while others are nearly empty inside. Asking one parameter to describe
such uneven structure is a misspecification, exactly the situation in which structural
ERGMs fail to converge, across decay values.

We therefore stop at the dyad-independent homophily term, which the ERGM estimates
exactly and interprets cleanly. The very feature that defeats a global clustering
term, that cohesion differs sharply from one sector to another, is what a blockmodel
is built to represent: rather than one clustering parameter for the whole network, it
estimates a separate connection rate for each pair of sectors. The structure of our
graph thus does not merely rule out structural ERGM terms; it points directly to the
model that comes next: the stochastic blockmodel.






# **STOCAZZI BLOCKMODEL**




```{r blockmodel-known}
comm = V(g)$community_ref

n_q = as_tibble(table(comm)) |>
  rename(community = comm, size = n) |>
  mutate(community = as.integer(community))   # <- fix: keep integer

el = as_edgelist(g, names = FALSE)

edge_counts = tibble(
  q = pmin(comm[el[, 1]], comm[el[, 2]]),
  r = pmax(comm[el[, 1]], comm[el[, 2]])
) |>
  count(q, r, name = "edges")

pairs_tbl = expand_grid(q = sort(unique(comm)), r = sort(unique(comm))) |>
  filter(q <= r) |>
  left_join(n_q |> rename(q = community, n_q = size), by = "q") |>
  left_join(n_q |> rename(r = community, n_r = size), by = "r") |>
  mutate(
    possible = if_else(q == r, choose(n_q, 2), as.numeric(n_q) * n_r)
  )

block_tbl = pairs_tbl |>
  left_join(edge_counts, by = c("q", "r")) |>
  mutate(
    edges  = replace_na(edges, 0),
    pi_hat = round(edges / possible, 4),
    type   = if_else(q == r, "intra", "inter")
  ) |>
  select(q, r, type, edges, possible, pi_hat)

block_tbl
```


```{r blockmodel-intra}
block_tbl |>
  filter(type == "intra") |>
  left_join(n_q |> rename(q = community), by = "q") |>
  arrange(desc(pi_hat)) |>
  select(community = q, size, edges, pi_hat)
```


```{r blockmodel-summary}
block_tbl |>
  group_by(type) |>
  summarise(
    mean_pi   = round(mean(pi_hat), 4),
    median_pi = round(median(pi_hat), 4),
    max_pi    = round(max(pi_hat), 4)
  )
```


```{r blockmodel-check}
# Total edges across all blocks must equal the network's edge count
sum(block_tbl$edges)        # should be 2532
ecount(g)                   # 2532

# Intra vs inter totals must match Section 3.5.1 (1966 intra, 566 inter)
block_tbl |>
  group_by(type) |>
  summarise(total_edges = sum(edges))
```


```{r sbm-variational}
library(sbm)

# SBM works on the adjacency matrix
A = as_adjacency_matrix(g, sparse = FALSE)

# Estimate the SBM: the algorithm discovers the blocks and probabilities
set.seed(SEED)
sbm_fit = estimateSimpleSBM(
  A,
  model = "bernoulli",
  estimOptions = list(verbosity = 1, plot = FALSE, exploreMax = 12)
)
```


```{r}
# Quanti blocchi ha selezionato l'ICL?
sbm_fit$nbBlocks

# La matrice delle probabilità tra blocchi
round(coef(sbm_fit)$mean, 4)

# Le dimensioni dei blocchi trovati
table(sbm_fit$memberships)
```


```{r sbm-comparison}
tibble(
  official = factor(V(g)$community_ref),
  sbm      = factor(sbm_fit$memberships)
) |>
  count(official, sbm) |>
  pivot_wider(names_from = sbm, values_from = n, values_fill = 0)
```






















Before concluding this section, it is crucial to acknowledge a fundamental theoretical trade-off shared by both the homophily ERGM and the standard Stochastic Blockmodel we just fitted. By moving from the Barabási–Albert baseline to these community-level models, we traded one structural focus for another. The BA model successfully captured degree heterogeneity (hubs) but remained blind to community boundaries. Conversely, our ERGM and standard SBM explicitly capture sectoral boundaries and measure intra/inter-group attraction, but they assume uniform connection probabilities within those blocks, effectively ignoring the heavy-tailed degree distribution. Capturing both phenomena simultaneously would require more complex model specifications capable of jointly handling group-level probabilities and individual node heterogeneity, which falls beyond the scope of our analysis. Nonetheless, isolating the community effect demonstrates mathematically what the global models lacked: sectoral homophily is a primary generative driver of the Product Space topology.