---
title: "The Product Space (HS): Network Modelling"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini | Beatrice Liuni | Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

# **Network Modelling**

# **Setup and Data Loading**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(here)
library(tidyverse)
library(igraph)

g = readRDS(here("data", "processed", "product_space_enriched.rds"))

SEED = 5836   
```

# **Introduction and Objectives**

The descriptive analysis (`02_descriptive_analysis.Rmd`) revealed a highly organized topology: an extremely low
density ($\approx 0.0068$) paired with a high global transitivity ($\approx 0.43$), a
heavy-tailed degree distribution concentrated in the heavy-machinery sector
(community 2), and a strong concentration of links within sectors (about 78% of edges
are within-community). Descriptive statistics, however, summarize these patterns
without explaining the mechanisms that produce them.

This chapter moves from description to a formal statistical assessment. The guiding
question is not which single model "wins", but how much structure is needed to
explain the Product Space: can it be reproduced by simple global mechanisms, or
does it fundamentally require the sector communities? To answer this, we fit a
sequence of models of increasing structural resolution, all of which describe the
same quantity, the probability that an edge exists between two products, but
condition it on progressively more information.

* **Global models (no communities)**: Erdős–Rényi assumes edges form at random;
Barabási–Albert adds preferential attachment to generate a heavy-tailed degree
distribution. Here the edge probability ignores the sectors entirely. These are
baselines: they reveal what the network has *beyond* randomness and degree.

* **Community-level models (communities as a variable)**: We then condition edge
probability on the sectors. An ERGM with a homophily term tests whether sharing a
community predicts edge formation, distinguishing only *within* from *between*
sectors. A stochastic blockmodel on the 12 sectors refines this, estimating a
separate connection rate for *every pair* of communities, which also lets us revisit
the question of community openness left open in the descriptive analysis (`02_descriptive_analysis.Rmd`).

* **Latent structure (communities discovered).** Finally, a stochastic blockmodel
with *latent* blocks, fitted by variational EM, reverses the logic: instead of
assuming the sectors, it infers groups from the topology alone, and we ask whether
the recovered structure matches the official sectors.

Together these models form a single axis,  the edge probability under increasingly
rich assumptions, along which we can locate exactly where the structure of the
Product Space resides.

# **Global Models**

Before modelling the community structure explicitly, we ask a more basic question:
how much of the Product Space can be explained by simple, global mechanisms that
ignore the sectors entirely? We consider two classical generative models. The
Erdős–Rényi model assumes that edges form completely at random; the Barabási–Albert
model assumes that the network grows by *preferential attachment*, where new products
attach preferentially to already well-connected ones. Neither model knows anything
about communities.

Our strategy is not simply to simulate one network from each model and compare a few
numbers. We proceed in two steps. First, we compare the global summary statistics
(transitivity, path length, maximum degree) of a simulated network against the
observed one. Second, we use a **parametric bootstrap**: we simulate many networks
from the null model, build the sampling distribution of a statistic under it, and
check whether the observed value is plausible or falls far outside its range. This
turns an informal comparison into a formal statistical assessment.

We expect both models to fail, but in informative ways: the question is *which*
features each one can and cannot reproduce, and what that failure tells us about the
forces shaping the network.


## **Erdős–Rényi: The Null Model**

The Erdős–Rényi (ER) model is the simplest random graph model and serves as our
baseline. In the $G(n, p)$ formulation, a graph has $n$ vertices and each of the
$\binom{n}{2}$ possible edges is present independently with the same probability $p$:

$$
A_{ij} \stackrel{iid}{\sim} \text{Bernoulli}(p), \quad i > j.
$$

Equivalently, in the $G(n, m)$ formulation used here, we fix the number of vertices
$n$ and edges $m$ and distribute the $m$ edges uniformly among all possible pairs.
We use $G(n, m)$ so that the simulated network has exactly the same size and density
as the observed one ($n = 866$, $m = 2532$), making the comparison fair.

This model produces homogeneous structure. Its degree distribution is approximately
Poisson (for $p = O(1/n)$), with exponentially decaying tails, and its expected local
clustering equals the edge probability, $\mathbb{E}[cl(v)] = p \approx 0.0068$, since
neighbours are connected only by chance. Both predictions clash with what we observed
in `02_descriptive_analysis.Rmd`: a heavy-tailed degree distribution and a high transitivity
($cl_T(G) \approx 0.43$). The ER model therefore lets us test directly
whether these features could arise from randomness alone, or whether they signal
genuine structure.

We simulate an ER graph of the same size as the Product Space and inspect its basic
properties.

```{r er-simulate}
set.seed(SEED)
 g_er = sample_gnm(n = vcount(g), m = ecount(g))

c(
  n_real = vcount(g),    n_er = vcount(g_er),
  m_real = ecount(g),    m_er = ecount(g_er),
  density_real = edge_density(g), density_er = edge_density(g_er)
)
```

## **Barabási–Albert: Preferential Attachment**

The Erdős–Rényi model treats all vertices identically. The Barabási–Albert (BA) model
instead introduces heterogeneity through *growth* and *preferential attachment*.
Unlike ER, which places edges on a fixed vertex set, BA builds the network over time:
starting from a small seed graph, one vertex is added at a time, and each new vertex
forms $m$ edges to existing vertices, chosen with probability proportional to their
current degree,

$$
P(\text{new vertex attaches to } i) = \frac{d_i}{\sum_j d_j}.
$$

This "rich-get-richer" rule concentrates new links on already well-connected
vertices. As the network grows, it produces a heavy-tailed degree distribution that
asymptotically follows a power law,

$$
P(d) \sim d^{-\gamma}, \qquad \gamma = 3,
$$

with theoretical exponent $\gamma = 3$ in the canonical model. Preferential attachment
is the mechanism most often invoked to explain heavy tails in real networks, so it
directly targets the degree concentration we found in the descriptive analysis (estimated exponent
$\alpha \approx 1.33$, `02_descriptive_analysis.Rmd`). Unlike ER, however, BA makes no provision for clustering: it
builds hubs but not triangle-rich neighbourhoods.

We simulate a BA network with $m = 3$ edges per new vertex, which yields about 2,600
edges, close to the observed 2,532, so that the density is comparable. Note that,
unlike $G(n, m)$, the BA model does not fix the edge count exactly, so the match is
approximate. The question is whether reproducing the degree mechanism is enough to
recover the rest of the structure, in particular the high clustering that
Erdős–Rényi could not generate.

```{r ba-simulate}
set.seed(SEED)
g_ba = sample_pa(n = vcount(g), m = 3, directed = FALSE)

c(
  n_ba = vcount(g_ba),
  m_ba = ecount(g_ba),
  density_ba = edge_density(g_ba)
)
```

### **Comparison of Summary Statistics**

Having simulated both null models at the size of the Product Space, we compare them
with the observed network on three global statistics: the transitivity (clustering),
the average path length, and the maximum degree. Each captures a different structural
feature, so the comparison shows not just whether the models fail, but which
aspects of the network each one can or cannot reproduce.

```{r global-comparison}
get_stats = function(graph, name) {
  tibble(
    Network         = name,
    Transitivity    = transitivity(graph, type = "global"),
    Avg_Path_Length = mean_distance(graph, weights = NA),
    Max_Degree      = max(degree(graph))
  )
}

bind_rows(
  get_stats(g,    "Empirical (Product Space)"),
  get_stats(g_er, "Erdős–Rényi"),
  get_stats(g_ba, "Barabási–Albert")
) |> knitr::kable(digits = 4)
```

Neither null model reproduces the structure of the Product Space, and the way each
one fails is informative.

The clearest failure is **clustering**. The empirical transitivity (0.434) is one to
two orders of magnitude higher than both null models (0.006 for Erdős–Rényi, 0.020
for Barabási–Albert). The Erdős–Rényi value matches its theoretical prediction
($cl_T \approx p \approx 0.007$): in a random graph, neighbours are connected only by
chance, so almost no triangles form. Barabási–Albert, despite its hubs, produces
almost no clustering either. The high transitivity of the Product Space therefore
cannot arise from randomness or preferential attachment alone.

The **maximum degree** is more nuanced. The empirical network reaches 48. Erdős–Rényi
underestimates this sharply (16), as expected from its light, Poisson-like tail.
Barabási–Albert instead overshoots (68): preferential attachment does generate a
heavy tail, confirming the mechanism is relevant, but in its pure form it produces
hubs larger than any observed product. BA captures the *presence* of a heavy tail,
but not its precise shape.

Finally, both null models have much shorter **average path lengths** (4.0 and 3.6)
than the empirical network (7.1). The Product Space is more stretched out: its
products are, on average, farther apart, consistent with the core–periphery layout
seen on the map.

The summary table reports only the maximum degree, but the whole shape of the degree
distribution matters. To compare the tails directly, we plot the complementary
cumulative distribution $P(\text{degree} \geq x)$ for the three networks on a
log–log scale: this shows, for each model, how far its degree tail reaches and how
much probability mass sits at high degrees.

```{r degree-ccdf-plot}
deg_tbl = bind_rows(
  tibble(model = "Empirical",       degree = igraph::degree(g)),
  tibble(model = "Erdős–Rényi",     degree = igraph::degree(g_er)),
  tibble(model = "Barabási–Albert", degree = igraph::degree(g_ba))
) |>
  filter(degree > 0) |>
  group_by(model, degree) |> 
  summarise(count = n(), .groups = 'drop') |> 
  group_by(model) |> 
  arrange(desc(degree)) |> 
  mutate(ccdf = cumsum(count) / sum(count)) |> 
  ungroup()

ggplot(deg_tbl, aes(x = degree, y = ccdf, color = model)) +
  geom_step(linewidth = 0.9, direction = "vh") + # direction="vh" per scalini corretti
  scale_x_log10() + scale_y_log10() +
  scale_color_manual(values = c("Empirical" = "black",
                                "Erdős–Rényi" = "steelblue",
                                "Barabási–Albert" = "firebrick")) +
  labs(title = "Degree distributions: complementary CDF (log–log)",
       x = "Degree", y = "P(Degree ≥ x)", color = NULL) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

The complementary CDF makes the differences in the tails clear. The Erdős–Rényi tail
(blue) collapses early, almost no vertex exceeds degree 16, confirming that random
wiring cannot produce hubs. The empirical (black) and Barabási–Albert (red) tails are
far heavier and broadly similar to each other: both extend to high degrees, with BA
reaching a slightly larger maximum. In the mid-to-high range the empirical curve even
sits above BA, meaning the Product Space has somewhat more medium-hub products than
pure preferential attachment produces. So on the degree distribution, BA is a
reasonable approximation, far better than ER, which is consistent with preferential
attachment being a relevant mechanism for the heavy tail. The gap
between the models is not in the degree sequence but, as we saw, in the clustering.

### **Parametric Bootstrap Assessment**

The comparisons so far are informative but informal: they contrast the observed
network with a *single* simulated draw from each model. A single draw cannot tell us
whether the gap we see, especially in clustering, is a real structural difference
or just sampling variability. To settle this, we use a **parametric bootstrap**. For
each null model we simulate 500 networks of the same size as the Product Space,
compute the global transitivity of each, and build its sampling distribution under
that model. We then ask how far the observed transitivity (0.434) falls from what the
model can plausibly produce.

We focus on transitivity because it is the statistic where both null models failed
most clearly: the degree tail is at least partly captured by Barabási–Albert, but the
clustering is not reproduced by either.


```{r bootstrap-run}
set.seed(SEED)
B = 500
obs = transitivity(g, type = "global")

# One function, applied to both models
bootstrap_transitivity = function(simulate_fn) {
  replicate(B, transitivity(simulate_fn(), type = "global"))
}

er_transitivity = bootstrap_transitivity(
  function() sample_gnm(n = vcount(g), m = ecount(g))
)
ba_transitivity = bootstrap_transitivity(
  function() sample_pa(n = vcount(g), m = 3, directed = FALSE)
)

# Summary as a tidy tibble
bootstrap_summary = bind_rows(
  tibble(model = "Erdős–Rényi",     value = er_transitivity),
  tibble(model = "Barabási–Albert", value = ba_transitivity)
) |>
  group_by(model) |>
  summarise(
    mean_sim     = round(mean(value), 4),
    sd_sim       = round(sd(value), 4),
    max_sim      = round(max(value), 4),
    observed     = round(obs, 4),
    std_distance = round((obs - mean(value)) / sd(value), 1)
  )

bootstrap_summary
```

The table reports the gap as a single distance in standard deviations; the plot below
shows it directly. 

On a logarithmic scale, the two bootstrap distributions are clearly
separated: Erdős–Rényi around 0.006, Barabási–Albert around 0.02. Both sit
orders of magnitude below the observed transitivity (0.434), marked by the dashed
line.

```{r bootstrap-plot}
boot_tbl = bind_rows(
  tibble(model = "ER", transitivity = er_transitivity),
  tibble(model = "BA", transitivity = ba_transitivity)
)
obs = transitivity(g, type = "global")

ggplot(boot_tbl, aes(x = transitivity, fill = model)) +
  geom_histogram(alpha = 0.6, bins = 40, position = "identity") +
  geom_vline(xintercept = obs, linetype = "dashed", linewidth = 0.8) +
  annotate("text", x = obs, y = Inf, label = "Observed (0.434)",
           hjust = 1.1, vjust = 2, size = 3.5) +
  scale_x_log10() +
  scale_fill_manual(values = c("ER" = "steelblue", "BA" = "firebrick")) +
  labs(title = "Bootstrap distributions of global transitivity (log scale)",
       x = "Global transitivity (log scale)", y = "Count", fill = NULL) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

The picture is unambiguous. Both null models concentrate in a narrow band well below
0.03, while the observed transitivity sits far to the right, separated from either
distribution by a large gap: about 353 standard deviations from the Erdős–Rényi mean
and 287 from the Barabási–Albert mean. No simulated network came remotely close to the
observed value.

This closes the analysis of the global models with a clear diagnosis. The two models
fail in complementary ways: Erdős–Rényi reproduces neither the hubs nor the
clustering, while Barabási–Albert captures the heavy degree tail, as the degree
distribution showed, but generates almost no clustering. What neither model can
reproduce is the **high transitivity** of the Product Space: the tendency of products
to form tightly connected groups. This is not a quantity that randomness or
preferential attachment can explain, because both treat the network as globally
homogeneous, with no notion of which products belong together.

The missing ingredient is therefore the **community structure** itself. The clustering
that the global models cannot generate is exactly what arises when products are
organised into sectors, with dense connections inside each sector and sparse ones
across them. This motivates the community-level models of the next section, which
build this structure into the model explicitly.

# **Community-Level Models**

The global models failed for a specific reason: they treat the network as
homogeneous, with no notion of which products belong together. The clustering they
cannot reproduce is precisely what arises when products are organised into sectors.
We now build this structure into the model, conditioning the probability of an edge
on the communities of the two products it would connect.

We do this in two steps of increasing resolution. First, an **exponential random
graph model (ERGM)** introduces the communities in the simplest possible way, through
a single *homophily* term that distinguishes only whether two products belong to the
same sector or to different ones. Because this term is *dyad-independent*, the model
reduces to a logistic regression on the edges and can be estimated exactly, avoiding
the degeneracy and heavy computation that affect general ERGMs. Then a **stochastic
blockmodel** refines this, estimating a separate connection rate for every pair of
sectors. The ERGM thus answers a first, coarse question: does sharing a community, in
itself, make an edge more likely?

## **Exponential Random Graph Model (ERGM)**

We begin with the exponential random graph model. An ERGM expresses the probability
of the whole network through a set of summary statistics; by choosing a statistic that counts within-community edges, we can test directly whether community membership drives edge formation. This is the first conditional step on the axis introduced above: we move from a single, global edge probability (Erdős–Rényi) to two: one for within-sector pairs, one for between-sector pairs.

### **The Model**

An exponential random graph model specifies the probability of the entire observed
graph $g$ through a set of *sufficient statistics* $T(g)$.

$$
P[G = g] = \frac{1}{\kappa(\theta)} \exp\left\{ \theta^\top T(g) \right\},
$$

where $T(g)$ is a vector of network statistics, $\theta$ the corresponding
parameters, and $\kappa(\theta)$ a normalising constant ensuring the probabilities
sum to one. The choice of $T(g)$ determines which structural features the model
targets: taking $T(g) = |E|$ (the number of edges) recovers the Erdős–Rényi model,
while adding the number of triangles would target clustering.

The statistics divide into two kinds, and the distinction is crucial here.
**Structural** statistics (such as the triangle count) make the existence of one edge
depend on others, so the likelihood involves the intractable normaliser
$\kappa(\theta)$ and must be estimated by Markov Chain Monte Carlo, an approach that
is computationally heavy and, for many parameter values, leads to *degenerate* models
that place almost all probability on the empty or the complete graph. **Dyad-
independent** statistics, by contrast, make each edge depend only on its two
endpoints. In that case the normaliser factorises and the ERGM reduces exactly to a
logistic regression on the edges:

$$
\text{logit}\, P(A_{ij} = 1) = \theta_{\text{edges}} + \delta \, \mathbb{1}[c_i = c_j],
$$

where $c_i$ denotes the community of product $i$ and the indicator $\mathbb{1}[c_i = c_j]$
is 1 when the two products share a sector. The coefficient $\delta$, that is the
**homophily** parameter (`nodematch`), measures the increase in the log-odds of an
edge when two products belong to the same community.

We deliberately restrict the model to this single dyad-independent term. It directly
encodes our question, whether sharing a sector drives connection, and, being
dyad-independent, it can be estimated exactly, free of the degeneracy and Monte Carlo
cost of structural terms. Adding a structural clustering statistic, even one designed
to be degeneracy-resistant, such as GWESP, would require Monte Carlo estimation and,
on a network as strongly and unevenly clustered as this one, is precisely the kind of
specification that tends to degenerate. We discuss this below, and leave the modelling
of clustering to the blockmodel, which represents it without structural terms.

### **Fitting the Homophily Model**

We convert the graph to the format used by the `ergm` package and fit the model in
two stages: a baseline with only the edge count, then the homophily term.

```{r ergm-setup}
library(intergraph)
library(network)
library(ergm)

g_net = asNetwork(g)
g_net
```

```{r ergm-models}
set.seed(SEED)

m1 = ergm(g_net ~ edges) # baseline, edges only

m2 = ergm(g_net ~ edges + nodematch("community_ref")) # add community homophily (dyad-independent term)

summary(m1)
summary(m2)
```

```{r ergm-comparison-table}
tibble(
  Model = c("Baseline (edges)", "Homophily (edges + nodematch)"),
  `Edges coefficient` = c(coef(m1)["edges"], coef(m2)["edges"]),
  `Nodematch coefficient` = c(NA, coef(m2)["nodematch.community_ref"]),
  AIC = c(AIC(m1), AIC(m2))
) |>
  knitr::kable(digits = 3)
```

Note that both models are estimated by maximum pseudolikelihood, not MCMC: this is the practical confirmation that dyad-independent terms admit exact estimation, with none of the Monte Carlo machinery the general case requires.

The coefficients are read on the log-odds (logit) scale, so we convert them to
probabilities to interpret them. Recall the dyad-independent form of the model:

$$
\text{logit}\, P(A_{ij} = 1) = \theta_{\text{edges}} + \delta \, \mathbb{1}[c_i = c_j],
$$

so that

$$
P(A_{ij} = 1) = \text{logit}^{-1}\!\left( \theta_{\text{edges}} + \delta \,
\mathbb{1}[c_i = c_j] \right) = \frac{1}{1 + e^{-(\theta_{\text{edges}} + \delta\,
\mathbb{1}[c_i = c_j])}}.
$$

The indicator $\mathbb{1}[c_i = c_j]$ switches the homophily term on or off, giving
two regimes:

**Between sectors** ($c_i \neq c_j$, indicator $=0$):
$$
P_{\text{inter}} = \text{logit}^{-1}(\theta_{\text{edges}})
= \text{logit}^{-1}(-6.34) \approx 0.0018.
$$

**Within a sector** ($c_i = c_j$, indicator $=1$):
$$
P_{\text{intra}} = \text{logit}^{-1}(\theta_{\text{edges}} + \delta)
= \text{logit}^{-1}(-6.34 + 3.09) = \text{logit}^{-1}(-3.25) \approx 0.037.
$$

On the odds scale the homophily coefficient has a direct multiplicative reading: the
*odds* of a within-sector edge are $e^{\delta} = e^{3.09} \approx 22$ times the odds of
a between-sector edge. Sharing a community multiplies the propensity to connect by
roughly twenty-two.

```{r ergm-prob-plot}
tibble(
  pair = c("Different sector", "Same sector"),
  probability = c(plogis(-6.34135), plogis(-6.34135 + 3.08996))
) |>
  ggplot(aes(x = pair, y = probability, fill = pair)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = scales::number(probability, accuracy = 0.0001)),
            vjust = -0.5, size = 4) +
  scale_fill_manual(values = c("Different sector" = "steelblue",
                               "Same sector" = "firebrick")) +
  labs(title = "Edge probability implied by the homophily model",
       subtitle = "Within-sector edges are ~22× more likely (on the odds scale)",
       x = NULL, y = "Probability of an edge") +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank())
```

The interpretation is unambiguous. The homophily coefficient is large and decisively
significant ($\delta = 3.09$, $z = 64.5$, $p < 0.001$): community membership is a
powerful predictor of edge formation. This is the formal counterpart of the
descriptive finding that 78% of edges fall within sectors. The sharp drop in AIC,
from 30352 to 25107, confirms that this single term captures a large part of the
structure that density alone misses.

It is worth recalling what an edge means here: two products are linked when the same countries tend to be competitive exporters of both (the proximity measure on which the Product Space is built), a proxy for shared productive capabilities. The strong sector homophily then has a natural economic reading: products in the same sector tend to require similar capabilities, so a country competitive in one is often competitive in the others, while crossing sectors demands different capabilities and is far rarer. We offer this interpretation with caution, since the network captures co-export proximity on a pruned backbone, not direct economic causation.

Yet this same reasoning exposes a limitation. If different sectors rest on different
capabilities, they need not be equally cohesive, and homophily, as specified here, cannot capture that: it forces a single within-sector probability ($\approx 0.037$) onto all communities alike. Given that the sectors differed greatly in size and internal density in the descriptive analysis, whether one uniform rate is adequate is exactly the question the blockmodel will address.

### **Why We Stop at Dyad-Independent Terms**

The homophily model captures *who* connects to whom at the sector level, but not
*clustering*, the tendency to form triangles that the global models failed to
reproduce. The natural extension would be to add a clustering statistic to the ERGM.

This is where dyad-independence ends and the difficulties begin. A clustering term
makes the probability of one edge depend on others, so the model can no longer be
estimated exactly: it requires Markov Chain Monte Carlo. The raw triangle count is
notoriously degenerate, triangles reinforce one another and the model collapses onto
the empty or the complete graph. The standard remedy, the *geometrically weighted
edgewise shared partner* (GWESP) term, was designed precisely to dampen this feedback
through a decay parameter that down-weights additional shared partners; lower decay
values are more resistant to degeneracy, higher ones closer to the raw triangle count.

Even so, such terms are well documented to degenerate on strongly clustered networks,
and the Product Space is an especially adverse case for two reasons established in the
descriptive analysis. First, its clustering is *extreme*: a global transitivity of
0.43 on a very sparse network means triangles are highly concentrated, the regime
where the triangle feedback is strongest. Second, the clustering is *highly
heterogeneous* across sectors, with internal densities ranging from about 0.18 in the
smallest communities to 0.007 in the largest. A GWESP term imposes a *single* global
clustering parameter on the whole network; it cannot represent that some sectors are
tightly knit while others are nearly empty inside. Asking one parameter to describe such uneven structure is a misspecification, exactly the kind of situation in which structural ERGMs are known to fail to converge, regardless of the decay chosen.

We therefore stop at the dyad-independent homophily term, which the ERGM estimates
exactly and interprets cleanly. The very feature that defeats a global clustering
term, that cohesion differs sharply from one sector to another, is what a blockmodel
is built to represent: rather than one clustering parameter for the whole network, it
estimates a separate connection rate for each pair of sectors. The structure of our
graph thus does not merely rule out structural ERGM terms; it points directly to the
model that comes next: the stochastic blockmodel.

## **Stochastic Blockmodel**

The ERGM established that community membership drives connection, but only coarsely:
it assigns one within-sector probability to all communities alike. The stochastic
blockmodel (SBM) removes that restriction. It is the natural next step on our axis,
from one global edge probability (Erdős–Rényi), to two (within vs. between, the
ERGM), to a full matrix of connection rates, one for *every pair* of sectors.

Following the course notation, the SBM is a generative model in which each vertex
belongs to one of $Q$ groups $C_1, \dots, C_Q,$ with $\alpha_q$ the probability of
belonging to group $q,$ and a pair of vertices $i \in C_q$, $j \in C_r$ is connected
with probability $\pi_{qr}$:

$$
P[\{i,j\} \in E \mid i \in C_q,\, j \in C_r] = \pi_{qr}.
$$

The matrix $\pi$ is the heart of the model. Its diagonal entries $\pi_{qq}$ are the
*within-sector* connection rates; its off-diagonal entries $\pi_{qr}$ ($q \neq r$) the
*between-sector* rates. Writing $z_{iq} = \mathbb{1}\{i \in C_q\}$ for the group
membership, the log-likelihood takes the form given in the course,

$$
\ell(\alpha, \pi; A, Z) = \sum_{i<j} \sum_{r,s} z_{ir} z_{js}\, b(A_{ij}, \pi_{rs})
+ \sum_i \sum_q z_{iq} \log \alpha_q,
\qquad b(a, \pi) = a \log \pi + (1-a)\log(1-\pi),
$$

where $b(\cdot)$ is simply the Bernoulli log-likelihood of an edge. This single
expression organises the whole section, because it can be used in two regimes that
differ only in whether the memberships $Z$ are known:

* **Known blocks.** If we *take* the 12 official sectors as the groups, $Z$ is
observed and the model reduces to estimating each $\pi_{qr}$ directly, the proportion
of realised edges among all possible pairs in each block. This refines the ERGM from
two probabilities to a full $12 \times 12$ matrix, and lets us revisit which sectors
are genuinely cohesive.

* **Latent blocks.** If instead we treat $Z$ as *unknown* and infer it from the
topology, the same likelihood must be maximised over both $\pi$ and $Z$. This is
intractable directly, so it is fitted by variational EM; we then ask whether the groups discovered from structure alone recover the
official sectors.

We take these in turn. Throughout, it helps to keep in mind what $\pi_{qr}$ means
economically. An edge encodes co-export proximity (shared productive capabilities)
so $\pi_{qq}$ measures how uniformly the products of a sector draw on the *same*
capabilities, while $\pi_{qr}$ measures how much two different sectors share. A dense
diagonal block is a technologically coherent sector; a sparse one is an internally
diverse aggregate; strong off-diagonal entries mark pairs of sectors built on related
capabilities. The blockmodel, in other words, does not just describe wiring: it maps
the capability structure of the economy.

### **Known Blocks: The 12 Sectors**

We begin with the case where the blocks are *given*: we take the 12 official sectors
as fixed groups, so the membership $Z$ is observed rather than inferred. The
likelihood then separates across blocks, and each connection rate is estimated
directly by its maximum-likelihood value, the observed proportion of edges among all
possible pairs in that block:

$$
\hat{\pi}_{qr} = \frac{e_{qr}}{n_{qr}},
$$

where $e_{qr}$ is the number of edges between sectors $q$ and $r$ and $n_{qr}$ the
number of possible such pairs. For an off-diagonal block this is $n_{qr} = n_q\, n_r$;
for a diagonal (within-sector) block it is $n_{qq} = \binom{n_q}{2}$, the number of
distinct pairs inside the sector.

Dividing by the number of *possible* pairs is what makes this estimate meaningful: it
corrects for sector size. The descriptive analysis showed that sector 2 holds the
most internal edges in absolute terms, but it is also by far the largest sector, so
it has many more possible pairs to fill. Only by normalising can we ask the right
question — not *how many* internal links a sector has, but how *densely* its products
are interconnected. This is precisely the size effect left open earlier, which the
$\hat{\pi}_{qq}$ now lets us resolve.

```{r blockmodel-known}
comm = igraph::V(g)$community_ref

n_q = as_tibble(table(comm)) |>
  rename(community = comm, size = n) |>
  mutate(community = as.integer(community)) 

el = igraph::as_edgelist(g, names = FALSE)

edge_counts = tibble(
  q = pmin(comm[el[, 1]], comm[el[, 2]]),
  r = pmax(comm[el[, 1]], comm[el[, 2]])
) |>
  count(q, r, name = "edges")

pairs_tbl = expand_grid(q = sort(unique(comm)), r = sort(unique(comm))) |>
  filter(q <= r) |>
  left_join(n_q |> rename(q = community, n_q = size), by = "q") |>
  left_join(n_q |> rename(r = community, n_r = size), by = "r") |>
  mutate(
    possible = if_else(q == r, choose(n_q, 2), as.numeric(n_q) * n_r)
  )

block_tbl = pairs_tbl |>
  left_join(edge_counts, by = c("q", "r")) |>
  mutate(
    edges  = replace_na(edges, 0),
    pi_hat = round(edges / possible, 4),
    type   = if_else(q == r, "intra", "inter")
  ) |>
  select(q, r, type, edges, possible, pi_hat)
```

```{r blockmodel-intra}
block_tbl |>
  filter(type == "intra") |>
  left_join(n_q |> rename(q = community), by = "q") |>
  arrange(desc(pi_hat)) |>
  select(community = q, size, edges, pi_hat)
```

```{r blockmodel-summary}
block_tbl |>
  group_by(type) |>
  summarise(
    mean_pi   = round(mean(pi_hat), 4),
    median_pi = round(median(pi_hat), 4),
    max_pi    = round(max(pi_hat), 4)
  )
```

```{r blockmodel-heatmap}
heat_tbl = block_tbl |>
  bind_rows(
    block_tbl |> filter(q != r) |> rename(q = r, r = q)
  )

ggplot(heat_tbl, aes(x = factor(q), y = factor(r), fill = pi_hat)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(q == r | pi_hat > 0.004, sprintf("%.3f", pi_hat), "")),
            size = 2.7) +
  scale_fill_gradient(low = "white", high = "purple", name = expression(hat(pi))) +
  scale_y_discrete(limits = rev) +
  labs(title = "Block connection rates between the 12 sectors",
       subtitle = "Diagonal = within-sector density; off-diagonal = between-sector",
       x = "Sector", y = "Sector") +
  coord_fixed() +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

The heatmap shows a clear assortative structure: the diagonal dominates, while almost
all off-diagonal cells are near-white. The full $12 \times 12$ matrix refines the
ERGM's two probabilities into block-specific rates, but the overall message is
consistent, connections form overwhelmingly within sectors. On average the
within-sector rate ($\bar{\pi}_{qq} = 0.083$) is about 26 times the between-sector
rate ($\bar{\pi}_{qr} = 0.0032$).

The diagonal, however, is far from uniform, and this is what the ERGM could not
show. Internal density ranges from 0.178 in sector 12 down to 0.007 in sector 9, a
spread of more than twenty-fold. The single within-sector probability the ERGM
estimated (0.037) was an average over these very different values; the blockmodel
reveals the heterogeneity it hid.

This resolves the question left open in the descriptive analysis. Sector 2 holds the
most internal edges in absolute terms (665), yet its internal density is only 0.021,
among the *lowest* of all sectors, and clearly pale on the diagonal of the heatmap
despite being by far the largest block (252 products). Its apparent cohesion was a
size effect: with so many products, it accumulates many internal edges while remaining
internally sparse. This agrees with its low internal transitivity found earlier, and
settles the matter: sector 2 is large and central, but not internally dense.

A striking regularity underlies this: internal density falls as sector size grows. The
smallest sectors (12, 6, 10, with 27–43 products) are the densest (0.15–0.18); the
largest (2, 11, 8, with 90–252) are among the sparsest (0.02–0.04). Recalling that an
edge encodes shared productive capabilities, this has a natural economic reading. A
small, dense sector groups products that draw on the *same* capabilities (textiles,
or paper and construction) so the countries competitive in one are competitive in
most, and internal proximity is high. A large, sparse sector such as heavy machinery
(sector 2) aggregates technologically diverse products (bearings, vehicles, engines)
that demand *different* capabilities; few countries are competitive across all of
them, so internal proximity stays low despite the sector's size. Sector 2 is central
to the Product Space not as a tight cluster but as a large, heterogeneous hub. We
offer this reading with the usual caution: the network captures co-export proximity on
a pruned backbone, not direct economic causation.

### **Latent Blocks: Variational EM**

We now reverse the logic. Instead of taking the 12 sectors as given, we treat the
block memberships $Z$ as *unknown* and infer them from the topology alone. The model
is the same SBM as before, with the same likelihood, but now both the connection
matrix $\pi$ *and* the memberships $Z$ must be estimated. The question changes
accordingly: not "how do the official sectors connect?", but "if we did not know the
sectors, would the network's structure recover them?"

The difficulty is that the likelihood now sums over all possible assignments of nodes
to blocks, a number that grows as $Q^{n}$ and is intractable for any real network.
The standard solution, which the course follows, is **variational EM** (Daudin,
Picard & Robin, 2008): it replaces the intractable likelihood with a tractable
*mean-field* approximation and alternates two steps until convergence, an E-step that
estimates, for each node, its probability of belonging to each block, and an M-step
that re-estimates $\pi$ given those memberships. The number of blocks $Q$ is itself
selected by the Integrated Completed Likelihood (ICL), which balances fit against
complexity.

```{r sbm-variational}
library(sbm)

A = igraph::as_adjacency_matrix(g, sparse = FALSE)

set.seed(SEED)
sbm_fit = estimateSimpleSBM(
  A,
  model = "bernoulli",
  estimOptions = list(verbosity = 1, plot = FALSE, exploreMax = 12)
)
```

```{r sbm-pi-heatmap}
pi_sbm = coef(sbm_fit)$mean
colnames(pi_sbm) = seq_len(ncol(pi_sbm))
rownames(pi_sbm) = seq_len(nrow(pi_sbm))
pi_tbl = as_tibble(as.table(pi_sbm), .name_repair = ~ c("block_q", "block_r", "pi")) |>
  mutate(
    block_q = factor(block_q, levels = as.character(seq_len(nrow(pi_sbm)))),
    block_r = factor(block_r, levels = as.character(seq_len(nrow(pi_sbm))))
  )
ggplot(pi_tbl, aes(x = block_q, y = block_r, fill = pi)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(pi >= 0.005, sprintf("%.3f", pi), "")), size = 2.5) +
  scale_fill_gradient(low = "white", high = "purple", name = expression(hat(pi))) +
  scale_y_discrete(limits = rev) +
  labs(title = "Connection rates between SBM-Discovered Blocks",
       x = "Block", y = "Block") +
  coord_fixed() +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

The variational EM selects **ten blocks**. The number is chosen by the Integrated
Completed Likelihood, which balances how well the model fits against how many blocks
it uses, so ten is not imposed but the resolution the data themselves prefer, close
to, yet deliberately distinct from, the twelve official sectors. That the topology
settles on a number so near twelve is already telling: left free to choose, it finds a
granularity close to the administrative one, but not identical to it.

The matrix makes the nature of these blocks clear, and it separates them into two
sharply different kinds. Nine are genuine communities: internally dense, with
within-block connection rates running from 0.77 in the tightest down to about 0.05,
and largely disconnected from one another (most off-diagonal rates are near zero). A
few exceptions stand out (blocks 1 and 3 connect at a rate of 0.30, blocks 4 and 5 at
0.08) marking the rare pairs of groups the model sees as genuinely related rather
than separate. These nine blocks are the organised, modular heart of the network.

The tenth block is something else entirely. It gathers **521 products (more than
half the entire network) into a single group whose internal density is just 0.003**,
two orders of magnitude below the others. This is not a community but a *residual*: it
absorbs the low-degree, peripheral products that the backbone leaves only loosely
attached, products that share no dense neighbourhood and therefore cannot be assigned
to any cohesive block. Its near-empty diagonal and its sheer size are the signature of
a periphery, not a sector.

What the model has recovered, then, is precisely the **core–periphery architecture**
that surfaced at every earlier stage of the analysis, but here it emerges
*endogenously*, from the topology alone, without being assumed. A dense, structured
core of nine modular groups sits embedded in a vast, amorphous periphery of weakly
connected products. The descriptive analysis suggested this shape through degree and
centrality; the latent blockmodel now confirms it as the network's fundamental
organising form.

```{r sbm-tile-plot}
contingency = tibble(
  official = factor(igraph::V(g)$community_ref),
  sbm      = factor(sbm_fit$memberships)
) |>
  count(official, sbm)

ggplot(contingency, aes(x = official, y = sbm, fill = n)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(n > 0, n, "")), size = 3) +
  scale_fill_gradient(low = "white", high = "purple", name = "Products") +
  labs(title = "Official sectors vs. SBM-discovered blocks",
       subtitle = "How the 10 latent blocks map onto the 12 official sectors",
       x = "Official sector", y = "SBM block") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

The tile plot reads the discovered blocks against the official sectors, and the
comparison is revealing precisely because the match is *partial*. When the topology is
free to redraw the groups, the twelve administrative sectors do not survive intact:
each one is either confirmed, merged, fragmented, or dissolved, and which of these
fates befalls it is itself the result.

A few sectors are **recovered as coherent groups**. Textiles (sector 4) split cleanly
into two dense blocks (32 and 21 products) with almost nothing left in the residual, 
the topology agrees that these products form tight communities, even refining the
official sector into two sub-groups. Electronics and chemicals (sectors 7 and 8)
behave differently but just as informatively: they are *merged*, both flowing into the
same block (24 and 40 products), which means the network cannot tell them apart:
their products are so densely co-exported across the two labels that they form one
structural community, not two.

Other sectors **dissolve**. The smaller, less cohesive ones (sectors 1, 5 and 9) are
absorbed almost wholesale into the residual block (19, 22 and 49 of their products
respectively), leaving little or nothing in any structured block. Their products are
too peripheral, too weakly connected, to constitute a community the model can see; as
groups, they exist in the official classification but not in the topology.

The most striking case is **heavy machinery (sector 2)**. Rather than forming a block
of its own, its 252 products *scatter* across five different blocks  (83 into one, 23
into another, then 11, 10, 8) with the remainder falling into the residual. The
largest and most central sector in the entire network is not recognised as a single
community at all; the variational EM decomposes it into the smaller coherent
sub-groups hidden inside it and discards the rest to the periphery. This is the same
internal heterogeneity the known-block model exposed through its low density (0.021),
now shown from the opposite direction: not "this sector is internally sparse", but
"this sector is not one group, it is several, plus noise". The two models, one
assuming the sectors and one ignoring them, describe the same object from both sides.

### **What the Models Tell Us Together**

We set out not to find the single "correct" model of the Product Space, but to ask how
much structure is needed to explain it. We approached this from two directions. First,
two global models that ignore the sectors entirely, Erdős–Rényi, which places edges
at random, and Barabási–Albert, which grows the network by preferential attachment,
to establish what the network has beyond randomness and degree. Then a sequence of
community-level models that estimate the same quantity, the probability of an edge, at
increasing resolution: one rate distinguishing within- from between-sector (the ERGM),
a full matrix over the known sectors (the blockmodel), and finally a matrix over groups
discovered from the topology alone (the latent SBM). Read together, they converge on
one coherent picture, and the answer to our question is unambiguous: global mechanisms
are not enough, the sector structure is essential, yet even the sectors do not tell
the whole story.

The global models fail, but informatively. Erdős–Rényi misses everything — neither the
hubs nor the clustering. Barabási–Albert does better on one axis: its
preferential-attachment mechanism reproduces the heavy degree tail almost as well as
the data, so the degree distribution is *not* where the network's structure hides. Yet
on clustering it fails just as decisively as Erdős–Rényi — the parametric bootstrap
places the observed transitivity hundreds of standard deviations beyond either null
model. This isolates the finding: what the global models cannot reproduce is not the
degree sequence but the *clustering*, and that clustering is the imprint of the
sectors. The ERGM confirms it formally — sharing a sector multiplies the odds of an
edge roughly twenty-two-fold. But the sectors are not uniform. The known-block model
shows their internal density ranging from 0.18 down to 0.007, and the latent model,
free to redraw the groups, recovers the specialised sectors cleanly while fragmenting
or dissolving the large, diverse ones. All four analyses — clustering, centrality, and
the two blockmodels — return the same verdict on the largest and most central sector,
heavy machinery: internally loose, a hub of distinct manufacturing capabilities rather
than a single tight cluster. That four methods, with entirely different assumptions,
agree on this is the strongest evidence the analysis offers.

This picture makes sense only in light of what the network *is*. The Product Space is
a pruned backbone of co-export proximities: an edge marks two products that the same
countries competitively export, a proxy for shared productive capabilities. Read this
way, the structural results map onto the capability structure of the global economy.
Specialised sectors — textiles, agriculture — are tightly knit because their products
rest on the same skills, infrastructure and institutions, so the countries that master
one master the others, and the latent model recovers them as clean blocks. Large
aggregates such as machinery are internally loose because they bundle products built
on different capabilities, which few countries command all at once — so no single
dense core exists for the model to find, and the sector fragments. The core–periphery
shape that every stage of the analysis returned — a structured core of coherent
sectors surrounded by a periphery of weakly connected products — is the topological
echo of a familiar economic logic: economies grow outward from the capabilities they
already possess, and the strongest of those relationships are exactly the ones the
backbone retains.

Before drawing the threads together, one theoretical trade-off deserves note. In
moving from the Barabási–Albert baseline to the community-level models, we traded one
structural focus for another. Barabási–Albert captured degree heterogeneity (the hubs) but was blind to community boundaries. The ERGM and the standard blockmodel do the
opposite: they capture sectoral boundaries and measure intra- versus inter-group
attraction, but they treat all products within a block as stochastically equivalent,
abstracting away the very degree heterogeneity that Barabási–Albert reproduced.
Capturing both at once (group structure *and* node-level heterogeneity) would require
a richer specification such as a degree-corrected blockmodel. We do not pursue it here:
on a pruned backbone it would inherit the same structural limits, and isolating the
community effect is what answers our question.

That scope is the limit we have respected throughout. Because the network is a
backbone, a maximum spanning tree plus the proximities above a high threshold, it
keeps only the strongest links and leaves many products with very few connections.
This is why the latent model gathers more than half the nodes into a single near-empty
residual block: the periphery is not noise, but it is partly a consequence of pruning,
and the partial recovery of the sectors must be read as a statement about this
co-export backbone, not about the full proximity matrix or direct economic causation.

Within that scope, however, the conclusion is firm and consistent across
four independent analyses: the Product Space is organised by sector, but unevenly.
Its structure is strong and recoverable where capabilities are specialised and shared,
and faint where they are diverse and dispersed. How much structure is needed to
explain the network? More than randomness, more than degree, and more than a single
uniform notion of community, exactly as much as its real, heterogeneous division into
productive sectors.




