---
title: "The Product Space (HS): Descriptive Analysis"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini, Beatrice Liuni, Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

> **A note on the backbone.** All the statistics in this analysis describe the
> distributed network, which is a *pruned backbone* of the full product space
> (a maximum spanning tree plus the links with $\varphi > 0.55$), not the complete
> proximity matrix. Weak proximity links were removed by construction, so measures
> such as density, degree, clustering, and distances refer to this strong-link
> backbone only. We keep this in mind when interpreting the results, and we note it
> explicitly where it matters.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(igraph)
library(here)
library(tidyverse)
```

# **1. Data Loading**

```{r load-graph}
g = readRDS(here("data", "processed", "product_space.rds"))
layout_xy <- readRDS(here("data", "processed", "layout_xy.rds"))
summary(g)
```

# **2. Exploratory Data Analysis**

This notebook performs the descriptive analysis of the Product Space (HS) network.
It loads the cleaned graph produced in the construction notebook and explores its
basic properties: data quality, size, connectivity, degree, clustering, and
centrality. We first check what the vertices and edges represent, whether we observe the full
system or only a part, and whether there is sampling or missing data.

```{r view-tables}
# A graph stores two tables: one for nodes, one for edges;
# We extract them and convert to tibbles for clean printing and later use
# NB: igraph:: prefix is needed because dplyr masks as_data_frame()

nodes_tbl <- igraph::as_data_frame(g, what = "vertices") |> as_tibble()
edges_tbl <- igraph::as_data_frame(g, what = "edges")    |> as_tibble()

# A first look at the node table (one row per product)
nodes_tbl |> select(name, pid, product, community_ref, size_layout) |> head(10)

# A first look at the edge table (one row per proximity link)
edges_tbl |> head(10)
```

 The edge table shows two numeric columns,
`width_layout` and `weight`, with the same values. This is on purpose. The
original value is the line *thickness* used to draw the network in the Atlas
visualization. According to the dataset documentation this thickness is
non-linear and meant for layout only, so it is **not** the true proximity score
$\varphi$ and should not be read as a real edge weight. We keep it under two names
for two reasons: `width_layout` records what the value really is (a drawing
thickness), while `weight` is the name that igraph looks for when running weighted
computations. Our main analyses treat the network as **unweighted**; we use the
weight only later, as a robustness check, to confirm that results do not depend on
this layout-only value.

## **2.1 Data Quality Check**

```{r}
# Missing values in the node and edge tables
sum(is.na(nodes_tbl))
sum(is.na(edges_tbl))

# Duplicated edges (same product pair appearing more than once)
any(which_multiple(g))

# Self-loops
any(which_loop(g))

# Isolated nodes
sum(degree(g) == 0)
```

The dataset is clean at the source: there are no missing values in either the node
or edge tables, no duplicated edges, no self-loops, and no isolated products. No
additional cleaning is needed before analysis. This also answers the *missingness*
question from Stage 1: there is no accidental missing data. The only reduction is
the backbone construction (maximum spanning tree + $\varphi > 0.55$), which removes
weak links on purpose, not by mistake.

## **2.2 Network Size and Connectivity**

```{r network-size}
# Nodes and edges
vcount(g)
ecount(g)

# Density: fraction of possible edges that actually exist
edge_density(g)

# Number of connected components
components(g)$no

# Diameter: the longest shortest path.
# weights = NA forces the unweighted version (count steps, not edge widths),
# our primary analysis given the layout-only weight.
diameter(g, weights = NA)

# Average shortest-path distance between two products
mean_distance(g, weights = NA)
```

The network has 866 products and 2,532
proximity links. Its density is very low ($\approx 0.0068$): only about 0.7% of
all possible edges are present, as expected for a pruned backbone. Despite being
sparse, the network forms a **single connected component**, so every product is
reachable from every other.

The **diameter is 21** (the longest shortest path) and the **average distance is $\approx 7.1$** steps between two products. 
We will interpret these values together with the degree and centrality results 
in the next sections, rather than drawing conclusions now.

## **2.3 Degree**

```{r degree}
deg = degree(g)
summary(deg) # it shows a numerical summary of the degree distribution

# Top 10 most connected products
nodes_tbl|>
  mutate(degree = deg)|>
  arrange(desc(degree))|>
  select(product, community_ref, degree)|>
  head(10)
```

The degree of a product is the number of other products it is directly
connected to. The summary shows a clear imbalance: the median degree is only 2,
so half of the products have two connections or fewer, while the mean is higher
(about 5.8) and the maximum reaches 48. When the mean is well above the median,
the distribution is skewed to the right: most products have few links, but a small
number act as highly connected hubs.

The ten most connected products are mostly basic manufacturing goods (forged steel
balls, transmission shafts, cast iron articles, bricks, wooden cases, rock wool),
plus T-shirts. We will examine the shape of this degree distribution more closely
in the next step, and return to the role of these hubs and their communities later.

### **2.3.1 Degree Distribution**

```{r degree-histogram}
deg_tbl = tibble(degree = deg)

ggplot(deg_tbl, aes(x = degree)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "Degree distribution",
       x = "Degree", y = "Number of products") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The histogram confirms the imbalance seen in the summary.
The large majority of products have a very low degree, with a tall peak at degree 1
(about 285 products) that drops off quickly. To the right there is a long, thin
tail of products with much higher degree, up to the single product with degree 48.
This long-tailed shape, with many low-degree nodes and few high-degree hubs, is
typical of many real-world networks. We look at it on a log-log scale next to see
its shape more clearly.

### **2.3.2 Degree Distribution on a log-log scale**

```{r degree-loglog}
# Frequency of each degree value
deg_dist = as.data.frame(table(deg)) |>
  rename(degree = deg, freq = Freq) |>
  mutate(degree = as.numeric(as.character(degree)))

# Log-log plot: log frequency vs log degree
ggplot(deg_dist, aes(x = log(degree), y = log(freq))) +
  geom_point(color = "darkblue") +
  labs(title = "Degree distribution (log-log scale)",
       x = "log(Degree)", y = "log(Frequency)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The histogram shows a heavy tail, but
not its precise shape. The log-log plot helps with this. If the degree
distribution follows a power law, $P(k) \propto k^{-\alpha}$, then taking logs of
both sides gives a straight line:

$\log P(k) = c - \alpha \, \log k$

so a power law appears as a descending straight line on a log-log scale, with
slope $-\alpha$.

In our plot the points do fall roughly along a descending line, which is
consistent with a power-law (or *scale-free*) degree distribution. The fit is
cleaner on the left (low degrees), where each degree value is shared by many
products, and noisier on the right (high degrees), where only a few products takeeach value and the frequencies are estimated less reliably. We will return to
whether a power-law mechanism such as preferential attachment can explain this
shape when we compare the network to generative models.

### **2.3.3 Fitting the Degree Distribution**

The log-log plot suggests a roughly straight, descending line. To describe this
more precisely, we fit a linear model to the log-frequency against the log-degree.
This gives a numerical estimate of the slope, which corresponds to the power-law
exponent $-\alpha$, and an $R^2$ value telling us how well a straight line
describes the points.

This is a **descriptive** fit of the observed shape, not a test of any generative
model: we are measuring the slope of the line we already see, not yet asking which
mechanism produced it. We return to generative mechanisms such as preferential
attachment in the modelling part of the project.

```{r degree-fit}
fit = lm(log(freq) ~ log(degree), data = deg_dist)
summary(fit)
```

The linear fit on the log-log scale gives a
slope of about $-1.40$ for `log(degree)`, with a very small p-value, so the
descending relationship is clear and well estimated. Following the power-law form
$\log P(k) = c - \alpha \log k$, this slope corresponds to an exponent
$\alpha \approx 1.40$.

This value is somewhat lower than the range often reported for scale-free networks
(usually between 2 and 3). We note this only as a description of the observed
shape, as mentioned above, our network is a pruned backbone, which affects the
tail, and we leave conclusions about the underlying mechanism to the modelling
part, where we compare the network to generative models.

## **Clustering**