---
title: "The Product Space (HS): Descriptive Analysis"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini, Beatrice Liuni, Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

> **A note on the backbone.** All the statistics in this analysis describe the
> distributed network, which is a *pruned backbone* of the full product space
> (a maximum spanning tree plus the links with $\varphi > 0.55$), not the complete
> proximity matrix. Weak proximity links were removed by construction, so measures
> such as density, degree, clustering, and distances refer to this strong-link
> backbone only. We keep this in mind when interpreting the results, and we note it
> explicitly where it matters.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(igraph)
library(here)
library(tidyverse)
```

# **1. Data Loading**

```{r load-graph}
g = readRDS(here("data", "processed", "product_space.rds"))
layout_xy = readRDS(here("data", "processed", "layout_xy.rds"))
summary(g)
```

# **2. Exploratory Data Analysis**

This notebook performs the descriptive analysis of the Product Space (HS) network.
It loads the cleaned graph produced in the construction notebook and explores its
basic properties: data quality, size, connectivity, degree, clustering, and
centrality. We first check what the vertices and edges represent, whether we observe the full
system or only a part, and whether there is sampling or missing data.

```{r view-tables}
# A graph stores two tables: one for nodes, one for edges;
# We extract them and convert to tibbles for clean printing and later use
# NB: igraph:: prefix is needed because dplyr masks as_data_frame()

nodes_tbl = igraph::as_data_frame(g, what = "vertices") |> as_tibble()
edges_tbl = igraph::as_data_frame(g, what = "edges")    |> as_tibble()

# A first look at the node table (one row per product)
nodes_tbl |> select(name, pid, product, community_ref, size_layout) |> head(10)

# A first look at the edge table (one row per proximity link)
edges_tbl |> head(10)
```

 The edge table shows two numeric columns,
`width_layout` and `weight`, with the same values. This is on purpose. The
original value is the line *thickness* used to draw the network in the Atlas
visualization. According to the dataset documentation this thickness is
non-linear and meant for layout only, so it is **not** the true proximity score
$\varphi$ and should not be read as a real edge weight. We keep it under two names
for two reasons: `width_layout` records what the value really is (a drawing
thickness), while `weight` is the name that igraph looks for when running weighted
computations. Our main analyses treat the network as **unweighted**; we use the
weight only later, as a robustness check, to confirm that results do not depend on
this layout-only value.

## **2.1 Data Quality Check**

```{r data-quality}
# Missing values in the node and edge tables
sum(is.na(nodes_tbl))
sum(is.na(edges_tbl))

# Duplicated edges (same product pair appearing more than once)
any(which_multiple(g))

# Self-loops
any(which_loop(g))

# Isolated nodes
sum(degree(g) == 0)
```

The dataset is clean at the source: there are no missing values in either the node
or edge tables, no duplicated edges, no self-loops, and no isolated products. No
additional cleaning is needed before analysis. This also answers the *missingness*
question from Stage 1: there is no accidental missing data. The only reduction is
the backbone construction (maximum spanning tree + $\varphi > 0.55$), which removes
weak links on purpose, not by mistake.

## **2.2 Network Size and Connectivity**

```{r network-size}
# Nodes and edges
vcount(g)
ecount(g)

# Density: fraction of possible edges that actually exist
edge_density(g)

# Number of connected components
components(g)$no

# Diameter: the longest shortest path.
# weights = NA forces the unweighted version (count steps, not edge widths),
# our primary analysis given the layout-only weight.
diameter(g, weights = NA)

# Average shortest-path distance between two products
mean_distance(g, weights = NA)
```

The network has 866 products and 2,532
proximity links. Its density is very low ($\approx 0.0068$): only about 0.7% of
all possible edges are present, as expected for a pruned backbone. Despite being
sparse, the network forms a **single connected component**, so every product is
reachable from every other.

The **diameter is 21** (the longest shortest path) and the **average distance is $\approx 7.1$** steps between two products. 
We will interpret these values together with the degree and centrality results 
in the next sections, rather than drawing conclusions now.

## **2.3 Degree**

```{r degree}
deg = degree(g)
summary(deg) # it shows a numerical summary of the degree distribution

# Top 10 most connected products
nodes_tbl|>
  mutate(degree = deg)|>
  arrange(desc(degree))|>
  select(product, community_ref, degree)|>
  head(10)
```

The degree of a product is the number of other products it is directly
connected to. The summary shows a clear imbalance: the median degree is only 2,
so half of the products have two connections or fewer, while the mean is higher
(about 5.8) and the maximum reaches 48. When the mean is well above the median,
the distribution is skewed to the right: most products have few links, but a small
number act as highly connected hubs.

The ten most connected products are mostly basic manufacturing goods (forged steel
balls, transmission shafts, cast iron articles, bricks, wooden cases, rock wool),
plus T-shirts. We will examine the shape of this degree distribution more closely
in the next step, and return to the role of these hubs and their communities later.

### **2.3.1 Degree Distribution**

```{r degree-histogram}
deg_tbl = tibble(degree = deg)

ggplot(deg_tbl, aes(x = degree)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "Degree distribution",
       x = "Degree", y = "Number of products") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The histogram confirms the imbalance seen in the summary.
The large majority of products have a very low degree, with a tall peak at degree 1
(about 285 products) that drops off quickly. To the right there is a long, thin
tail of products with much higher degree, up to the single product with degree 48.
This long-tailed shape, with many low-degree nodes and few high-degree hubs, is
typical of many real-world networks. We look at it on a log-log scale next to see
its shape more clearly.

### **2.3.2 Degree Distribution on a log-log scale**

```{r degree-loglog}
# Frequency of each degree value
deg_dist = as.data.frame(table(deg)) |>
  rename(degree = deg, freq = Freq) |>
  mutate(degree = as.numeric(as.character(degree)))

# Log-log plot: log frequency vs log degree
ggplot(deg_dist, aes(x = log(degree), y = log(freq))) +
  geom_point(color = "darkblue") +
  labs(title = "Degree distribution (log-log scale)",
       x = "log(Degree)", y = "log(Frequency)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The histogram shows a heavy tail, but
not its precise shape. The log-log plot helps with this. If the degree
distribution follows a power law, $P(k) \propto k^{-\alpha}$, then taking logs of
both sides gives a straight line:

$$
\log P(k) = c - \alpha \, \log k,
$$

so a power law appears as a descending straight line on a log-log scale, with
slope $-\alpha$.

In our plot the points do fall roughly along a descending line, which is
consistent with a power-law (or *scale-free*) degree distribution. The fit is
cleaner on the left (low degrees), where each degree value is shared by many
products, and noisier on the right (high degrees), where only a few products take each value and the frequencies are estimated less reliably. We will return to
whether a power-law mechanism such as preferential attachment can explain this
shape when we compare the network to generative models.

### **2.3.3 Fitting the Degree Distribution**

# we do not need the linear model (lm, just the poisson)

The log-log plot suggests a roughly straight, descending line. To describe this
more precisely, we fit a linear model to the log-frequency against the log-degree.
This gives a numerical estimate of the slope, which corresponds to the power-law
exponent $-\alpha$.

This is a **descriptive** fit of the observed shape, not a test of any generative
model: we are measuring the slope of the line we already see, not yet asking which
mechanism produced it. We return to generative mechanisms such as preferential
attachment in the modelling part of the project.

```{r degree-fit}
fit = lm(log(freq) ~ log(degree), data = deg_dist)
summary(fit)
```

The linear fit on the log-log scale gives a slope of about $-1.40$ for
`log(degree)`, with a very small p-value, so the descending relationship is clear
and well estimated. Following the power-law form $\log P(k) = c - \alpha \log k$,
this slope corresponds to an exponent $\alpha \approx 1.40$.

This value is somewhat lower than the range often reported for scale-free networks. We note this only as a description of the observed shape:
as mentioned above, our network is a pruned backbone, which affects the tail. We
leave conclusions about the underlying mechanism to the modelling part, where we
compare the network to generative models.

The linear fit above is the simplest way to estimate the slope, but it has a known
weakness: it takes the logarithm of the frequencies and treats them as if they were
normal, while they are actually **counts** (how many products have each degree). A
more appropriate model for count data is a **Poisson regression**, which models the
counts directly without logging them. We fit it and compare the two estimates.

```{r degree-glm}
fit_glm = glm(freq ~ log(degree), family = poisson, data = deg_dist)
summary(fit_glm)
```

The two methods agree closely. The linear fit on the
log-log scale gave a slope of about $-1.40$, and the Poisson regression gives
$-1.33$ (z = $-43$, p < 0.001), so both point to an exponent of roughly
$\alpha \approx 1.3$–$1.4$. The fact that the simpler linear fit and the more
appropriate count model produce nearly the same estimate means the result is
robust: it does not depend on the choice of method. The Poisson model also shows a
large drop from the null deviance (2032) to the residual deviance (46), confirming
that degree explains the frequencies well.

Both estimates place the exponent on the low side. We do not compare it to a fixed
external reference here; instead, we will contrast it with the exponent produced by
a simulated Barabási-Albert model in the modelling part, which grounds the
comparison in our own analysis. As noted above, our network is also a pruned
backbone, which affects the tail.

## **2.4 Clustering and Transitivity**

After looking at the degree of each product, we ask whether
the neighbours of a product are also connected to each other, i.e. whether the
network forms *triangles*.

For a vertex $v$, let $\tau_\triangle(v)$ be the number of triangles involving $v$,
and $\tau_3(v) = \binom{d_v}{2}$ the number of connected triples (wedges) centred
on $v$. The **local clustering coefficient** is

$$
cl(v) = \frac{\tau_\triangle(v)}{\tau_3(v)},
$$

defined for vertices with degree at least 2. It measures, among the pairs of
neighbours of $v$, how many are actually connected.

These local values can be aggregated to the whole network in two ways:

$$
cl(G) = \frac{1}{|V'|}\sum_{v \in V'} cl(v); 
\qquad cl_T(G) = \frac{\sum_{v \in V'}\tau_\triangle(v)}{\sum_{v \in V'}\tau_3(v)}.
$$

The first is a simple average of the local coefficients; the second, the
**transitivity**, is a weighted average. Following Bollobás and Riordan, we use the
transitivity $cl_T(G)$ as our main measure, since the weighting makes it less
sensitive to low-degree vertices. Values close to 1 indicate a highly clustered
network; values close to 0 indicate few triangles.

```{r clustering}
# Transitivity: cl_T(G)
transitivity(g, type = "global")

# Clustering Coefficient: cl(G)
transitivity(g, type = "average")
```

The transitivity is $cl_T(G) \approx 0.43$ and the
simple clustering coefficient is $cl(G) \approx 0.28$. Both are moderate-to-high
values, showing that the network forms a fair number of triangles: the neighbours
of a product are often connected to each other. This points to the presence of
dense groups of products, an early sign of community structure that we examine
later. The value is notable given that we work with a pruned backbone, where
removing weak links also removes potential triangles.

The two measures also differ ($cl_T(G) > cl(G)$). In the transitivity, each vertex
$v$ is weighted by $\tau_3(v) = \binom{d_v}{2}$, the number of connected triples
centred on it. Since this grows quickly with degree, hubs receive much larger
weights than peripheral nodes. The fact that the weighted measure is higher than
the simple average therefore suggests that the more connected products (the hubs)
tend to be more clustered than the peripheral ones: their neighbours are also
often linked to one another.

Overall, the network shows a clear tendency to form triangles, and this clustering
is concentrated around the hubs. Together with the heavy-tailed degree
distribution seen earlier, this gives a first, consistent picture: a few highly
connected, densely interconnected products alongside many peripheral ones. We test
whether this translates into well-defined communities in the modelling part.

## **2.5 Centrality**

The degree section asked *how many* links a product has. We now ask a broader
question: **how important is a product in the network?** This is the idea of
*centrality*. There is no single definition of importance, so rather than pick one
measure and call it "the" importance, we consider several standard measures, each
built on a different idea of what makes a vertex important, and then select those
that are most informative for *this* particular network.

We consider four measures:

- **Degree centrality** counts direct neighbours. A product is important if it is
  connected to many others. We already computed this in Section 2.3, so here we use
  it as a baseline for comparison.
- **Closeness centrality** is based on *distance*. A product is important if it is
  close, on average, to all other products through short paths.
- **Betweenness centrality** is based on *flow*. A product is important if it lies
  on many shortest paths between other products, acting as a bridge.
- **Eigenvector centrality** is based on *feedback*. A product is important if it is
  connected to other important products.

For a vertex $v \in V$:

**Closeness centrality** is the inverse of the total distance from $v$ to all
other vertices,

$$
c_{Cl}(v) = \frac{1}{\sum_{u \in V} \mathrm{dist}(v, u)},
$$

where $\mathrm{dist}(v, u)$ is the shortest-path distance between $u$ and $v$. A
small total distance gives a large closeness. It is only well-defined on a
connected graph; ours is a single connected component, so this is not a problem.

**Betweenness centrality** counts how often $v$ sits on shortest paths between
other vertices,

$$
c_B(v) = \sum_{s \neq t \neq v \in V} \frac{\sigma(s, t \mid v)}{\sigma(s, t)},
$$

where $\sigma(s, t)$ is the number of shortest paths between $s$ and $t$, and
$\sigma(s, t \mid v)$ is the number of those that pass through $v$. A product with
high betweenness connects parts of the network that would otherwise be far apart.

**Eigenvector centrality** (Bonacich) defines importance recursively: the
importance of $v$ is proportional to the sum of the importances of its neighbours,

$$
c_{Ei}(u) = \alpha \sum_{\{u, v\}} c_{Ei}(v),
$$

which in matrix form is the eigenvector problem for the adjacency matrix $A$,

$$
A\,\vec{c}_{Ei} = \alpha^{-1}\,\vec{c}_{Ei}.
$$

The centrality vector is the leading eigenvector of $A$. All these measures are
computed on the pruned backbone, since they depend on the shortest paths within it.

### **2.5.1 Choosing the Centrality Measures**

Before relying on these measures, we check which of them are actually informative
for this network. A measure is useful only if it **discriminates** between
vertices: if it gives almost the same value to every product, it cannot separate
central from peripheral ones. We also want to avoid **redundancy**: if two measures
rank products in nearly the same way, keeping both adds little.

We look at this in two steps. First, we compare the distributions of the four
measures, rescaled to a common 0–1 range, to see which spread the products out.
Second, we plot degree against the other measures to see which carry information
that degree alone does not.

```{r centrality-explore}
# Compute the four candidate measures (unweighted)
clos = closeness(g, weights = NA)
betw = betweenness(g, weights = NA)
eig  = eigen_centrality(g, weights = NA)$vector

# Put them in a tibble, rescaled to 0-1 so they share a common axis
cent_explore = tibble(
  degree      = deg  / max(deg),
  closeness   = clos / max(clos),
  betweenness = betw / max(betw),
  eigenvector = eig  / max(eig)
)
```

```{r centrality-distributions}
cent_explore |>
  pivot_longer(everything(), names_to = "measure", values_to = "value") |>
  ggplot(aes(x = measure, y = value)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(title = "Distribution of centrality measures (rescaled to 0-1)",
       x = NULL, y = "Rescaled centrality") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```


The boxplots reveal four very different shapes.
*Closeness* has the widest, most balanced spread: its values fill the upper part
of the scale (box roughly 0.54–0.77) with no outliers,
so it separates products across a broad range. *Degree* and *betweenness* are both
concentrated near zero with long upper tails of outliers, the classic heavy-tailed
shape; betweenness in particular has many outliers spread continuously up to the
maximum, so the bridging products form a sizeable group rather than a handful of
isolated cases. *Eigenvector* is the most extreme: its box collapses onto zero and
only a few outliers jump to the top, splitting products sharply into a near-zero
majority and a small élite.

```{r centrality-pairs}
cent_explore |>
  pivot_longer(c(closeness, betweenness, eigenvector),
               names_to = "measure", values_to = "value") |>
  ggplot(aes(x = degree, y = value)) +
  geom_point(alpha = 0.4, color = "darkblue") +
  facet_wrap(~ measure, scales = "free_y") +
  labs(title = "Degree vs other centrality measures",
       x = "Degree (rescaled)", y = "Other measure (rescaled)") +
  theme_minimal()
```

The scatter plots against degree clarify what each measure adds. For *betweenness*,
the points form no clear line: products with the same degree can have very
different betweenness, and high-degree products are not always strong bridges, so
betweenness is independent of degree and captures a distinct bridging role. For
*closeness*, there is a dense vertical band of low-degree products whose closeness
ranges all the way from about 0.35 to 0.9, while high-degree products always have
high closeness; closeness therefore adds information precisely among the peripheral
products, distinguishing well-placed ones from truly isolated ones with the same
degree. For *eigenvector*, the points trace a curve that stays near zero until
degree is high and then rises steeply: eigenvector is essentially a function of
degree in this network and adds little.

We therefore keep **degree**, **betweenness**, and **closeness** as our main
measures and set eigenvector centrality aside as largely redundant with degree.
This is the opposite of our initial expectation, we had thought closeness might be
the weak measure on such a sparse, elongated network, which shows the value of
checking the measures on the data before relying on them.

### **2.5.2 Analysis of the selected Measures**

We now compute the three selected measures (degree, betweenness, closeness), store
them as vertex attributes of the network, and examine which products they identify
as central.

```{r centrality-compute}
# Store the three selected measures as vertex attributes
V(g)$degree      = deg
V(g)$betweenness = betw
V(g)$closeness   = clos

# Confirm the attributes are now part of the graph
vertex_attr_names(g)
```

The three measures are now part of the network object, alongside the original
attributes. We build tables to rank and compare the products.

```{r centrality-top-betweenness}
cent_tbl = tibble(
  product       = V(g)$product,
  community_ref = V(g)$community_ref,
  degree        = V(g)$degree,
  betweenness   = V(g)$betweenness,
  closeness     = V(g)$closeness
)

# Top 10 by betweenness (the bridging role)
cent_tbl |> arrange(desc(betweenness)) |>
  select(product, community_ref, betweenness, degree) |> head(10)
```

```{r centrality-top-closeness}
# Top 10 by closeness (the short-reach role)
cent_tbl |> arrange(desc(closeness)) |>
  select(product, community_ref, closeness, degree) |> head(10)
```

The two top-ten lists differ in a revealing way. In the betweenness ranking, three
of the top ten products have a very low degree (electrical machinery parts, degree
6; spark-ignition aircraft engines, degree 3; dried potatoes, degree 4): they lie
on many shortest paths despite having few links, acting as pure bridges between
otherwise separated parts of the network. In the closeness ranking, by contrast,
all ten top products have medium-to-high degree (between 22 and 48): no low-degree
product reaches the top of closeness. The forged steel balls (degree 48) rank first
in both, as the single dominant hub.

```{r centrality-scatter}
ggplot(cent_tbl, aes(x = degree, y = betweenness)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  labs(title = "Degree vs betweenness",
       x = "Degree", y = "Betweenness") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

```{r centrality-scatter-closeness}
ggplot(cent_tbl, aes(x = degree, y = closeness)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  labs(title = "Degree vs closeness",
       x = "Degree", y = "Closeness") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

# be more focus on the interpetations and check better which products belomngs to communities and which are the communities

The two scatter plots explain this. Degree versus betweenness is a dispersed cloud
with no clear trend: products with the same degree can have very different
betweenness, and high degree does not guarantee high betweenness. Degree versus
closeness, instead, has a clear shape: closeness rises steeply with degree at first
and then flattens towards an upper limit, so high closeness is only reached by
products with at least medium degree, while low-degree products spread over a wide
range of closeness values.

This contrast shows that the two measures capture genuinely different things.
Betweenness identifies bridging products regardless of how many links they have,
which is why low-degree bridges can rank at the very top. Closeness, on the other
hand, rewards products that are well embedded in the network, and being well
embedded requires a reasonable number of connections; this is why it tracks degree
more closely and never places a sparsely connected product at the top.

Overall, the centrality analysis points to two distinct kinds of important
products: well-connected, well-embedded products that score high on both degree
and closeness (such as the forged steel balls and transmission shafts), and pure
bridges that score high on betweenness despite a low degree (such as aircraft
engines and dried potatoes). No single measure captures both roles, which confirms
that keeping several measures was the right choice. We also note that most of the
top products by both measures belong to community 2; we do not interpret this here,
but return to community structure in the modelling part.

## **2.6 Network Visualization**

So far we have described the network with numbers: degree, clustering, and
centrality. We now *look* at it. The goal of visualization is to embed the graph,
a combinatorial object, into a two-dimensional space, so that its structure
becomes visible. This embedding is not unique: it requires choices about where to
place the vertices (the layout), and about how to use colour and size to show node
attributes.

Usually the layout has to be computed with an algorithm, such as Kamada-Kawai or
Fruchterman-Reingold, which place vertices by treating edges as physical forces.
For a network of our size (866 products) such algorithms often produce a
"hairball", a dense ball of points and lines that is hard to read. We avoid this
problem because the dataset already provides **canonical coordinates**: the
official Product Space layout, in which related products are placed close together.
We use these coordinates directly, so the positions are meaningful and consistent
with the way the Product Space is usually shown, rather than the result of an
arbitrary algorithm.

We build the visualization in two steps, following the "decorating graphs" approach:
first a plain map showing the structure, then a decorated map where colour encodes
the product communities and size encodes importance.

```{r network-base, fig.width=10, fig.height=10}
plot(g,
     layout = layout_xy,        
     vertex.label = NA,          
     vertex.size = 2,            
     vertex.color = "grey40",
     edge.color = "grey65",      
     main = "Product Space — Base Map")
```

Using the canonical coordinates, the network already shows a clear
structure. There is a dense central band where many products are packed closely
together and heavily connected (the darker area of overlapping edges), and a sparser
periphery where products are more spread out and linked by thin chains. A few small
groups hang off the bottom, connected to the rest by single links. This is the
typical shape of the Product Space: a connected core of closely related products
surrounded by more isolated ones. The map confirms visually what the numbers
suggested earlier (low density, long distances, a few hubs), and we now add colour
and size to read it in more detail.

```{r network-decorated, fig.width=10, fig.height=10}
V(g)$plot_size = 1.5 + 3 * sqrt(V(g)$degree) / max(sqrt(V(g)$degree))

plot(g,
     layout = layout_xy,
     vertex.label = NA,
     vertex.size = V(g)$plot_size,
     vertex.color = V(g)$color,
     vertex.frame.color = NA,
     edge.color = "grey65",
     edge.width = 0.3,
     main = "Product Space — Communities and Hubs")
```

This single map ties together everything the numerical
analysis found so far. Three features stand out.

First, **community structure**. The communities are not mixed across the map: each
colour occupies its own region (purple in the lower left, cyan in the upper left,
the large blue group through the centre, red and green on the right). Products that
the dataset assigns to the same community are also placed close together and densely
linked, which is the visual counterpart of the high transitivity ($cl_T \approx 0.43$) 
we measured: the triangles that drive that clustering are concentrated inside
these coloured regions.

Second, **core–periphery organisation**. A dense central band holds most of the
large, highly connected products, while the periphery thins out into scattered
products and a few chains that hang off the bottom by single links. This is exactly
what the global numbers anticipated: a low density (0.0068) with a large diameter
(21) and long average distance (7.1) only make sense if the network is stretched
between a compact core and a far-reaching periphery, rather than being a uniform
blob.

Third, **the position of the hubs**. The largest nodes are not spread evenly: they
sit in the centre, inside the densest communities. This matches the centrality
results, where the same products ranked high on degree and closeness, and it is why
the weighted transitivity exceeded the simple one: the well-connected core products
are also the most clustered.

The largest and most central community is the blue one, which runs through the
middle of the map and connects to most others. We do not yet name the sectors these
communities represent, nor test whether these colour groups coincide with what a
detection algorithm would recover: the official communities shown here become the
benchmark for that comparison in the modelling part.

## **2.7 Derived Edge Attributes**

So far our derived measures described vertices. We now compute two attributes of
the *edges*, to describe the links themselves rather than the products.

- **Edge betweenness** is the betweenness applied to edges: for each link, the
  number of shortest paths in the network that pass through it. A link with high
  edge betweenness is a critical bridge — removing it would lengthen or cut the
  connection between parts of the network.
- **Intra/inter-community flag** marks, for each link, whether it connects two
  products of the same community or of two different communities. In short, an edge is *intra* when both products it connects belong to the same community, and *inter* when they belong to different ones.

We expect these two to be related: links *between* communities should tend to have
higher edge betweenness, since they are the few bridges connecting otherwise dense
groups, so many shortest paths are forced through them. We check this.

```{r edge-attributes}
# Edge betweenness, stored as an edge attribute
E(g)$betweenness = edge_betweenness(g, weights = NA)

# Intra vs inter-community flag.
# For each edge, compare the community of its two endpoints.
el = as_edgelist(g, names = FALSE)          
comm = V(g)$community_ref
E(g)$cross_community = ifelse(
  comm[el[, 1]] == comm[el[, 2]], "intra", "inter"
)

edge_attr_names(g)
```

```{r edge-counts}
# How many links stay within communities vs cross between them
table(E(g)$cross_community)
round(prop.table(table(E(g)$cross_community))*100, 2)
```

Of the 2,532 links, 1,966 (77.65%) stay within a community and only
566 (22.35%) cross between communities. Most connections therefore join products
of the same community, while links between communities are a minority. This fits
the high transitivity and the well-separated coloured regions seen on the map:
products link mainly within their own group, and only a few links bridge across
groups. We now check whether these few between-community links are also the ones
that carry the most shortest paths.

We now bring the two edge attributes together. A common idea in network analysis is
that between-community links should carry more shortest paths than within-community
ones, because they act as the few bridges connecting otherwise dense groups. Within
a community, instead, there are many alternative short paths, so each single link
should carry fewer of them.

To check whether this holds in our network, we compare the distribution of edge
betweenness for the two types of link with side-by-side boxplots. If the idea is
correct, the *inter* box should sit higher than the *intra* box.

```{r edge-betweenness-boxplot}
tibble(
  betweenness    = E(g)$betweenness,
  cross_community = E(g)$cross_community
) |>
  ggplot(aes(x = cross_community, y = betweenness)) +
  geom_boxplot(fill = "steelblue", alpha = 0.6) +
  labs(title = "Edge betweenness: within vs between communities",
       x = NULL, y = "Edge betweenness") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The two distributions turn out to be very similar. The *inter* box is
only slightly higher than the *intra* box, and high-betweenness links, the ones
carrying many shortest paths, appear in both groups, with the largest values
(around 28,000–29,000) found on both sides. So in our network, between-community
links are not clearly the main bridges: within-community links can carry just as
many shortest paths.

A likely reason is that our network is a pruned backbone, built from a maximum
spanning tree plus the strongest links. The spanning tree keeps the strongest
connections and turns many of them into obligatory passages of the skeleton, even
when they lie inside a community. As a result, several within-community links also
work as bridges, which blurs the difference between the two types. This is a
reminder that, as noted before, all these measures describe the backbone and not
the full proximity network.