---
title: "The Product Space (HS): Descriptive Analysis"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini, Beatrice Liuni, Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(igraph)
library(here)
library(tidyverse)
```

# **Data Loading**

```{r load-graph}
g <- readRDS(here("data", "processed", "product_space.rds"))
layout_xy <- readRDS(here("data", "processed", "layout_xy.rds"))
summary(g)
```

# **Exploratory Data Analysis**

```{r view-tables}
# A graph stores two tables underneath: one for nodes, one for edges.
# We extract them and convert to tibbles for clean printing and later use.
# Note: igraph:: prefix is needed because dplyr masks as_data_frame().

nodes_tbl <- igraph::as_data_frame(g, what = "vertices") |> as_tibble()
edges_tbl <- igraph::as_data_frame(g, what = "edges")    |> as_tibble()

# A first look at the node table (one row per product)
nodes_tbl |> select(name, pid, product, community_ref, size_layout) |> head(10)

# A first look at the edge table (one row per proximity link)
edges_tbl |> head(10)
```

## **Data Quality Check**

```{r}
# Missing values in the node and edge tables
sum(is.na(nodes_tbl))
sum(is.na(edges_tbl))

# Duplicated edges (same product pair appearing more than once)
any(which_multiple(g))

# Self-loops
any(which_loop(g))

# 4. Isolated nodes
sum(degree(g) == 0)
```

The dataset is clean at the source: there are no missing values
in either the node or edge tables, no duplicated edges, no self-loops, and no
isolated products. No additional cleaning is therefore required before analysis.
This also answers the *missingness* question from Stage 1: there is no accidental
missing data, the only data reduction is the deliberate backbone thinning
(maximum spanning tree + $\varphi > 0.55$), which removes weak proximity
links by design rather than by error.

## **Network Size and Connectivity**


```{r}
# Number of nodes (products) and edges (proximity links)
vcount(g)
ecount(g)

# Density: fraction of possible edges that actually exist.
edge_density(g)

# Number of connected components
components(g)$no

# Diameter: the longest shortest path in the network.
# weights = NA forces the UNWEIGHTED version (count steps, not edge widths)
# our primary analysis, given the layout-only weight.
diameter(g, weights = NA)

# Average shortest-path distance between any two products
mean_distance(g, weights = NA)
```

**Network size and connectivity.** The network has 866 products and 2,532
proximity links. Its density is very low ($\approx 0.0068$): only about 0.7% of all
possible edges are present, as expected for a pruned backbone. Despite being
sparse, the network forms a **single connected component**, so every product is
reachable from every other.

The **diameter is 21** (the longest shortest path) and the **average distance is $\approx 7.1$** steps between two products. These values are fairly large for a network of this size, which suggests the products are not all tightly packed together but
spread out across the network. We will look at this structure more closely with centrality and community detection.