---
title: "The Product Space (HS): Descriptive Analysis"
subtitle: "Statistical and Computational Methods for Network Data"
author: "Matteo Gerevini, Beatrice Liuni, Lorenzo Meroni"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

> **A note on the backbone.** All the statistics in this analysis describe the
> distributed network, which is a *pruned backbone* of the full product space
> (a maximum spanning tree plus the links with $\varphi > 0.55$), not the complete
> proximity matrix. Weak proximity links were removed by construction, so measures
> such as density, degree, clustering, and distances refer to this strong-link
> backbone only. We keep this in mind when interpreting the results, and we note it
> explicitly where it matters.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(igraph)
library(here)
library(tidyverse)
```

# **1. Data Loading**

```{r load-graph}
g = readRDS(here("data", "processed", "product_space.rds"))
layout_xy <- readRDS(here("data", "processed", "layout_xy.rds"))
summary(g)
```

# **2. Exploratory Data Analysis**

This notebook performs the descriptive analysis of the Product Space (HS) network.
It loads the cleaned graph produced in the construction notebook and explores its
basic properties: data quality, size, connectivity, degree, clustering, and
centrality. We first check what the vertices and edges represent, whether we observe the full
system or only a part, and whether there is sampling or missing data.

```{r view-tables}
# A graph stores two tables: one for nodes, one for edges;
# We extract them and convert to tibbles for clean printing and later use
# NB: igraph:: prefix is needed because dplyr masks as_data_frame()

nodes_tbl <- igraph::as_data_frame(g, what = "vertices") |> as_tibble()
edges_tbl <- igraph::as_data_frame(g, what = "edges")    |> as_tibble()

# A first look at the node table (one row per product)
nodes_tbl |> select(name, pid, product, community_ref, size_layout) |> head(10)

# A first look at the edge table (one row per proximity link)
edges_tbl |> head(10)
```

 The edge table shows two numeric columns,
`width_layout` and `weight`, with the same values. This is on purpose. The
original value is the line *thickness* used to draw the network in the Atlas
visualization. According to the dataset documentation this thickness is
non-linear and meant for layout only, so it is **not** the true proximity score
$\varphi$ and should not be read as a real edge weight. We keep it under two names
for two reasons: `width_layout` records what the value really is (a drawing
thickness), while `weight` is the name that igraph looks for when running weighted
computations. Our main analyses treat the network as **unweighted**; we use the
weight only later, as a robustness check, to confirm that results do not depend on
this layout-only value.

## **2.1 Data Quality Check**

```{r}
# Missing values in the node and edge tables
sum(is.na(nodes_tbl))
sum(is.na(edges_tbl))

# Duplicated edges (same product pair appearing more than once)
any(which_multiple(g))

# Self-loops
any(which_loop(g))

# Isolated nodes
sum(degree(g) == 0)
```

The dataset is clean at the source: there are no missing values in either the node
or edge tables, no duplicated edges, no self-loops, and no isolated products. No
additional cleaning is needed before analysis. This also answers the *missingness*
question from Stage 1: there is no accidental missing data. The only reduction is
the backbone construction (maximum spanning tree + $\varphi > 0.55$), which removes
weak links on purpose, not by mistake.

## **2.2 Network Size and Connectivity**

```{r network-size}
# Nodes and edges
vcount(g)
ecount(g)

# Density: fraction of possible edges that actually exist
edge_density(g)

# Number of connected components
components(g)$no

# Diameter: the longest shortest path.
# weights = NA forces the unweighted version (count steps, not edge widths),
# our primary analysis given the layout-only weight.
diameter(g, weights = NA)

# Average shortest-path distance between two products
mean_distance(g, weights = NA)
```

The network has 866 products and 2,532
proximity links. Its density is very low ($\approx 0.0068$): only about 0.7% of
all possible edges are present, as expected for a pruned backbone. Despite being
sparse, the network forms a **single connected component**, so every product is
reachable from every other.

The **diameter is 21** (the longest shortest path) and the **average distance is $\approx 7.1$** steps between two products. 
We will interpret these values together with the degree and centrality results 
in the next sections, rather than drawing conclusions now.

## **2.3 Degree**

```{r degree}
deg = degree(g)
summary(deg) # it shows a numerical summary of the degree distribution

# Top 10 most connected products
nodes_tbl|>
  mutate(degree = deg)|>
  arrange(desc(degree))|>
  select(product, community_ref, degree)|>
  head(10)
```

The degree of a product is the number of other products it is directly
connected to. The summary shows a clear imbalance: the median degree is only 2,
so half of the products have two connections or fewer, while the mean is higher
(about 5.8) and the maximum reaches 48. When the mean is well above the median,
the distribution is skewed to the right: most products have few links, but a small
number act as highly connected hubs.

The ten most connected products are mostly basic manufacturing goods (forged steel
balls, transmission shafts, cast iron articles, bricks, wooden cases, rock wool),
plus T-shirts. We will examine the shape of this degree distribution more closely
in the next step, and return to the role of these hubs and their communities later.

### **2.3.1 Degree Distribution**

```{r degree-histogram}
deg_tbl = tibble(degree = deg)

ggplot(deg_tbl, aes(x = degree)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "Degree distribution",
       x = "Degree", y = "Number of products") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The histogram confirms the imbalance seen in the summary.
The large majority of products have a very low degree, with a tall peak at degree 1
(about 285 products) that drops off quickly. To the right there is a long, thin
tail of products with much higher degree, up to the single product with degree 48.
This long-tailed shape, with many low-degree nodes and few high-degree hubs, is
typical of many real-world networks. We look at it on a log-log scale next to see
its shape more clearly.

### **2.3.2 Degree Distribution on a log-log scale**

```{r degree-loglog}
# Frequency of each degree value
deg_dist = as.data.frame(table(deg)) |>
  rename(degree = deg, freq = Freq) |>
  mutate(degree = as.numeric(as.character(degree)))

# Log-log plot: log frequency vs log degree
ggplot(deg_dist, aes(x = log(degree), y = log(freq))) +
  geom_point(color = "darkblue") +
  labs(title = "Degree distribution (log-log scale)",
       x = "log(Degree)", y = "log(Frequency)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

The histogram shows a heavy tail, but
not its precise shape. The log-log plot helps with this. If the degree
distribution follows a power law, $P(k) \propto k^{-\alpha}$, then taking logs of
both sides gives a straight line:

$$
\log P(k) = c - \alpha \, \log k,
$$

so a power law appears as a descending straight line on a log-log scale, with
slope $-\alpha$.

In our plot the points do fall roughly along a descending line, which is
consistent with a power-law (or *scale-free*) degree distribution. The fit is
cleaner on the left (low degrees), where each degree value is shared by many
products, and noisier on the right (high degrees), where only a few products takeeach value and the frequencies are estimated less reliably. We will return to
whether a power-law mechanism such as preferential attachment can explain this
shape when we compare the network to generative models.

### **2.3.3 Fitting the Degree Distribution**

The log-log plot suggests a roughly straight, descending line. To describe this
more precisely, we fit a linear model to the log-frequency against the log-degree.
This gives a numerical estimate of the slope, which corresponds to the power-law
exponent $-\alpha$, and an $R^2$ value telling us how well a straight line
describes the points.

This is a **descriptive** fit of the observed shape, not a test of any generative
model: we are measuring the slope of the line we already see, not yet asking which
mechanism produced it. We return to generative mechanisms such as preferential
attachment in the modelling part of the project.

```{r degree-fit}
fit = lm(log(freq) ~ log(degree), data = deg_dist)
summary(fit)
```

The linear fit on the log-log scale gives a
slope of about $-1.40$ for `log(degree)`, with a very small p-value, so the
descending relationship is clear and well estimated. Following the power-law form
$\log P(k) = c - \alpha \log k$, this slope corresponds to an exponent
$\alpha \approx 1.40$.

This value is somewhat lower than the range often reported for scale-free networks
(usually between 2 and 3). We note this only as a description of the observed
shape, as mentioned above, our network is a pruned backbone, which affects the
tail, and we leave conclusions about the underlying mechanism to the modelling
part, where we compare the network to generative models.

## **2.4 Clustering and Transitivity**

After looking at the degree of each product, we ask whether
the neighbours of a product are also connected to each other, i.e. whether the
network forms *triangles*.

For a vertex $v$, let $\tau_\triangle(v)$ be the number of triangles involving $v$,
and $\tau_3(v) = \binom{d_v}{2}$ the number of connected triples (wedges) centred
on $v$. The **local clustering coefficient** is

$$
cl(v) = \frac{\tau_\triangle(v)}{\tau_3(v)},
$$

defined for vertices with degree at least 2. It measures, among the pairs of
neighbours of $v$, how many are actually connected.

These local values can be aggregated to the whole network in two ways:

$$
cl(G) = \frac{1}{|V'|}\sum_{v \in V'} cl(v); 
\qquad cl_T(G) = \frac{\sum_{v \in V'}\tau_\triangle(v)}{\sum_{v \in V'}\tau_3(v)}.
$$

The first is a simple average of the local coefficients; the second, the
**transitivity**, is a weighted average. Following Bollobás and Riordan, we use the
transitivity $cl_T(G)$ as our main measure, since the weighting makes it less
sensitive to low-degree vertices. Values close to 1 indicate a highly clustered
network; values close to 0 indicate few triangles.

```{r clustering}
# Transitivity: cl_T(G)
transitivity(g, type = "global")

# Clustering Coefficient: cl(G)
transitivity(g, type = "average")
```

The transitivity is $cl_T(G) \approx 0.43$ and the
simple clustering coefficient is $cl(G) \approx 0.28$. Both are moderate-to-high
values, showing that the network forms a fair number of triangles: the neighbours
of a product are often connected to each other. This points to the presence of
dense groups of products, an early sign of community structure that we examine
later. The value is notable given that we work with a pruned backbone, where
removing weak links also removes potential triangles.

The two measures also differ ($cl_T(G) > cl(G)$). In the transitivity, each vertex
$v$ is weighted by $\tau_3(v) = \binom{d_v}{2}$, the number of connected triples
centred on it. Since this grows quickly with degree, hubs receive much larger
weights than peripheral nodes. The fact that the weighted measure is higher than
the simple average therefore suggests that the more connected products (the hubs)
tend to be more clustered than the peripheral ones: their neighbours are also
often linked to one another.

Overall, the network shows a clear tendency to form triangles, and this clustering
is concentrated around the hubs. Together with the heavy-tailed degree
distribution seen earlier, this gives a first, consistent picture: a few highly
connected, densely interconnected products alongside many peripheral ones. We test
whether this translates into well-defined communities in the modelling part.

## **2.5 Centrality**

The degree section asked *how many* links a product has. We now ask a broader
question: **how important is a product in the network?** This is the idea of
*centrality*. The difficulty is that there is no single definition of importance.
Rather than pick one measure and call it "the" importance, we compute several
measures, each built on a different idea of what makes a vertex important, and
compare them. Where they agree, the picture is robust; where they disagree, the
difference itself is informative.

We use four measures:

- **Degree centrality** counts direct neighbours. A product is important if it is
  connected to many others. We already computed this in Section 2.3, so here we use
  it only as a baseline for comparison.
- **Closeness centrality** is based on *distance*. A product is important if it is
  close, on average, to all other products through short paths.
- **Betweenness centrality** is based on *flow*. A product is important if it lies
  on many shortest paths between other products, acting as a bridge.
- **Eigenvector centrality** is based on *feedback*. A product is important if it is
  connected to other important products.

Each captures a different role: many connections (degree), short reach (closeness),
bridging position (betweenness), and connection to influential neighbours
(eigenvector).

### Mathematical formulation

For a vertex $v \in V$:

**Closeness centrality** is the inverse of the total distance from $v$ to all
other vertices,

$$
c_{Cl}(v) = \frac{1}{\sum_{u \in V} \mathrm{dist}(v, u)},
$$

where $\mathrm{dist}(v, u)$ is the shortest-path distance between $u$ and $v$. A
small total distance gives a large closeness. Note that $c_{Cl}(v)$ is only
well-defined when the graph is connected; otherwise distances to unreachable
vertices are infinite and the measure collapses to zero. Our network is a single
connected component, so this is not a problem here.

**Betweenness centrality** counts how often $v$ sits on shortest paths between
other vertices,

$$
c_B(v) = \sum_{s \neq t \neq v \in V} \frac{\sigma(s, t \mid v)}{\sigma(s, t)},
$$

where $\sigma(s, t)$ is the number of shortest paths between $s$ and $t$, and
$\sigma(s, t \mid v)$ is the number of those paths that pass through $v$. A product
with high betweenness connects parts of the network that would otherwise be far
apart.

**Eigenvector centrality** (Bonacich) defines importance recursively: the
importance of $v$ is proportional to the sum of the importances of its neighbours,

$$
c_{Ei}(u) = \alpha \sum_{\{u, v\}} c_{Ei}(v),
$$

which in matrix form is the eigenvector problem for the adjacency matrix $A$,

$$
A\,\vec{c}_{Ei} = \alpha^{-1}\,\vec{c}_{Ei}.
$$

The centrality vector is the leading eigenvector of $A$. For an undirected,
connected graph this eigenvector has all entries of the same sign; we report them
in absolute value, scaled to $(0, 1]$.

All these measures are computed on the pruned backbone, since they depend on the
shortest paths within it. Given what we already know about the network, we expect
the high-degree hubs from Section 2.3 to also score high on the other measures,
since a product with many links tends to be both close to others and connected to
important neighbours. The interesting cases will be products that rank high on one
measure but not on others, especially betweenness: a product can sit on many
shortest paths without having a high degree, if it bridges two otherwise separated
groups. We do not assume this happens; we check it.

